---
title: "Convergence"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)

result_summary <- function(res) {
  objective <- if (is.null(res$f)) NA_real_ else res$f
  data.frame(
    status = res$status,
    terminate = res$terminate$what,
    iter = res$iter,
    nf = res$nf,
    ng = res$ng,
    objective = signif(objective, 6),
    row.names = NULL
  )
}
```

An optimization can stop for three importantly different reasons:

| Outcome | `status` | `converged` | Typical termination reasons |
|:--|:--|:--|:--|
| A tolerance was satisfied | `"converged"` | `TRUE` | `abs_tol`, `rel_tol`, `grad_tol`, `ginf_tol`, `step_tol` |
| A hard limit was exhausted | `"budget_exhausted"` | `FALSE` | `max_iter`, `max_fn`, `max_gr`, `max_fg` |
| The optimization failed | `"failed"` | `FALSE` | a non-finite callback or a failed line search |

The result contract also reserves `status = "terminated"` for other terminal
reasons.

A limit often marks the intended end of a bounded run. It leaves convergence
unestablished. Inspect `status` and `converged` for the broad outcome and
`terminate$what` and `terminate$val` for the detailed reason. This article
shows how to choose among the controls and interpret tolerance, budget, and
failure outcomes.

## Choose stopping controls

Each control answers a different question:

| Control | What it detects | Main limitation | Additional observation |
|:--|:--|:--|:--|
| `abs_tol` / `rel_tol` | Little change in the objective | Stagnation need not mean stationarity | Objective |
| `grad_tol` / `ginf_tol` | A small gradient | Stationary points include saddles and maxima; absolute thresholds depend on scaling | Gradient |
| `step_tol` | Little movement in the parameters | May reflect poor scaling or a rejected step | Parameter update |
| `max_iter` | An outer iteration limit | Iterations can contain different amounts of work | None |
| `max_fn` / `max_gr` / `max_fg` | A callback cost limit | Says nothing about convergence | None |

For a general smooth problem, use a gradient criterion when its scale is
meaningful, retain callback or iteration budgets as safety limits, and treat
function-change or step tolerances as evidence of stagnation without assuming
that they establish stationarity.

We'll use the two-dimensional Rosenbrock function, which has a minimum at
`c(1, 1)` where the function equals zero.

```{r example-problem}
rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  }
)
rb0 <- c(-1.2, 1)
```

## Iteration limit

`max_iter` bounds the number of optimizer iterations. Disable the tolerances
when the purpose of an example is to demonstrate that limit:

```{r termination-by-max-iter}
res <- mize(
  rb0,
  rb_fg,
  max_iter = 10,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-max-iter, include = FALSE}
stopifnot(
  identical(res$terminate$what, "max_iter"),
  identical(res$status, "budget_exhausted"),
  identical(res$converged, FALSE)
)
```

Iteration counts are method-specific: one method may perform much more work
inside an iteration than another. For comparisons, report callback counts and
elapsed time as appropriate. Equal iteration limits do not imply equal
computational effort.

## Function-change tolerances

`abs_tol` compares consecutive function values and is reached when
`abs(fn_old - fn_new) < abs_tol`.

```{r absolute-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = 1e-8,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-absolute-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "abs_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

At a `mize()` call, `rel_tol` defaults to the value supplied for `abs_tol`.
Set `rel_tol = NULL` explicitly when you want the absolute criterion alone;
similarly, set `abs_tol = NULL` when isolating relative change.

Relative change is calculated as
`abs(fn_old - fn_new) / min(abs(fn_old), abs(fn_new))`. If the denominator is
zero, the relative change is zero when the values are equal and `Inf`
otherwise.

Consecutive objective values can become similar because a run has approached a
solution or because its steps have become too small. The following
one-dimensional quadratic isolates that distinction. Its minimizer is 10, and
a deliberately small constant step produces little relative change while the
parameter remains far away:

```{r relative-tolerance}
slow_minimum <- 10
slow_fg <- list(
  fn = function(x) (x - slow_minimum)^2,
  gr = function(x) 2 * (x - slow_minimum)
)

relative_result <- mize(
  0,
  slow_fg,
  method = "SD",
  line_search = "constant",
  step0 = 1e-4,
  max_iter = 5,
  abs_tol = NULL,
  rel_tol = 1e-3,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
```

```{r relative-tolerance-table, echo=FALSE}
relative_summary <- data.frame(
  Reason = relative_result$terminate$what,
  `Relative change` = relative_result$terminate$val,
  Parameter = relative_result$par,
  `Distance from 10` = abs(relative_result$par - slow_minimum),
  check.names = FALSE
)
knitr::kable(relative_summary, digits = 4, align = c("l", "r", "r", "r"))
```

```{r assert-relative-tolerance, include = FALSE}
stopifnot(
  identical(relative_result$terminate$what, "rel_tol"),
  identical(relative_result$status, "converged"),
  identical(relative_result$converged, TRUE),
  relative_result$terminate$val < 1e-3,
  abs(relative_result$par - slow_minimum) > 9
)
```

The relative criterion has answered its specific question: recent objective
values changed little. The distance column shows why that observation alone
does not establish that the parameter is near the minimizer. Pair it with a
stationarity criterion when that is the conclusion the run must support.

## Gradient tolerances

`grad_tol` tests the Euclidean (2-)norm of the current gradient.

```{r gradient-2-norm-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = 1e-3,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-gradient-2-norm-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "grad_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

A small gradient indicates approximate stationarity. It does not distinguish a
minimum from a saddle point or maximum, and an absolute threshold depends on
the scaling of both the parameters and the objective. Gradient norms are still
commonly used when comparing optimizers; see, for example, [Nocedal and
co-workers](https://doi.org/10.1023/A:1014897230089).

`ginf_tol` instead tests the infinity norm: the largest absolute component of
the gradient. This is another common choice, particularly for larger problems;
see the [conjugate-gradient paper by Hager and Zhang](https://doi.org/10.1137/030601880).

```{r gradient-infinity-norm-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = 1e-3,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-gradient-infinity-norm-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "ginf_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

Checking either gradient tolerance requires a gradient at the current
position. A cached current gradient avoids an additional callback; otherwise
the check requests one. The method and line search determine whether the value
is already available.

## Step tolerance

`step_tol` detects when the Euclidean distance between consecutive parameter
vectors is small.

```{r step-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = 1e-4
)
result_summary(res)
```

```{r assert-step-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "step_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

Some methods can reject an iteration and restart from the same parameters with
different optimizer state. `step_tol` ignores that restart. Because the
criterion uses an absolute Euclidean distance, rescaling one parameter can also
change when it triggers even when the underlying path describes the same
problem.

## Evaluation budgets

When function or gradient callbacks dominate runtime, their counts can be a
useful proxy for work. The relative costs still depend on the problem, and
methods and line searches differ in which callbacks they need and can reuse.

`max_fn` caps function evaluations, `max_gr` caps gradient evaluations, and
`max_fg` caps their combined total. The following runs use the same problem and
disable every ordinary tolerance so that each budget supplies the terminal
condition:

```{r evaluation-budget-comparison}
budget_names <- c("max_fn", "max_gr", "max_fg")

run_with_budget <- function(budget) {
  arguments <- list(
    par = rb0,
    fg = rb_fg,
    abs_tol = NULL,
    rel_tol = NULL,
    grad_tol = NULL,
    ginf_tol = NULL,
    step_tol = NULL
  )
  arguments[[budget]] <- 10
  do.call(mize, arguments)
}

budget_runs <- setNames(
  lapply(budget_names, run_with_budget),
  budget_names
)
```

```{r evaluation-budget-table, echo=FALSE}
budget_summary <- do.call(
  rbind,
  lapply(names(budget_runs), function(budget) {
    run <- budget_runs[[budget]]
    data.frame(
      Control = budget,
      Reason = run$terminate$what,
      `Function calls` = run$nf,
      `Gradient calls` = run$ng,
      `Total calls` = run$nf + run$ng,
      check.names = FALSE
    )
  })
)
rownames(budget_summary) <- NULL
knitr::kable(
  budget_summary,
  align = c("l", "l", "r", "r", "r")
)
```

```{r assert-evaluation-budgets, include=FALSE}
budget_reasons <- vapply(
  budget_runs,
  function(run) run$terminate$what,
  character(1)
)
budget_statuses <- vapply(budget_runs, function(run) run$status, character(1))
budget_convergence <- vapply(
  budget_runs,
  function(run) run$converged,
  logical(1)
)

stopifnot(
  identical(unname(budget_reasons), budget_names),
  all(budget_statuses == "budget_exhausted"),
  !any(budget_convergence),
  budget_runs$max_fn$nf == 10,
  budget_runs$max_gr$ng == 10,
  budget_runs$max_fg$nf + budget_runs$max_fg$ng == 10
)
```

The separate limits allow ten evaluations of each callback in this example;
the combined limit stops after ten callbacks in total. The table reports the
actual work because an iteration can consume different numbers of callbacks.

Each actual callback consumes the applicable budget; reuse of a cached value
does not. Once a callback budget is exhausted, `mize` makes no later callback
covered by that cap. This is why a final objective or requested gradient
diagnostic can be absent when its value was not already available.

## When several criteria are active

All non-`NULL` criteria are active. `terminate$what` identifies the terminal
condition reported by the optimizer; it does not imply that every other measure
was far from its threshold. Hard callback limits are checked before a covered
callback is made, so the configured cap is respected. When the final permitted
callback itself establishes convergence or produces a non-finite result, that
observed tolerance or failure is reported. The [Convergence section of the
`mize()` reference](../reference/mize.html#convergence) gives the complete
contract.

## Failure result

A non-finite callback produces a failure outcome. Here the objective is finite,
but the gradient callback returns `NaN`:

```{r non-finite-gradient}
bad_fg <- list(
  fn = function(x) sum(x^2),
  gr = function(x) rep(NaN, length(x))
)

failure_result <- mize(c(1, 1), bad_fg)
failure_result[c("status", "message", "terminate")]
```

```{r assert-non-finite-gradient, include=FALSE}
stopifnot(
  identical(failure_result$status, "failed"),
  identical(failure_result$converged, FALSE),
  identical(failure_result$terminate$what, "gr_inf")
)
```

The `gr_inf` reason identifies the non-finite gradient as the immediate
problem. Begin failure triage with `message` and `terminate`, then inspect
available progress or line-search diagnostics for the callback and iteration
that produced it.

## Checking tolerances less often

By default, `mize` checks tolerances every iteration. Function-change
tolerances need a function value at the current parameters, and gradient
tolerances need a current gradient. When the method or line search has not
already produced the needed value, checking can add a callback; when it has,
the cached value is reused.

`check_conv_every` changes the cadence of tolerance detection. For example,
this run tests the gradient criterion only every five iterations:

```{r checking-convergence-less-often}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = 1e-3,
  ginf_tol = NULL,
  step_tol = NULL,
  check_conv_every = 5
)
result_summary(res)
```

```{r assert-checking-convergence-less-often, include = FALSE}
stopifnot(
  identical(res$terminate$what, "grad_tol"),
  identical(res$status, "converged"),
  res$iter %% 5 == 0
)
```

A larger interval can delay recognition of a satisfied tolerance and therefore
overshoot it. Callback budgets remain hard: `max_fn`, `max_gr`, and `max_fg`
are enforced when a covered callback would be made. Non-finite
callback results and line-search no-step failures are also handled independently
of the ordinary tolerance cadence.

If `verbose = TRUE` or `store_progress = TRUE`, `log_every` controls how often
the corresponding summary is emitted or stored. It defaults to
`check_conv_every` and must be an integer multiple of it; otherwise `mize`
uses `check_conv_every` for both.

## Best and last results

`mize` returns explicit best-versus-last fields: `best_par` and `best_f` name
the best observed result, while `last_par` and `last_f` describe the final
optimizer state. The traditional `par` and `f` fields currently match the best
result.

Cached function or gradient information can make the best and last results
differ even with every tolerance disabled. A hard callback budget can also
prevent a final observation, leaving an objective field absent or `last_f` as
`NA`. The named fields identify the returned points; stopping settings alone
leave them ambiguous.

## See also

- [Getting started](mize.html) introduces status and result fields at the first
  optimization call.
- [Choosing methods and tuning](methods.html) explains method-specific work and
  line-search controls.
- [Stateful optimization](stateful.html) applies these termination controls in
  a caller-owned loop.
