---
title: "Convergence"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

There are a variety of ways that the optimization can terminate, running the 
gamut from good (you have reached the minimum and further work is pointless) to
bad (the solution diverged and an infinity or NaN turned up in a calculation).

We'll use the 2D Rosenbrock function for the examples, which has a minimum at
`c(1, 1)`, where the function equals `0`.

```{r Test data}
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
rb0 <- c(-1.2, 1)
```

## Iteration tolerance

An obvious way for the optimization to terminate is if you run out of 
iterations:

```{r termination by max_iter}
res <- mize(rb0, rb_fg, max_iter = 10)
res$terminate
res$f
res$par
```

When comparing different methods, the number of iterations is obviously less
important than the amount of actual CPU time you spent. Comparing results
with a fixed number of iterations is not a very good idea, because different
methods may do a lot more work within an iteration than others. See the 
section on function and gradient tolerance below.

## Function tolerance

There are two ways to specify a function tolerance, based on comparing the
difference between consecutive function values. `abs_tol` measures absolute
tolerance.

```{r Absolute tolerance}
res <- mize(rb0, rb_fg, max_iter = 100, abs_tol = 1e-8)
res$terminate
res$f
res$par
```

However, relative tolerance is often preferred, because it measures the change 
in value relative to the size of the values themselves. 

```{r Relative tolerance}
res <- mize(rb0, rb_fg, max_iter = 100, rel_tol = 1e-3)
# hit relative tolerance
res$terminate

# but stopped too early!
res$iter
res$f
res$par
```
In this example we stopped way too early. Even efficient methods like L-BFGS 
may make little progress on some iterations, so don't be too aggressive with 
relative tolerance.

## Gradient tolerance

Gradient tolerances measure the difference between the size of the gradient on
consecutive step. `grad_tol` uses the 2-norm (sometimes referred to as the
Euclidean norm) of the gradient to measure convergence.

```{r Gradient 2-norm tolerance}
res <- mize(rb0, rb_fg, abs_tol = 0, grad_tol = 1e-3)

res$terminate
res$f
res$par
```

This seems like a good stopping criterion because it is always zero at a
minimum, even if the function isn't. It is also used to compare different
methods in Nocedal and Wright's book. However, it has also been recognized that
it is not always reliable, see for instance this 
[paper by Nocedal and co-workers](http://dx.doi.org/10.1023/A:1014897230089).

Other workers suggest using the infinity norm (the maximum absolute component)
of the gradient vector, particularly for larger problems. For example, see this
[conjugate gradient paper by Hager and Zhang](http://dx.doi.org/10.1137/030601880).
To use the infinity norm, set the `ginf_norm` parameter.

```{r Gradient infinity-norm tolerance}
res <- mize(rb0, rb_fg, rel_tol = NULL, abs_tol = NULL, ginf_tol = 1e-3)

res$terminate
res$f
res$par
```

While the gradient norms aren't as reliable for checking convergence, they 
almost never incur any overhead for checking, because the gradient that's 
calculated at the end of the iteration for this purpose can nearly always be
re-used for the gradient descent calculation at the beginning of the next
iteration, whereas the function-based convergence requires the function to be
calculated at the end of the iteration and this is not always reused, although
for many line search methods it is.

## Step tolerance

You can also look out for the change in `par` itself getting too small:

```{r Step tolerance}
# set abs_tol to zero to stop it from triggering instead of step_tol
res <- mize(rb0, rb_fg, abs_tol = 0, step_tol = .Machine$double.eps)
res$terminate
res$iter
res$f
res$par
```

In most cases, the step tolerance should be a reasonable way to spot 
convergence. Some optimization methods may allow for a step size of zero for 
some iterations, preferring to commence the next iteration using the same 
initial value of `par`, but with different optimization settings. The step 
tolerance criterion knows when this sort of "restart" is being attempted, and 
does not triggered under these conditions.

## Function and gradient count tolerance

For most problems, the time spent calculating the function and gradient values 
will drown out any of the house-keeping that individual methods do, so the 
number of function and gradient evaluations is the usual determinant of how long
an optimization takes. You can therefore decide to terminate based on the number
of function evaluations:

```{r Maximum number of function evaluations}
res <- mize(rb0, rb_fg, max_fn = 10)

res$terminate
res$nf
res$f
res$par
```

Number of gradient evaluations:

```{r Maximum number of gradient evaluations}
res <- mize(rb0, rb_fg, max_gr = 10)

res$terminate
res$ng
res$f
res$par
```

or both:

```{r Maximum number of function and gradient evaluations}
res <- mize(rb0, rb_fg, max_fg = 10)

res$terminate
res$nf
res$ng
res$f
res$par
```

The function and gradient termination criteria are checked both between 
iterations and during line search. On the assumption that if you specify 
a maximum number of evaluations, that means these calculations are expensive,
`mize` errs on the side of caution and will sometimes calculate fewer 
evaluations than you ask for, because it thinks that attempting another 
iteration will exceed the limit.

## A minor complication with convergence checking

By default, convergence is checked at every iteration. For `abs_tol` and 
`rel_tol`, this means that the function needs to have been evaluated at the
current value of `par`. A lot of optimization methods do this as part of their
normal working, so it doesn't cost very much to do the convergence check.
However, not all optimization methods do. If you specify a non-`NULL` value for
`rel_tol` and `abs_tol` and the function value isn't available, it will be
calculated. This could, for some methods, add a lot of overhead.

If this is important, then using a gradient-based tolerance will be a better 
choice.

`mize` internally uses the function value as a way to keep track of the best 
`par` found during optimization. If this isn't available, it will use a gradient
norm if that is being calculated. This is less reliable than using function
values, but better than nothing. If you turn off all function and gradient
tolerances then `mize` will be unable to return the best set of parameters found
over the course of the optimization. Instead, you'll get the last set of
parameters it used.

If convergence checking at every iteration is too much of a burden, you can 
reduce the frequency with which it is carried out with the `check_conv_every` 
parameter:

```{r Checking convergence less often}
res <- mize(rb0, rb_fg, grad_tol = 1e-3, check_conv_every = 5, verbose = TRUE)
```

This also has the side effect of producing less output to the console when 
`verbose = TRUE`, because `log_every` is set to the same value of 
`check_conv_every` by default. If you set them to different values, `log_every` 
must be an integer multiple of `check_conv_every`. If it's not, it will be 
silently set to be equal to `check_conv_every`.

In many cases, however, convergence checking every iteration imposes no 
overhead, so this is a non-issue. The vignette that runs through the methods
available in `mize` mentions where it might be an issue.
