---
title: "Convergence"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)

result_summary <- function(res) {
  objective <- if (is.null(res$f)) NA_real_ else res$f
  data.frame(
    status = res$status,
    terminate = res$terminate$what,
    iter = res$iter,
    nf = res$nf,
    ng = res$ng,
    objective = signif(objective, 6),
    row.names = NULL
  )
}
```

An optimization can stop for three importantly different reasons:

| Outcome | `status` | `converged` | Typical termination reasons |
|:--|:--|:--|:--|
| A tolerance was satisfied | `"converged"` | `TRUE` | `abs_tol`, `rel_tol`, `grad_tol`, `ginf_tol`, `step_tol` |
| A hard limit was exhausted | `"budget_exhausted"` | `FALSE` | `max_iter`, `max_fn`, `max_gr`, `max_fg` |
| The optimization failed | `"failed"` | `FALSE` | a non-finite callback or a failed line search |

The result contract also reserves `status = "terminated"` for other terminal
reasons.

Reaching a limit is often an intentional end to a bounded run, but it is not
evidence of convergence. Always inspect `status` and `converged` for the broad
outcome and `terminate$what` and `terminate$val` for the detailed reason. This
article concentrates on tolerances and limits; it does not manufacture a
failure merely to complete the table.

We'll use the two-dimensional Rosenbrock function, which has a minimum at
`c(1, 1)` where the function equals zero.

```{r example-problem}
rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  }
)
rb0 <- c(-1.2, 1)
```

## Iteration limit

`max_iter` bounds the number of optimizer iterations. Disable the tolerances
when the purpose of an example is to demonstrate that limit:

```{r termination-by-max-iter}
res <- mize(
  rb0,
  rb_fg,
  max_iter = 10,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-max-iter, include = FALSE}
stopifnot(
  identical(res$terminate$what, "max_iter"),
  identical(res$status, "budget_exhausted"),
  identical(res$converged, FALSE)
)
```

Iteration counts are method-specific: one method may perform much more work
inside an iteration than another. For comparisons, report callback counts and
elapsed time as appropriate rather than treating equal iteration limits as
equal computational effort.

## Function-change tolerances

`abs_tol` compares consecutive function values and is reached when
`abs(fn_old - fn_new) < abs_tol`.

```{r absolute-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = 1e-8,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-absolute-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "abs_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

At a `mize()` call, `rel_tol` defaults to the value supplied for `abs_tol`.
Set `rel_tol = NULL` explicitly when you want the absolute criterion alone;
similarly, set `abs_tol = NULL` when isolating relative change.

Relative change is calculated as
`abs(fn_old - fn_new) / min(abs(fn_old), abs(fn_new))`. If the denominator is
zero, the relative change is zero when the values are equal and `Inf`
otherwise. A large relative tolerance can declare convergence after an
iteration that made little progress:

```{r relative-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = 1e-2,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-relative-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "rel_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE),
  res$f > 1
)
```

Here the result is still far from the known minimum even though the relative
criterion was satisfied. Efficient methods such as L-BFGS can make little
progress on an individual iteration, so choose `rel_tol` with care.

## Gradient tolerances

`grad_tol` tests the Euclidean (2-)norm of the current gradient.

```{r gradient-2-norm-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = 1e-3,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-gradient-2-norm-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "grad_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

The gradient is zero at a differentiable local minimum, even when the function
value is not zero, and gradient norms are commonly used when comparing
optimizers. They are not infallible convergence tests; see, for example,
[Nocedal and co-workers](https://doi.org/10.1023/A:1014897230089).

`ginf_tol` instead tests the infinity norm: the largest absolute component of
the gradient. This is another common choice, particularly for larger problems;
see the [conjugate-gradient paper by Hager and Zhang](https://doi.org/10.1137/030601880).

```{r gradient-infinity-norm-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = 1e-3,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-gradient-infinity-norm-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "ginf_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

Checking either gradient tolerance requires a gradient at the current
position. That may be free because the method already calculated and cached
it, or it may require an additional callback. The cost depends on the method
and its line search rather than on the tolerance alone.

## Step tolerance

`step_tol` detects when the Euclidean distance between consecutive parameter
vectors is small.

```{r step-tolerance}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = 1e-4
)
result_summary(res)
```

```{r assert-step-tolerance, include = FALSE}
stopifnot(
  identical(res$terminate$what, "step_tol"),
  identical(res$status, "converged"),
  identical(res$converged, TRUE)
)
```

Some methods can reject an iteration and restart from the same parameters with
different optimizer state. `mize` does not treat that restart as satisfying
`step_tol`.

## Evaluation budgets

When function or gradient callbacks dominate runtime, their counts can be a
useful proxy for work. The relative costs still depend on the problem, and
methods and line searches differ in which callbacks they need and can reuse.

`max_fn` is a hard function-evaluation budget:

```{r maximum-function-evaluations}
res <- mize(
  rb0,
  rb_fg,
  max_fn = 10,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-maximum-function-evaluations, include = FALSE}
stopifnot(
  identical(res$terminate$what, "max_fn"),
  identical(res$status, "budget_exhausted"),
  identical(res$converged, FALSE),
  res$nf <= 10
)
```

`max_gr` applies the same rule to gradient evaluations:

```{r maximum-gradient-evaluations}
res <- mize(
  rb0,
  rb_fg,
  max_gr = 10,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-maximum-gradient-evaluations, include = FALSE}
stopifnot(
  identical(res$terminate$what, "max_gr"),
  identical(res$status, "budget_exhausted"),
  identical(res$converged, FALSE),
  res$ng <= 10
)
```

`max_fg` bounds the combined total:

```{r maximum-combined-evaluations}
res <- mize(
  rb0,
  rb_fg,
  max_fg = 10,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
result_summary(res)
```

```{r assert-maximum-combined-evaluations, include = FALSE}
stopifnot(
  identical(res$terminate$what, "max_fg"),
  identical(res$status, "budget_exhausted"),
  identical(res$converged, FALSE),
  res$nf + res$ng <= 10
)
```

Each actual callback consumes the applicable budget; reuse of a cached value
does not. Once a callback budget is exhausted, `mize` makes no later callback
covered by that cap. This is why a final objective or requested gradient
diagnostic can be absent when its value was not already available.

## Checking tolerances less often

By default, `mize` checks tolerances every iteration. Function-change
tolerances need a function value at the current parameters, and gradient
tolerances need a current gradient. When the method or line search has not
already produced the needed value, checking can add a callback; when it has,
the cached value is reused.

`check_conv_every` changes the cadence of tolerance detection. For example,
this run tests the gradient criterion only every five iterations:

```{r checking-convergence-less-often}
res <- mize(
  rb0,
  rb_fg,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = 1e-3,
  ginf_tol = NULL,
  step_tol = NULL,
  check_conv_every = 5
)
result_summary(res)
```

```{r assert-checking-convergence-less-often, include = FALSE}
stopifnot(
  identical(res$terminate$what, "grad_tol"),
  identical(res$status, "converged"),
  res$iter %% 5 == 0
)
```

A larger interval can delay recognition of a satisfied tolerance and therefore
overshoot it. It does not soften callback budgets: `max_fn`, `max_gr`, and
`max_fg` are enforced when a covered callback would be made. Non-finite
callback results and line-search no-step failures are also handled independently
of the ordinary tolerance cadence.

If `verbose = TRUE` or `store_progress = TRUE`, `log_every` controls how often
the corresponding summary is emitted or stored. It defaults to
`check_conv_every` and must be an integer multiple of it; otherwise `mize`
uses `check_conv_every` for both.

## Best and last results

`mize` returns explicit best-versus-last fields: `best_par` and `best_f` name
the best observed result, while `last_par` and `last_f` describe the final
optimizer state. The traditional `par` and `f` fields currently match the best
result.

Disabling all tolerances does not necessarily force the best and last results
to be identical. A method or line search can leave cached function or gradient
information that lets `mize` identify an earlier best point even without a
tolerance check. Conversely, a hard callback budget can prevent a final
observation, so an objective field may be absent or `last_f` may be `NA`.
Inspect the named fields instead of inferring which point was returned from the
stopping settings alone.
