---
title: "Choosing methods and tuning"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing methods and tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)

rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  },
  fg = function(x) {
    a <- x[2] - x[1] * x[1]
    b <- 1 - x[1]
    list(
      fn = 100 * a^2 + b^2,
      gr = c(-400 * x[1] * a - 2 * b, 200 * a)
    )
  }
)
rb0 <- c(-1.2, 1)
```

This guide collects the advanced method, line-search, step-size, momentum, and
restart options available in `mize`. Start with [Getting started](mize.html)
for the basic optimization workflow, then use this guide when choosing or
tuning an optimization strategy.

## Choosing a method

`mize()` defaults to `"L-BFGS"`, a good starting point for many smooth
problems when an analytic gradient is available. Change methods for a reason:
because the problem size changes the storage budget, Hessian information is
available, or the experiment is specifically about another update rule.

| Need | Suggested starting point | Main trade-off |
|---|---|---|
| General smooth problem | `"L-BFGS"` | Limited-memory curvature model; usually the first method to try |
| Small or medium smooth problem | `"BFGS"`; consider `"SR1"` for comparison | Dense inverse-Hessian state uses quadratic storage |
| Reliable Hessian or inverse Hessian available | `"NEWTON"` | Strong local model, but Hessian construction and factorization can dominate |
| Large Hessian-free problem | `"TN"` | Approximates a Newton direction with an inner solve and extra gradient work |
| Specialized nonlinear conjugate-gradient study | `"CG"` | Low storage, but update choice, scaling, and line search matter |
| Momentum experiment | `"NAG"` or `"Momentum"` | Requires a defined schedule and careful comparison of nonmonotone trajectories |
| Known or deliberately controlled learning rate | `"SD"` | Simple baseline; performance depends heavily on the step-size strategy |
| Per-coordinate adaptive learning rate | `"DBD"` | Heuristic update controls both direction scaling and step size |

For algorithmic background, see
[Nocedal and Wright](https://doi.org/10.1007/978-0-387-40065-5). The choices
above are starting points, not rankings that hold for every objective.

### Current method surface

- `"L-BFGS"` stores the most recent `memory` curvature pairs instead of a
  dense inverse-Hessian approximation. Increasing `memory` retains more
  history and costs more storage; it does not make the method equivalent to
  BFGS.
- `"BFGS"` maintains a dense inverse-Hessian approximation, so its storage is
  quadratic in the number of parameters.
- `"SR1"` is a dense symmetric-rank-one alternative. Its update is safeguarded,
  and `mize` falls back to a BFGS or steepest-descent direction when necessary
  to obtain descent.
- `"NEWTON"` uses `fg$hs` or `fg$hi`. A Hessian or inverse Hessian may be a
  full matrix or a diagonal vector; unsuccessful or non-descent directions
  fall back to steepest descent and can be inspected through
  `direction_reason`.
- `"TN"` (truncated Newton) approximately solves for a Newton direction using
  gradient differences. `tn_init`, `tn_exit`, and the optional L-BFGS
  `preconditioner` control its inner solve.
- `"CG"` is nonlinear conjugate gradient. `cg_update = "PR+"` is the default;
  `"HZ+"` is another restart-enabled choice. The reference topic lists the
  complete update vocabulary.
- `"SD"` is steepest descent. It is useful as a baseline and with a deliberately
  chosen constant or adaptive step size, but it is not guaranteed to improve
  the objective for an arbitrary step.
- `"NAG"` is Nesterov accelerated gradient. Its momentum depends on `nest_q`
  and related controls, so compare bounded runs rather than assuming a
  universal improvement over another method.
- `"Momentum"` (which may be abbreviated to `"MOM"`) combines steepest descent
  with a classical or Nesterov-style momentum stage and a chosen schedule.
- `"DBD"` is Delta-Bar-Delta. It owns a per-coordinate learning rate, so its
  step controls replace the requested `line_search`.
- `"PHESS"` reuses a partial-Hessian factorization. It is accepted by the
  factory but is experimental/internal, requires `fg$hs`, and is not a
  recommended starting method.

```{r method-surface-check, include = FALSE}
documented_methods <- c(
  "SD", "BFGS", "SR1", "L-BFGS", "CG", "TN", "NEWTON", "PHESS", "NAG",
  "Momentum", "DBD"
)
stopifnot(all(vapply(
  documented_methods,
  function(method) is.list(make_mize(method = method)),
  logical(1)
)))
```

Method-specific controls are not interchangeable. For example, `memory`
applies to L-BFGS and to the optional L-BFGS preconditioner for CG or TN;
`cg_update` applies only to CG; and `tn_init` and `tn_exit` apply only to TN.
Other methods ignore those controls. The same rule applies to step-size
controls: DBD ignores `line_search`, and Wolfe-only controls are ignored by
constant and bold-driver steps. See [`?mize`](../reference/mize.html) for the
complete accepted values and control scopes.

### Nesterov Accelerated Gradient (`"NAG"`)

The Nesterov Accelerated Gradient method consists of doing a steepest
descent step immediately followed by a step that resembles momentum (see below),
but with the previous gradient term replaced with the current gradient.

```{r NAG}
res <- mize(rb0, rb_fg, max_iter = 10, method = "NAG")
```

Be aware that the convergence of NAG is not guaranteed to be monotone, that is
you may (and probably will) see the function value increase between iterations.
Getting the optimum convergence rates guaranteed by theory requires you to be
able to estimate how strong the convexity of the function you are minimizing is.

To see this in action, let's increase the iteration count to 100 and plot the 
progress of the function:

```{r NAG with 100 steps}
res <- mize(rb0, rb_fg, max_iter = 100, method = "NAG", store_progress = TRUE)
plot(res$progress$nf, log(res$progress$f), type = "l")
res$f
```

Yikes. If you let this optimization continue, it does eventually settle down,
but it's a real test of patience. Clearly, in this case, there's probably
a better estimate of the strong convexity parameter.

You can control the strong convexity estimate with the `nest_q` parameter, which
is inversely related to the amount of momentum that is applied. It should take 
a value between `0` (you think your function is strongly convex) and `1` 
(method becomes steepest descent). The default is zero, and this is in fact
the assumed value in most applications and papers.

Let's try turning down the momentum a little bit and repeat:

```{r NAG with 100 steps and less aggressive momentum}
resq <- mize(rb0, rb_fg, max_iter = 100, method = "NAG", nest_q = 0.001, 
            store_progress = TRUE)
plot(res$progress$nf, log(res$progress$f), type = "l",
     ylim = range(log(res$progress$f), log(resq$progress$f)))
lines(resq$progress$nf, log(resq$progress$f), col = "red")
resq$f
```

The red line is the result with `nest_q = 0.001`. That's better. Unfortunately, 
I have no advice on how you should go about estimating the right value for 
`nest_q`. However, see the section on Adaptive Restart for ways to ameliorate 
the problem.

To get the theoretical guarantees about convergence, you actually need to follow
a fairly strict formula about what the step size should be (it's related to the
reciprocal of the Lipschitz constant of the function), and 
[Sutskever](https://hdl.handle.net/1807/36012) notes that for non-convex 
functions (with sparse stochastic gradient) the theoretically optimal momentum 
and step size values are too aggressive. Alas, the best advice (at least for
deep learning applications) given was "manual tuning".

There are connections between NAG and classical momentum, and there is some
evidence (at least in deep learning) that replacing classical momentum
with the NAG method, but using the same momentum schedule (see below) can
improve results. I have written more (much more) on the subject
[here](https://jlmelville.github.io/mize/articles/nesterov.html).

### Classical Momentum (`"MOM"`)

Momentum methods use steepest descent and then add a fraction of the previous
iteration's direction. Often the effectiveness of this approach is likened
to inertia and has the effect of damping the oscillation in direction that 
slows the progress of steepest descent with poorly-scaled functions.

Originally popular for training neural networks, the momentum value is often
a constant and set to a fairly high value, like `0.9`. You can set a constant
momentum by setting the `mom_schedule` parameter:

```{r Momentum}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = 0.9)
```

Momentum-based methods share the same issues as Nesterov Accelerated Gradient:
convergence need not be monotone. The plot below demonstrates:

```{r Momentum plot}
res <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9,
            store_progress = TRUE)
plot(res$progress$nf, log(res$progress$f), type = "l")
res$f
```

To be fair, there is little reason to believe that the sort of settings used
for neural network training would be effective for minimizing the Rosenbrock
function.

Non-constant momentum schedules are also available, by passing a string to 
`mom_schedule`. These include:

* A `switch` function, that steps the momentum from one value (`mom_init`), 
to another (`mom_final`) at a specified iteration (`mom_switch_iter`):
```{r momentum with a switch function}
# Switch from a momentum of 0.4 to 0.8 at iteration 5
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = "switch",
            mom_init = 0.4, mom_final = 0.8, mom_switch_iter = 5)
```
* A `ramp` function, the linearly increases the momentum from `mom_init`
to `mom_final` over `max_iter` iterations:
```{r momentum with a ramp function}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = "ramp",
            mom_init = 0.4, mom_final = 0.8)
```
* The schedule used in NAG can be used here by asking for a `"nsconvex"` 
schedule.
```{r momentum with nesterov schedule}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = "nsconvex")
```
* The `nest_q` parameter can also be used here as it can be with NAG:
```{r momentum with nesterov schedule and non-zero q}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = "nsconvex",
            nest_q = 0.001)
```

If none of this meets your needs, you can pass a function that takes the current
iteration number and the maximum number of iterations as input and returns a 
scalar, which will be used as the momentum value, although the value will be 
clamped between 0 and 1.

For example, you could create a momentum schedule that randomly picks a value
between 0 and 1:
```{r momentum with random momentum}
mom_fn <- function(iter, max_iter) {
  runif(n = 1, min = 0, max = 1)
}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = mom_fn)
```
I'm not saying it's a *good* idea.

### Simplified Nesterov Momentum

Interest in NAG in the neural network community (specifically deep learning),
was sparked by papers from 
[Sutskever and co-workers](https://proceedings.mlr.press/v28/sutskever13.html)
and [Bengio and co-workers](https://arxiv.org/abs/1212.0901) who showed that 
it could be related to classical momentum in a way that made it easy to 
implement if you already had a working classical momentum implementation, and
more importantly, that it seemed to give improvements over classical momentum.

As `mize` already offers NAG by setting `method = "NAG"`, being able to use it
in a different form related to classical momentum may seem superfluous. But in 
fact, the use of what's often called "NAG" in deep learning is slightly more
general (and as far as I'm aware, has no convergence theory to go with it, so
improvements are not guaranteed). It's probably better to refer to it as 
(simplified) Nesterov momentum.

At any rate, the key observation of Sutskever was that NAG could be considered
a momentum method where the momentum stage occurs before the gradient descent
stage. In this formulation, you can have whatever momentum schedule you like,
and the hope is that the gradient descent stage can more easily correct if the 
momentum schedule is too aggressive because it gets to "see" the result of the 
momentum update and hence the descent occurs from a different location.

To get Nesterov momentum, you need only add a `mom_type = "nesterov"` parameter
to a momentum optimizer:

```{r Simplified Nesterov momentum}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", mom_schedule = 0.9, 
            mom_type = "nesterov")
```

I know you're itching to see whether Nesterov momentum can calm down the bad 
case of non-monotone convergence we got with a large classical momentum value
on the Rosenbrock function:

```{r Nesterov versus classical momentum}
resc <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             store_progress = TRUE)
resn <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             mom_type = "nesterov", 
             store_progress = TRUE)
# Best f found for Nesterov momentum
resn$f
# Best f found for classical momentum
resc$f
plot(resc$progress$nf, log(resc$progress$f), type = "l",
     ylim = range(log(resc$progress$f), log(resn$progress$f)))
lines(resn$progress$nf, log(resn$progress$f), col = "red")
```

The Nesterov momentum is in red. And yes, it's a lot smoother. On the other 
hand, you can see that the cost function increases and then levels off.
Clearly, some tweaking of the correct momentum schedule to use is needed.

Sutskever also came up with a simplified approximation to the momentum schedule
used in NAG. It doesn't have a big effect on the run time, only applies for
the default `nest_q` of 0, and at early iterations the momentum is much larger.
To use it, set the `nest_convex_approx = TRUE` when using the 
`mom_schedule = "nesterov"`:

```{r Nesterov momentum with convex approximation}
res <- mize(rb0, rb_fg, max_iter = 10, method = "MOM", 
            mom_schedule = "nsconvex", nest_convex_approx = TRUE, 
            mom_type = "nesterov")
```

## Line searches and step sizes

Most methods first choose a direction and then use `line_search` to choose how
far to move. DBD is the main exception: it chooses per-coordinate learning
rates itself and ignores `line_search`.

| Choice | What it accepts | Useful starting context |
|---|---|---|
| `"More-Thuente"` | Strong Wolfe conditions by default | Default for direction methods, including quasi-Newton and CG |
| `"Rasmussen"` or `"Schmidt"` | Strong Wolfe conditions by default | Alternative Wolfe implementations for comparison or troubleshooting |
| `"Hager-Zhang"` | Standard curvature plus approximate Armijo by default | CG-oriented Wolfe search with Hager-Zhang initializers |
| `"Backtracking"` | Armijo sufficient decrease | Simpler search when a curvature condition is unnecessary |
| `"Bold Driver"` | Any finite objective decrease | Heuristic adaptation from the preceding accepted step |
| `"Constant"` | No search; always uses numeric `step0` | Known learning rate or a controlled experiment |

These choices are not freely interchangeable with every control. `c2`,
`strong_curvature`, and `approx_armijo` apply only to Wolfe searches. `c1`
also applies to Armijo backtracking, but not to bold driver or constant steps.
`ls_max_alpha` and `ls_safe_cubic` are More-Thuente-specific. The DBD controls
are described separately below.

### Wolfe conditions and `c2`

For a direction $p$, define the line function
$\phi(\alpha) = f(x + \alpha p)$ and assume the initial slope
$\phi'(0) = g(x)^T p$ is negative. Wolfe searches require sufficient decrease
and a curvature condition. Under the strong curvature condition,

$$
|\phi'(\alpha)| \leq c_2 |\phi'(0)|.
$$

A smaller `c2` makes the right-hand side smaller, so it is **stricter**; a
larger `c2` is looser. The same direction holds for the standard curvature
condition because the initial slope is negative. `c2` must lie strictly
between `c1` and 1.

The CG default is `c2 = 0.1`. The next bounded example deliberately increases
it to `0.5`, making the curvature test looser. This may change accepted steps
and callback counts; it is a tuning example, not a claim that looser is better.
It also selects the Hager-Zhang next-step initializer so its possible probe is
covered by the displayed callback diagnostics.

```{r looser-c2-diagnostics}
looser_c2 <- mize(
  rb0,
  rb_fg,
  method = "CG",
  cg_update = "HZ+",
  line_search = "More-Thuente",
  c2 = 0.5,
  step_next_init = "hz",
  max_iter = 4,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL,
  store_progress = TRUE
)

line_search_columns <- c(
  "alpha_init", "slope_init", "ls_reason", "ls_outcome", "ls_nf", "ls_ng"
)
line_search_diagnostics <- looser_c2$progress[
  !is.na(looser_c2$progress$ls_reason),
  line_search_columns,
  drop = FALSE
]
rownames(line_search_diagnostics) <- NULL
line_search_diagnostics
```

`alpha_init` is the safeguarded starting proposal and `slope_init` is
$\phi'(0)$. `ls_reason` says why the search stopped, while `ls_outcome` says
what kind of point was ultimately selected; the two need not be identical.
`ls_nf` and `ls_ng` count callbacks owned by that outer line search, excluding
its cached starting point. These columns are method-specific and appear only
when their owning stage supplies them.

```{r looser-c2-check, include = FALSE}
stopifnot(
  identical(looser_c2$status, "budget_exhausted"),
  identical(looser_c2$terminate$what, "max_iter"),
  identical(names(line_search_diagnostics), line_search_columns),
  nrow(line_search_diagnostics) > 0,
  all(is.finite(line_search_diagnostics$alpha_init)),
  all(line_search_diagnostics$alpha_init > 0),
  all(is.finite(line_search_diagnostics$slope_init)),
  all(line_search_diagnostics$slope_init < 0),
  all(line_search_diagnostics$ls_nf >= 0),
  all(line_search_diagnostics$ls_ng >= 0)
)
```

More-Thuente, Rasmussen, and Schmidt use strong curvature with the exact Armijo
condition by default. Hager-Zhang uses the standard curvature condition with
the approximate Armijo condition by default. `strong_curvature` and
`approx_armijo` can override those choices, but doing so changes the conditions
whose theory applies; treat it as an explicit experiment.

### Initial step estimates

On the first outer iteration, the named `step0` choices use the starting
gradient $g$ and direction $p$. Writing the directional derivative as
$\phi'(0) = g^T p$, the current formulas are:

- `"rasmussen"`: $1 / (1 - \phi'(0))$. For steepest descent, this becomes
  $1 / (1 + \lVert g \rVert_2^2)$.
- `"scipy"`: $-\lVert g \rVert_2 / \phi'(0)$. For steepest descent, where
  $p = -g$, this reduces to $1 / \lVert g \rVert_2$.
- `"schmidt"`: $1 / \lVert g \rVert_1$; there is no added 1 in the
  denominator.
- `"hz"`: Hager-Zhang scaling based first on the relative infinity norms of
  the parameters and gradient, then on the objective and squared gradient norm,
  with a safeguarded fallback.

A positive finite numeric `step0` bypasses these formulas. Hager-Zhang uses its
own first-step estimator by default; the other gradient searches default to
Rasmussen.

After the first iteration, `step_next_init` can use a slope ratio, quadratic
interpolation, the Hager-Zhang `"hz"` rule, or a fixed positive number.
Hager-Zhang's next-step rule may make one objective-only probe to fit a
quadratic. The probe is attempted only when there is a previous step and both
the local line-search budgets and the optimizer's remaining hard budgets leave
enough callbacks for the probe and a subsequent search. Otherwise the rule
falls back to an enlarged, safeguarded version of the previous step. When the
probe occurs, it is included in `ls_nf`, as in the diagnostic vocabulary above;
it is not an unconditional extra evaluation.

`try_newton_step` gives the calculated proposal a small upward adjustment, up
to the natural unit step of 1. Its current default is `TRUE` for `"NEWTON"`,
`"PHESS"`, `"BFGS"`, `"SR1"`, `"L-BFGS"`, and `"TN"`, and `FALSE` otherwise.
`"PHESS"` remains experimental/internal despite sharing this default. Set
`try_newton_step` explicitly when comparing initialization policies.

### Backtracking, constant, and bold-driver steps

Armijo backtracking repeatedly reduces the candidate until it provides the
decrease controlled by `c1`. With `step_down = NULL`, the current implementation
uses cubic interpolation and evaluates function and gradient at candidate
points. A numeric `step_down` instead multiplies the candidate by that fixed
factor and uses function-only candidate evaluations.

```{r armijo-and-fixed-steps}
backtracking_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Backtracking",
  step0 = 1,
  step_down = 0.5,
  max_iter = 10
)

constant_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Constant",
  norm_direction = TRUE,
  step0 = 0.01,
  max_iter = 10
)

bold_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Bold Driver",
  step_down = 0.5,
  step_up = 1.1,
  max_iter = 10
)
```

The constant strategy requires a numeric `step0` and performs no search. With
steepest descent, `norm_direction = TRUE` makes `step0` the length of the
parameter update rather than a multiplier of the gradient norm.

Bold driver accepts the first tested point with a lower objective, reduces
failed proposals by `step_down`, and initializes the next iteration by
multiplying the last accepted step by `step_up`. In the current implementation
the first bold-driver proposal is 1; `step0`, `c1`, `c2`, and `step_up_fun` do
not configure it. `ls_max_fn` is its local callback limit.

For Wolfe and backtracking searches, `ls_max_fn`, `ls_max_gr`, and `ls_max_fg`
limit callbacks within one outer line search. They do not replace the optimizer
budgets `max_fn`, `max_gr`, and `max_fg`, which remain hard limits for the
entire run. The default local function limit is 20. See the
[line-search reference](../reference/mize.html#line-search) for the remaining
specialized safeguards.

## Delta-Bar-Delta learning rates

`method = "DBD"` changes a per-coordinate learning rate from the agreement in
sign between the current gradient and a weighted history. Because DBD chooses
both the direction scaling and the step, `line_search` is ignored.

The main controls are:

- `step0`, the initial per-coordinate learning rate (a positive number or a
  named first-step estimator);
- `step_down`, the multiplicative decrease when signs disagree;
- `step_up`, the increase when signs agree;
- `step_up_fun`, either `"*"` for a proportional increase (the default) or
  `"+"` for an additive increase; and
- `dbd_weight`, the gradient-history weight when no momentum schedule is used.

The following uses the additive update associated with the t-SNE-style
variant. It is an example of the control semantics, not a generally recommended
setting for smooth optimization.

```{r additive-dbd}
dbd_res <- mize(
  rb0,
  rb_fg,
  method = "DBD",
  step0 = "rasmussen",
  step_down = 0.8,
  step_up = 0.2,
  step_up_fun = "+",
  dbd_weight = 0.5,
  max_iter = 10
)
```

If a momentum schedule is supplied with DBD, `dbd_weight` is ignored and the
momentum update supplies the history. DBD's step calculation itself is
gradient-based, but function-based convergence criteria still request objective
evaluations, and `mize()` may need an objective observation to report a best
result. Choose convergence controls according to the observations the run can
afford rather than assuming DBD is function-free.

## Adaptive Restart

[O'Donoghue and Candes](https://arxiv.org/abs/1204.3982) described a restart
scheme for NAG as a cure for the non-monotonic convergence. They suggest using
either a function-based check (did the function value after the gradient descent
stage decrease between iterations?) or a gradient-based check (is the momentum
direction a descent direction?). If the check fails, then the momentum is 
restarted: the previous update vector is discarded and the momentum set back to
zero.

`mize` allows these checks to be used to restart any momentum scheme, not just 
NAG. The only change is that the function-based check is carried out at the
beginning and end of each iteration, rather than at the end of the gradient
descent stage. Restarting is applied using the `restart` argument, supplying
either `fn` or `gr` for function-based and gradient-based restarting,
respectively:

```{r momentum with restart}
resc <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             store_progress = TRUE)
resf <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             store_progress = TRUE, restart = "fn")
resg <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             store_progress = TRUE, restart = "gr")
plot(resc$progress$nf, log(resc$progress$f), type = "l", 
     ylim = range(log(resc$progress$f), log(resf$progress$f),
                  log(resg$progress$f)))
lines(resf$progress$nf, log(resf$progress$f), col = "red")
lines(resg$progress$nf, log(resg$progress$f), col = "blue")
```

Adaptive restart (red and blue lines) certainly seems to help here, although
with either gradient or function-based restart seeming equally effective.

The only downside to adaptive restart is that you may want to choose the type
of restart to not require more overhead of function and gradient calls than
necessary. In practice, the gradient-based restart is economical because in
most cases, the gradient calculated at the end of the iteration can be re-used
for the gradient descent at the start of the next iteration. The one exception
here is when using nesterov momentum, where the gradients aren't calculated
in the right location to be of use. In that case, you are better off using
function-based restarting, although if you aren't checking convergence
with a function calculation or using a line search method that makes use of
the function at the starting point, you will obviously incur the overhead
of the function call. This is one reason to favour using the NAG method directly
rather than emulating it with Nesterov momentum.

There is also a `restart_wait` parameter. This determines how many iterations 
to wait between attempted restarts. By default, using the value from the paper
of Su and co-workers, we wait 10 iterations before allowing another restart.
Making the wait time too short may lead to premature convergence:

```{r momentum with restart and wait time}
resfw <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             store_progress = TRUE, restart = "fn", restart_wait = 1)
resgw <- mize(rb0, rb_fg, max_iter = 100, method = "MOM", mom_schedule = 0.9, 
             store_progress = TRUE, restart = "gr", restart_wait = 1)
plot(resc$progress$nf, log(resc$progress$f), type = "l", 
     ylim = range(log(resc$progress$f), log(resf$progress$f),
                  log(resg$progress$f), log(resfw$progress$f),
                  log(resgw$progress$f)))
lines(resf$progress$nf, log(resf$progress$f), col = "red")
lines(resfw$progress$nf, log(resfw$progress$f), col = "blue")
lines(resgw$progress$nf, log(resgw$progress$f), col = "orange")
```

Here, reducing the wait time to 1 had a great effect on gradient restarting
(the orange line), it substantially outperforms the other methods.
On the other hand, function restarting more often (the blue line) led to worse 
performance.

There is also a `"speed"` restart type, related to a restart strategy introduced
by [Su and co-workers](https://papers.nips.cc/paper/5322-a-differential-equation-for-modeling-nesterovs-accelerated-gradient-method-theory-and-insights),
which restarts if the update vector doesn't increase in length (as measured by
Euclidean 2-norm) between iterations. This seems less performant in their
experiments than gradient-based restarting, but might provide more stability
under some circumstances.
