---
title: "Choosing methods and tuning"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing methods and tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)

rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  },
  fg = function(x) {
    a <- x[2] - x[1] * x[1]
    b <- 1 - x[1]
    list(
      fn = 100 * a^2 + b^2,
      gr = c(-400 * x[1] * a - 2 * b, 200 * a)
    )
  }
)
rb0 <- c(-1.2, 1)
```

This guide helps you decide whether a concrete problem or diagnostic justifies
changing the default optimizer. Start with [Getting started](mize.html) for a
first optimization, then return here to choose a direction method, investigate
a line search, or run a controlled experiment with a specialized update.

## Start here

For a smooth objective with an analytic gradient, begin with the default
`"L-BFGS"` method and line search. It builds a limited-memory curvature model,
so it is a useful baseline across a wide range of problem sizes. Record the
metric that matters for your application---such as callback count, elapsed
time, or objective value at a fixed budget---before comparing alternatives.

## Before changing the method

Check the problem setup and the evidence first:

- Validate the analytic gradient at representative points.
- Examine parameter and objective scaling.
- Distinguish an expensive line search from a failed search by inspecting its
  diagnostic fields.
- Confirm that memory or derivative availability is the actual constraint.
- Decide whether the comparison concerns iterations, callbacks, elapsed time,
  best objective, or the final iterate.

These checks often identify a local tuning problem without changing the
direction method. The [Convergence guide](convergence.html) explains how
stopping rules and budgets affect the result being compared.

## Change the direction method when

| Method | Consider it when | Main trade-off or diagnostic |
|---|---|---|
| `"L-BFGS"` | You need a general smooth-problem baseline | `memory` trades curvature history for storage |
| `"BFGS"` | Dense state is affordable | Inverse-Hessian storage is quadratic in the parameter count |
| `"SR1"` | You want a dense alternative to BFGS | Safeguards may fall back to BFGS or steepest descent |
| `"NEWTON"` | A reliable Hessian or inverse Hessian is available | Construction and factorization can dominate; inspect `direction_reason` after a fallback |
| `"TN"` | The problem is large and Hessian-free inner work is acceptable | Gradient differences drive an iterative inner solve |
| `"CG"` | Low storage matters and you can tune a nonlinear-CG configuration | Update choice, scaling, and line-search behavior all matter |
| `"SD"` | You need a baseline or know a suitable learning rate | Progress depends strongly on the step-size strategy |
| `"NAG"` or `"Momentum"` | Momentum behavior is the subject of the experiment | Define a schedule and compare nonmonotone trajectories explicitly |
| `"DBD"` | You want a per-coordinate adaptive learning-rate experiment | Its update controls both direction scaling and step size |

`"PHESS"` reuses a partial-Hessian factorization when `fg$hs` is available.
It remains experimental/internal, so it is documented for completeness rather
than recommended as a normal starting point. For algorithmic background on
the general direction methods, see
[Nocedal and Wright](https://doi.org/10.1007/978-0-387-40065-5).

When supplied curvature is expensive, compare `nh` and `nhi` alongside `nf`
and `ng`. The first pair count accepted `fg$hs` and `fg$hi` callback results;
they do not count the internal updates used to construct BFGS, SR1, or L-BFGS
approximations. This keeps callback cost distinct from algorithmic curvature
work. See the [`mize()` result reference](../reference/mize.html#value) for the
counting contract and the [Convergence guide](convergence.html#evaluation-budgets)
for its relationship to callback budgets.

```{r method-surface-check, include = FALSE}
documented_methods <- c(
  "SD", "BFGS", "SR1", "L-BFGS", "CG", "TN", "NEWTON", "PHESS", "NAG",
  "Momentum", "DBD"
)
stopifnot(all(vapply(
  documented_methods,
  function(method) is.list(make_mize(method = method)),
  logical(1)
)))
```

Each method-specific control has a fixed scope. For example, `memory` applies
to L-BFGS and to the optional L-BFGS preconditioner for CG or TN;
`cg_update` applies only to CG; and `tn_init` and `tn_exit` apply only to TN.
Other methods ignore those controls. The same rule applies to step-size
controls: DBD ignores `line_search`, and Wolfe-only controls are ignored by
constant and bold-driver steps. See [`?mize`](../reference/mize.html) for the
complete accepted values and control scopes.

## Tune the line search when

Most methods choose a direction and then use `line_search` to choose how far
to move. With `store_progress = TRUE`, inspect the diagnostics produced by the
search in `res$progress`. In a stateful loop, the corresponding diagnostics are
returned by `mize_step_summary()`. Start with the field associated with the
symptom before changing a control.

| Symptom | First evidence to inspect | Possible next action |
|---|---|---|
| Many evaluations inside each search | `ls_nf`, `ls_ng`, `alpha_init`, and `ls_reason` | Revisit scaling, the initial-step rule, or the curvature tolerance |
| Repeated searches find no usable step | Gradient accuracy, `ls_outcome`, and `direction_reason` | Correct the gradient or scaling; then reconsider method/search compatibility |
| Quasi-Newton steps are repeatedly shortened | Accepted `alpha`, local model quality, and parameter scaling | Compare `try_newton_step` policies or rescale the problem |
| A stable learning rate is already known | Whether a search supplies useful adaptation | Use a constant step in a controlled run |
| Callback cost must be capped per search | `ls_nf`, `ls_ng`, and the global callback totals | Set local search budgets while retaining global hard budgets |

Once the diagnostics indicate that the search policy itself is worth changing,
the available choices differ in the acceptance conditions and adaptation they
provide:

| Choice | What it accepts | Useful starting context |
|---|---|---|
| `"More-Thuente"` | Strong Wolfe conditions by default | Default for direction methods, including quasi-Newton and CG |
| `"Rasmussen"` or `"Schmidt"` | Strong Wolfe conditions by default | Alternative Wolfe implementations for comparison or troubleshooting |
| `"Hager-Zhang"` | Standard curvature plus approximate Armijo by default | CG-oriented Wolfe search with Hager-Zhang initializers |
| `"Backtracking"` | Armijo sufficient decrease | Simpler search when a curvature condition is unnecessary |
| `"Bold Driver"` | Any finite objective decrease | Heuristic adaptation from the preceding accepted step |
| `"Constant"` | No search; always uses numeric `step0` | Known learning rate or a controlled experiment |

`c2`, `strong_curvature`, and `approx_armijo` apply only to Wolfe searches.
`c1` also applies to Armijo backtracking. Bold driver and constant steps ignore
those controls, and `ls_max_alpha` and `ls_safe_cubic` are
More-Thuente-specific.

### Wolfe conditions and `c2`

For a direction $p$, define the line function
$\phi(\alpha) = f(x + \alpha p)$ and assume the initial slope
$\phi'(0) = g(x)^T p$ is negative. Wolfe searches require sufficient decrease
and a curvature condition. Under the strong curvature condition,

$$
|\phi'(\alpha)| \leq c_2 |\phi'(0)|.
$$

A smaller `c2` makes the right-hand side smaller and the condition stricter;
a larger `c2` makes it looser. The same relationship holds for the standard
curvature condition because the initial slope is negative. `c2` must lie
strictly between `c1` and 1.

The CG default is `c2 = 0.1`. The next comparison changes only `c2`, holding
the start, direction update, line search, initializer, and four-iteration
budget fixed. Proposed and selected step lengths show how the search responded;
local work reports function and gradient callbacks owned by that outer search.

```{r c2-diagnostics}
c2_runs <- lapply(c(`0.1` = 0.1, `0.5` = 0.5), function(c2_value) {
  mize(
    rb0,
    rb_fg,
    method = "CG",
    cg_update = "HZ+",
    line_search = "More-Thuente",
    c2 = c2_value,
    step_next_init = "hz",
    max_iter = 4,
    abs_tol = NULL,
    rel_tol = NULL,
    grad_tol = NULL,
    ginf_tol = NULL,
    step_tol = NULL,
    store_progress = TRUE
  )
})

c2_diagnostics <- do.call(rbind, lapply(seq_along(c2_runs), function(i) {
  progress <- c2_runs[[i]]$progress
  progress <- progress[!is.na(progress$ls_reason), , drop = FALSE]
  data.frame(
    `c2` = names(c2_runs)[i],
    Iteration = seq_len(nrow(progress)),
    `Proposed alpha` = progress$alpha_init,
    `Selected alpha` = progress$alpha,
    `Reason / outcome` = paste(progress$ls_reason, progress$ls_outcome, sep = " / "),
    `Local work (fn / gr)` = paste(progress$ls_nf, progress$ls_ng, sep = " / "),
    check.names = FALSE
  )
}))
rownames(c2_diagnostics) <- NULL
c2_display <- c2_diagnostics
c2_display[["Proposed alpha"]] <- formatC(
  c2_display[["Proposed alpha"]],
  format = "g",
  digits = 4
)
c2_display[["Selected alpha"]] <- formatC(
  c2_display[["Selected alpha"]],
  format = "g",
  digits = 4
)
knitr::kable(c2_display)
```

For this run, changing the curvature tolerance changes accepted step lengths
and callback work. `ls_reason` records why a search stopped, while
`ls_outcome` classifies the point it selected; the two fields answer different
diagnostic questions. The callback counts exclude the cached starting point
and belong only to the current outer line search.

```{r c2-check, include = FALSE}
stopifnot(
  all(vapply(c2_runs, function(result) {
    identical(result$status, "budget_exhausted") &&
      identical(result$terminate$what, "max_iter")
  }, logical(1))),
  all(is.finite(c2_diagnostics[["Proposed alpha"]])),
  all(c2_diagnostics[["Proposed alpha"]] > 0),
  all(is.finite(c2_diagnostics[["Selected alpha"]])),
  all(c2_diagnostics[["Selected alpha"]] > 0),
  !isTRUE(all.equal(
    c2_runs[[1]]$progress$alpha,
    c2_runs[[2]]$progress$alpha
  ))
)
```

More-Thuente, Rasmussen, and Schmidt use strong curvature with the exact Armijo
condition by default. Hager-Zhang uses the standard curvature condition with
the approximate Armijo condition by default. `strong_curvature` and
`approx_armijo` can override those choices. Such an override changes the
conditions whose theory applies, so make it an explicit part of the
experiment.

### Initial step estimates

Leave the initializer defaults in place unless diagnostics show repeated poor
proposals or you need to reproduce another implementation. Hager-Zhang uses
its own first-step estimator by default; the other gradient searches default
to Rasmussen. A positive finite numeric `step0` bypasses the named formulas.

<details>
<summary>Exact initial-step formulas</summary>

On the first outer iteration, the named `step0` choices use the starting
gradient $g$ and direction $p$. Write the directional derivative as
$\phi'(0) = g^T p$.

- `"rasmussen"`: $1 / (1 - \phi'(0))$. For steepest descent, this becomes
  $1 / (1 + \lVert g \rVert_2^2)$.
- `"scipy"`: $-\lVert g \rVert_2 / \phi'(0)$. For steepest descent, where
  $p = -g$, this reduces to $1 / \lVert g \rVert_2$.
- `"schmidt"`: $1 / \lVert g \rVert_1$; there is no added 1 in the
  denominator.
- `"hz"`: Hager-Zhang scaling based first on the relative infinity norms of
  the parameters and gradient, then on the objective and squared gradient norm,
  with a safeguarded fallback.

</details>

After the first iteration, `step_next_init` can use a slope ratio, quadratic
interpolation, the Hager-Zhang `"hz"` rule, or a fixed positive number. The
Hager-Zhang rule may make one objective-only probe to fit a quadratic. It does
so only when a previous step exists and both the local line-search budgets and
the remaining hard budgets leave room for the probe and a subsequent search.
Otherwise it falls back to an enlarged, safeguarded version of the previous
step. A probe is included in `ls_nf`.

`try_newton_step` adjusts a calculated proposal upward, at most to the natural
unit step of 1. Its current default is `TRUE` for `"NEWTON"`, `"PHESS"`,
`"BFGS"`, `"SR1"`, `"L-BFGS"`, and `"TN"`, and `FALSE` otherwise. Set it
explicitly when comparing initialization policies.

### Backtracking, constant, and bold-driver steps

Armijo backtracking repeatedly reduces the candidate until it provides the
decrease controlled by `c1`. With `step_down = NULL`, the implementation uses
cubic interpolation and evaluates function and gradient at candidate points.
A numeric `step_down` instead multiplies the candidate by that factor and uses
function-only candidate evaluations.

```{r armijo-and-fixed-steps}
backtracking_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Backtracking",
  step0 = 1,
  step_down = 0.5,
  max_iter = 10
)

constant_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Constant",
  norm_direction = TRUE,
  step0 = 0.01,
  max_iter = 10
)

bold_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Bold Driver",
  step_down = 0.5,
  step_up = 1.1,
  max_iter = 10
)
```

The constant strategy requires a numeric `step0` and performs no search. With
steepest descent, `norm_direction = TRUE` makes `step0` the parameter-update
length directly.

Bold driver accepts the first tested point with a lower objective, reduces
failed proposals by `step_down`, and initializes the next iteration by
multiplying the last accepted step by `step_up`. Its first proposal is 1;
`step0`, `c1`, `c2`, and `step_up_fun` do not configure that proposal.
`ls_max_fn` is its local callback limit.

For Wolfe and backtracking searches, `ls_max_fn`, `ls_max_gr`, and `ls_max_fg`
limit callbacks within one outer line search. The optimizer budgets `max_fn`,
`max_gr`, and `max_fg` remain hard limits for the entire run. See the
[line-search reference](../reference/mize.html#line-search) for specialized
safeguards and defaults.

## Momentum and adaptive learning-rate methods

The trajectory comparisons below use a fixed outer-iteration budget.
`abs_tol = 0` asks the high-level loop to observe the objective at every
iteration so the tables and plots describe the accepted trajectory; the other
ordinary tolerances are disabled. They illustrate one problem and do not
establish convergence or a generally superior configuration. Comparison tables
report both best and last objectives because momentum trajectories may be
nonmonotone.

```{r bounded-comparison-helpers, include = FALSE}
bounded_controls <- list(
  max_iter = 100,
  abs_tol = 0,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL,
  store_progress = TRUE
)

run_bounded <- function(method, ...) {
  do.call(
    mize,
    c(list(par = rb0, fg = rb_fg, method = method), list(...), bounded_controls)
  )
}

summarize_runs <- function(results) {
  rows <- lapply(seq_along(results), function(i) {
    result <- results[[i]]
    data.frame(
      Run = names(results)[[i]],
      `Best objective` = result$best_f,
      `Last objective` = result$last_f,
      `Function callbacks` = result$nf,
      `Gradient callbacks` = result$ng,
      Status = result$status,
      Termination = result$terminate$what,
      check.names = FALSE
    )
  })
  summary <- do.call(rbind, rows)
  rownames(summary) <- NULL
  summary
}

assert_bounded_runs <- function(results) {
  summary <- summarize_runs(results)
  stopifnot(
    all(summary$Status == "budget_exhausted"),
    all(summary$Termination == "max_iter"),
    all(is.finite(summary[["Best objective"]])),
    all(is.finite(summary[["Last objective"]])),
    all(summary[["Best objective"]] <= summary[["Last objective"]])
  )
}

plot_current_objective <- function(results) {
  traces <- lapply(results, function(result) {
    keep <- is.finite(result$progress$nf) &
      is.finite(result$progress$f) & result$progress$f > 0
    result$progress[keep, c("nf", "f"), drop = FALSE]
  })
  stopifnot(all(vapply(traces, nrow, integer(1)) > 0L))

  colors <- rep(
    c("#000000", "#0072B2", "#D55E00", "#009E73", "#CC79A7"),
    length.out = length(traces)
  )
  line_types <- rep(c(1, 2, 3, 4, 5), length.out = length(traces))
  plot(
    NA_real_,
    NA_real_,
    xlim = range(unlist(lapply(traces, `[[`, "nf"))),
    ylim = range(unlist(lapply(traces, `[[`, "f"))),
    log = "y",
    xlab = "Cumulative function callbacks",
    ylab = "Current objective (log scale)"
  )
  grid()
  for (i in seq_along(traces)) {
    lines(traces[[i]]$nf, traces[[i]]$f, col = colors[[i]], lty = line_types[[i]])
  }
  legend(
    "topright",
    legend = names(results),
    col = colors,
    lty = line_types,
    bty = "n",
    cex = 0.75,
    ncol = if (length(results) > 3L) 2 else 1
  )
}
```

### Nesterov Accelerated Gradient (`"NAG"`)

NAG takes a steepest-descent step and then applies a momentum update. Its
current objective can increase on an iteration, so inspect the full trajectory
and the returned best result.

`nest_q` controls the built-in NAG schedule. At `nest_q = 0`, the
non-strongly-convex limit gives the largest momentum; at `nest_q = 1`, momentum
is zero and the parameter updates reduce to steepest descent. The default is
zero. A value between the endpoints is a tuning choice unless the objective
supplies the strong-convexity information needed to motivate it.

```{r nag-q-comparison, fig.width = 7, fig.height = 4.5, fig.cap = "Current objective against cumulative function callbacks for two NAG schedules. Both runs use the same start and fixed 100-iteration budget.", fig.alt = "Log-scale line plot comparing current objective over function callbacks for NAG with q equal to zero and q equal to 0.001."}
nag_runs <- list(
  `q = 0` = run_bounded("NAG", nest_q = 0),
  `q = 0.001` = run_bounded("NAG", nest_q = 0.001)
)
assert_bounded_runs(nag_runs)
knitr::kable(summarize_runs(nag_runs), digits = 6)
plot_current_objective(nag_runs)
```

The objective paths are nonmonotone, so the final value does not summarize
every iterate a run visited. Choose either the best or last objective before
the comparison and use that metric consistently.

The theoretical step-size and momentum guarantees require assumptions and
problem constants that this Rosenbrock example does not supply. For derivations
and historical context, see the
[Nesterov background note](https://jlmelville.github.io/mize/articles/nesterov.html).

### Momentum schedules and update order

With `method = "Momentum"`, `mom_schedule` defines the momentum fraction:

| Value | Schedule |
|---|---|
| A number such as `0.9` | Constant momentum |
| `"switch"` | Change from `mom_init` to `mom_final` at `mom_switch_iter` |
| `"ramp"` | Increase linearly from `mom_init` to `mom_final` |
| `"nsconvex"` | Use the NAG schedule, controlled by the `nest_*` arguments |
| A function | Compute a scalar from the iteration, optionally also `max_iter` |

A deterministic custom schedule is easiest to inspect independently of an
optimizer trajectory:

```{r deterministic-momentum-schedule}
cosine_schedule <- function(iter, max_iter) {
  phase <- min(iter / max_iter, 1)
  0.4 + 0.2 * (1 - cos(pi * phase))
}

schedule_values <- data.frame(
  Iteration = c(1, 5, 10),
  Momentum = vapply(
    c(1, 5, 10),
    cosine_schedule,
    numeric(1),
    max_iter = 10
  )
)
knitr::kable(schedule_values, digits = 3)
```

The values show the intended smooth increase over the ten-iteration schedule.
`mize` clamps custom schedules to the supported momentum range. Prefer a
built-in schedule when it already expresses the intended experiment.

`mom_type = "classical"` applies momentum after the gradient step;
`mom_type = "nesterov"` applies it before that step. The next comparison holds
the schedule, start, objective, and budget fixed.

```{r momentum-type-comparison, fig.width = 7, fig.height = 4.5, fig.cap = "Current objective for classical and Nesterov-style momentum under one fixed schedule and 100-iteration budget.", fig.alt = "Log-scale line plot comparing current objective over function callbacks for classical and Nesterov-style momentum."}
momentum_runs <- list(
  Classical = run_bounded("Momentum", mom_schedule = 0.9),
  Nesterov = run_bounded(
    "Momentum",
    mom_schedule = 0.9,
    mom_type = "nesterov"
  )
)
assert_bounded_runs(momentum_runs)
knitr::kable(summarize_runs(momentum_runs), digits = 6)
plot_current_objective(momentum_runs)
```

The table and trajectories show how update order changes this bounded run.
Treat both orders as choices to test on the problem at hand. With
`mom_schedule = "nsconvex"`, `nest_convex_approx = TRUE` selects Sutskever's
approximation for the `nest_q = 0` case and causes `nest_q` to be ignored.

### Delta-Bar-Delta learning rates

`method = "DBD"` changes a per-coordinate learning rate according to agreement
in sign between the current gradient and a weighted history. Because DBD
chooses both direction scaling and step size, it ignores `line_search`.

The main controls are:

- `step0`, the initial per-coordinate learning rate;
- `step_down`, the multiplicative decrease when signs disagree;
- `step_up`, the increase when signs agree;
- `step_up_fun`, either `"*"` for a proportional increase or `"+"` for an
  additive increase; and
- `dbd_weight`, the gradient-history weight when no momentum schedule is used.

The following configuration uses the additive update associated with the
t-SNE-style variant. It demonstrates control semantics; adopting it for a
smooth optimization problem requires problem-specific evidence.

```{r additive-dbd}
dbd_res <- mize(
  rb0,
  rb_fg,
  method = "DBD",
  step0 = "rasmussen",
  step_down = 0.8,
  step_up = 0.2,
  step_up_fun = "+",
  dbd_weight = 0.5,
  max_iter = 10
)
```

If a momentum schedule is supplied with DBD, `dbd_weight` is ignored and the
momentum update supplies the history. DBD's step calculation is gradient-based,
although function-based convergence criteria and best-result reporting may
still request objective evaluations. Budget explicitly for the observations
selected by the convergence controls.

### Adaptive restart

[O'Donoghue and Candes](https://arxiv.org/abs/1204.3982) proposed restarting
accelerated schemes when momentum appears unhelpful. A restart discards the
stored momentum update and resets iteration-dependent momentum schedules. In
`mize`, it can be applied to any optimizer configuration that uses momentum.

| `restart` | Trigger |
|---|---|
| `"fn"` | The objective does not decrease on consecutive iterations |
| `"gr"` | The proposed update is not a descent direction |
| `"speed"` | The update vector does not grow in Euclidean norm |

Gradient restart is unavailable with `mom_type = "nesterov"` because its
gradient is evaluated at a different location. `restart_wait` specifies the
number of iterations after a restart before another restart is allowed; its
default is 10.

The comparison below uses classical momentum and varies only restart type and
waiting period. The no-restart result is reused from the preceding comparison.

```{r adaptive-restart-comparison, fig.width = 7, fig.height = 4.5, fig.cap = "Current objective for classical momentum with no restart and four adaptive-restart configurations under the same 100-iteration budget.", fig.alt = "Log-scale line plot comparing current objective over function callbacks for momentum without restart and with function- or gradient-based restart at waiting periods of ten or one iterations."}
restart_runs <- list(
  None = momentum_runs[["Classical"]],
  `fn, wait = 10` = run_bounded(
    "Momentum",
    mom_schedule = 0.9,
    restart = "fn",
    restart_wait = 10
  ),
  `gr, wait = 10` = run_bounded(
    "Momentum",
    mom_schedule = 0.9,
    restart = "gr",
    restart_wait = 10
  ),
  `fn, wait = 1` = run_bounded(
    "Momentum",
    mom_schedule = 0.9,
    restart = "fn",
    restart_wait = 1
  ),
  `gr, wait = 1` = run_bounded(
    "Momentum",
    mom_schedule = 0.9,
    restart = "gr",
    restart_wait = 1
  )
)
assert_bounded_runs(restart_runs)
knitr::kable(summarize_runs(restart_runs), digits = 6)
plot_current_objective(restart_runs)
```

The table places trajectory summaries beside callback demand. Function-based
validation may add an objective callback when the value is not cached; gradient
validation has the same relationship to available gradients. Compare both
callback columns and tune restart policy under the budget and metric that
matter for the application.

## See also

- [Getting started](mize.html) provides the shortest route to a first result.
- [Convergence](convergence.html) explains the status and termination fields
  used in every bounded comparison.
- [Nesterov accelerated gradient and
  momentum](https://jlmelville.github.io/mize/articles/nesterov.html) develops
  the schedule derivations and historical background.
- [Quasi-hyperbolic
  momentum](https://jlmelville.github.io/mize/articles/qhm.html) covers another
  momentum formulation.
