---
title: "Choosing methods and tuning"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing methods and tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)

rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  },
  fg = function(x) {
    a <- x[2] - x[1] * x[1]
    b <- 1 - x[1]
    list(
      fn = 100 * a^2 + b^2,
      gr = c(-400 * x[1] * a - 2 * b, 200 * a)
    )
  }
)
rb0 <- c(-1.2, 1)
```

This guide collects the advanced method, line-search, step-size, momentum, and
restart options available in `mize`. Start with [Getting started](mize.html)
for the basic optimization workflow, then use this guide when choosing or
tuning an optimization strategy.

## Choosing a method

`mize()` defaults to `"L-BFGS"`, a good starting point for many smooth
problems when an analytic gradient is available. Change methods for a reason:
because the problem size changes the storage budget, Hessian information is
available, or the experiment is specifically about another update rule.

| Need | Suggested starting point | Main trade-off |
|---|---|---|
| General smooth problem | `"L-BFGS"` | Limited-memory curvature model; usually the first method to try |
| Small or medium smooth problem | `"BFGS"`; consider `"SR1"` for comparison | Dense inverse-Hessian state uses quadratic storage |
| Reliable Hessian or inverse Hessian available | `"NEWTON"` | Strong local model; Hessian construction and factorization can dominate |
| Large Hessian-free problem | `"TN"` | Approximates a Newton direction with an inner solve and extra gradient work |
| Specialized nonlinear conjugate-gradient study | `"CG"` | Low storage; update choice, scaling, and line search matter |
| Momentum experiment | `"NAG"` or `"Momentum"` | Requires a defined schedule and careful comparison of nonmonotone trajectories |
| Known or deliberately controlled learning rate | `"SD"` | Simple baseline; performance depends heavily on the step-size strategy |
| Per-coordinate adaptive learning rate | `"DBD"` | Heuristic update controls both direction scaling and step size |

For algorithmic background, see
[Nocedal and Wright](https://doi.org/10.1007/978-0-387-40065-5). The choices
above are starting points. Benchmark promising methods on the objective at
hand.

### Current method surface

- `"L-BFGS"` maintains its limited-memory approximation from the most recent
  `memory` curvature pairs. Increasing `memory` retains more history and costs
  more storage.
- `"BFGS"` maintains a dense inverse-Hessian approximation, so its storage is
  quadratic in the number of parameters.
- `"SR1"` is a dense symmetric-rank-one alternative. Its update is safeguarded,
  and `mize` falls back to a BFGS or steepest-descent direction when necessary
  to obtain descent.
- `"NEWTON"` uses `fg$hs` or `fg$hi`. A Hessian or inverse Hessian may be a
  full matrix or a diagonal vector; unsuccessful or non-descent directions
  fall back to steepest descent and can be inspected through
  `direction_reason`.
- `"TN"` (truncated Newton) approximately solves for a Newton direction using
  gradient differences. `tn_init`, `tn_exit`, and the optional L-BFGS
  `preconditioner` control its inner solve.
- `"CG"` is nonlinear conjugate gradient. `cg_update = "PR+"` is the default;
  `"HZ+"` is another restart-enabled choice. The reference topic lists the
  complete update vocabulary.
- `"SD"` is steepest descent. It is useful as a baseline and with a deliberately
  chosen constant or adaptive step size. An arbitrary step can increase the
  objective.
- `"NAG"` is Nesterov accelerated gradient. Its momentum depends on `nest_q`
  and related controls. Compare bounded runs with explicit metrics.
- `"Momentum"` (which may be abbreviated to `"MOM"`) combines steepest descent
  with a classical or Nesterov-style momentum stage and a chosen schedule.
- `"DBD"` is Delta-Bar-Delta. It owns a per-coordinate learning rate, so its
  step controls replace the requested `line_search`.
- `"PHESS"` reuses a partial-Hessian factorization. The factory accepts this
  experimental/internal method when `fg$hs` is available. For ordinary work,
  choose a starting method from the table above.

```{r method-surface-check, include = FALSE}
documented_methods <- c(
  "SD", "BFGS", "SR1", "L-BFGS", "CG", "TN", "NEWTON", "PHESS", "NAG",
  "Momentum", "DBD"
)
stopifnot(all(vapply(
  documented_methods,
  function(method) is.list(make_mize(method = method)),
  logical(1)
)))
```

Each method-specific control has a fixed scope. For example, `memory` applies
to L-BFGS and to the optional L-BFGS preconditioner for CG or TN;
`cg_update` applies only to CG; and `tn_init` and `tn_exit` apply only to TN.
Other methods ignore those controls. The same rule applies to step-size
controls: DBD ignores `line_search`, and Wolfe-only controls are ignored by
constant and bold-driver steps. See [`?mize`](../reference/mize.html) for the
complete accepted values and control scopes.

### Nesterov Accelerated Gradient (`"NAG"`)

NAG takes a steepest-descent step and then applies a momentum update. Its
current objective can increase on an iteration. Inspect the full bounded
trajectory and the returned best result.

`nest_q` controls the built-in NAG schedule. The endpoints are important:
`nest_q = 0` is the non-strongly-convex limit and gives the largest momentum,
whereas `nest_q = 1` gives zero momentum and reduces the parameter updates to
steepest descent. The default is zero. Treat a value between the endpoints as a
tuning choice. A known strong-convexity ratio requires corresponding
information about the objective.

Each comparison reports best and last objectives, callback counts, status, and
termination reason, then plots the same result objects against cumulative
function callbacks.

```{r bounded-comparison-helpers, include = FALSE}
summarize_runs <- function(results) {
  rows <- lapply(seq_along(results), function(i) {
    result <- results[[i]]
    data.frame(
      Run = names(results)[[i]],
      `Best objective` = result$best_f,
      `Last objective` = result$last_f,
      `Function callbacks` = result$nf,
      `Gradient callbacks` = result$ng,
      Status = result$status,
      Termination = result$terminate$what,
      check.names = FALSE
    )
  })
  summary <- do.call(rbind, rows)
  rownames(summary) <- NULL
  summary
}

assert_bounded_runs <- function(results) {
  summary <- summarize_runs(results)
  stopifnot(
    all(summary$Status == "budget_exhausted"),
    all(summary$Termination == "max_iter"),
    all(is.finite(summary[["Best objective"]])),
    all(is.finite(summary[["Last objective"]])),
    all(summary[["Best objective"]] <= summary[["Last objective"]])
  )
}

plot_current_objective <- function(results, title) {
  traces <- lapply(results, function(result) {
    keep <- is.finite(result$progress$nf) &
      is.finite(result$progress$f) & result$progress$f > 0
    result$progress[keep, c("nf", "f"), drop = FALSE]
  })
  stopifnot(all(vapply(traces, nrow, integer(1)) > 0L))

  colors <- rep(
    c("#000000", "#0072B2", "#D55E00", "#009E73", "#CC79A7"),
    length.out = length(traces)
  )
  line_types <- rep(c(1, 2, 3, 4, 5), length.out = length(traces))
  plot(
    NA_real_,
    NA_real_,
    xlim = range(unlist(lapply(traces, `[[`, "nf"))),
    ylim = range(unlist(lapply(traces, `[[`, "f"))),
    log = "y",
    xlab = "Cumulative function calls",
    ylab = "Current objective (log scale)",
    main = title
  )
  grid()
  for (i in seq_along(traces)) {
    lines(traces[[i]]$nf, traces[[i]]$f, col = colors[[i]], lty = line_types[[i]])
  }
  legend(
    "topright",
    legend = names(results),
    col = colors,
    lty = line_types,
    bty = "n",
    cex = 0.75,
    ncol = if (length(results) > 3L) 2 else 1
  )
}
```

```{r nag-q-comparison, fig.width = 7, fig.height = 4.5}
nag_runs <- list(
  `q = 0` = mize(
    rb0, rb_fg, method = "NAG", nest_q = 0,
    max_iter = 100, store_progress = TRUE
  ),
  `q = 0.001` = mize(
    rb0, rb_fg, method = "NAG", nest_q = 0.001,
    max_iter = 100, store_progress = TRUE
  )
)
assert_bounded_runs(nag_runs)
knitr::kable(summarize_runs(nag_runs), digits = 6)
plot_current_objective(nag_runs, "NAG and nest_q")
```

```{r nag-order-check, include = FALSE}
stopifnot(
  nag_runs[["q = 0.001"]]$last_f < nag_runs[["q = 0"]]$last_f,
  nag_runs[["q = 0"]]$best_f < nag_runs[["q = 0.001"]]$best_f
)
```

Both rows stop at `max_iter` and demonstrate progress within a fixed outer
iteration budget. Convergence remains unestablished. Here `q = 0.001` has the
lower last objective, while `q = 0` has the lower returned best objective. The plot shows
current iterates; the table reports both best and last objectives. The
conflicting order is the lesson: define the metric before calling one bounded
trajectory “better.”

The theoretical step size and momentum guarantees require assumptions and
problem constants that this Rosenbrock example does not supply. For derivations
and historical context, see the [Nesterov background note](nesterov.html).

### Classical Momentum (`"Momentum"`)

Classical momentum adds a fraction of the previous update to a
steepest-descent update. `mom_schedule` defines that fraction:

| Value | Schedule |
|---|---|
| A number such as `0.9` | Constant momentum |
| `"switch"` | Change from `mom_init` to `mom_final` at `mom_switch_iter` |
| `"ramp"` | Increase linearly from `mom_init` to `mom_final` |
| `"nsconvex"` | Use the NAG schedule, controlled by the `nest_*` arguments |
| A function | Compute a scalar from the iteration, optionally also `max_iter` |

A custom schedule can be deterministic when randomness adds no lesson:

```{r deterministic-momentum-schedule}
cosine_schedule <- function(iter, max_iter) {
  phase <- min(iter / max_iter, 1)
  0.4 + 0.2 * (1 - cos(pi * phase))
}
custom_momentum <- mize(
  rb0, rb_fg, method = "Momentum", mom_schedule = cosine_schedule,
  max_iter = 10
)
```

The schedule is clamped to the supported momentum range. Built-in schedules
are preferable when they express the intended experiment directly.

### Classical and Nesterov-style momentum

With `method = "Momentum"`, `mom_type = "classical"` applies momentum after
the gradient step; `mom_type = "nesterov"` applies it before that step. The
schedule remains an independent choice. The next comparison holds the
schedule, start, objective, and iteration budget fixed.

```{r momentum-type-comparison, fig.width = 7, fig.height = 4.5}
momentum_runs <- list(
  Classical = mize(
    rb0, rb_fg, method = "Momentum", mom_schedule = 0.9,
    max_iter = 100, store_progress = TRUE
  ),
  Nesterov = mize(
    rb0, rb_fg, method = "Momentum", mom_schedule = 0.9,
    mom_type = "nesterov", max_iter = 100, store_progress = TRUE
  )
)
assert_bounded_runs(momentum_runs)
knitr::kable(summarize_runs(momentum_runs), digits = 6)
plot_current_objective(momentum_runs, "Momentum update order")
```

These runs also exhaust `max_iter`. Their table records what happened under
this configuration. Treat the two update orders as choices to test on the
problem at hand. Use `mom_schedule = "nsconvex"` to combine either order with
the NAG schedule. With that schedule, `nest_convex_approx = TRUE` selects Sutskever's
approximation for the `nest_q = 0` case and causes `nest_q` to be ignored.

## Line searches and step sizes

Most methods first choose a direction and then use `line_search` to choose how
far to move. DBD is the main exception: it chooses per-coordinate learning
rates itself and ignores `line_search`.

| Choice | What it accepts | Useful starting context |
|---|---|---|
| `"More-Thuente"` | Strong Wolfe conditions by default | Default for direction methods, including quasi-Newton and CG |
| `"Rasmussen"` or `"Schmidt"` | Strong Wolfe conditions by default | Alternative Wolfe implementations for comparison or troubleshooting |
| `"Hager-Zhang"` | Standard curvature plus approximate Armijo by default | CG-oriented Wolfe search with Hager-Zhang initializers |
| `"Backtracking"` | Armijo sufficient decrease | Simpler search when a curvature condition is unnecessary |
| `"Bold Driver"` | Any finite objective decrease | Heuristic adaptation from the preceding accepted step |
| `"Constant"` | No search; always uses numeric `step0` | Known learning rate or a controlled experiment |

Each line-search control has a fixed scope. `c2`, `strong_curvature`, and
`approx_armijo` apply only to Wolfe searches. `c1`
also applies to Armijo backtracking. Bold driver and constant steps ignore it.
`ls_max_alpha` and `ls_safe_cubic` are More-Thuente-specific. The DBD controls
are described separately below.

### Wolfe conditions and `c2`

For a direction $p$, define the line function
$\phi(\alpha) = f(x + \alpha p)$ and assume the initial slope
$\phi'(0) = g(x)^T p$ is negative. Wolfe searches require sufficient decrease
and a curvature condition. Under the strong curvature condition,

$$
|\phi'(\alpha)| \leq c_2 |\phi'(0)|.
$$

A smaller `c2` makes the right-hand side smaller, so it is **stricter**; a
larger `c2` is looser. The same direction holds for the standard curvature
condition because the initial slope is negative. `c2` must lie strictly
between `c1` and 1.

The CG default is `c2 = 0.1`. The next bounded example sets `c2 = 0.5`, making
the curvature test looser. It also selects the Hager-Zhang next-step initializer.
The table displays the resulting initial proposals, slopes, line-search
outcomes, and callback counts.

```{r looser-c2-diagnostics}
looser_c2 <- mize(
  rb0,
  rb_fg,
  method = "CG",
  cg_update = "HZ+",
  line_search = "More-Thuente",
  c2 = 0.5,
  step_next_init = "hz",
  max_iter = 4,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL,
  store_progress = TRUE
)

line_search_columns <- c(
  "alpha_init", "slope_init", "ls_reason", "ls_outcome", "ls_nf", "ls_ng"
)
line_search_diagnostics <- looser_c2$progress[
  !is.na(looser_c2$progress$ls_reason),
  line_search_columns,
  drop = FALSE
]
rownames(line_search_diagnostics) <- NULL
line_search_diagnostics
```

`alpha_init` is the safeguarded starting proposal and `slope_init` is
$\phi'(0)$. `ls_reason` says why the search stopped, while `ls_outcome` says
what kind of point was ultimately selected; the two need not be identical.
`ls_nf` and `ls_ng` count callbacks owned by that outer line search, excluding
its cached starting point. These columns are method-specific and appear only
when their owning stage supplies them.

```{r looser-c2-check, include = FALSE}
stopifnot(
  identical(looser_c2$status, "budget_exhausted"),
  identical(looser_c2$terminate$what, "max_iter"),
  identical(names(line_search_diagnostics), line_search_columns),
  nrow(line_search_diagnostics) > 0,
  all(is.finite(line_search_diagnostics$alpha_init)),
  all(line_search_diagnostics$alpha_init > 0),
  all(is.finite(line_search_diagnostics$slope_init)),
  all(line_search_diagnostics$slope_init < 0),
  all(line_search_diagnostics$ls_nf >= 0),
  all(line_search_diagnostics$ls_ng >= 0)
)
```

More-Thuente, Rasmussen, and Schmidt use strong curvature with the exact Armijo
condition by default. Hager-Zhang uses the standard curvature condition with
the approximate Armijo condition by default. `strong_curvature` and
`approx_armijo` can override those choices. An override changes the conditions
whose theory applies; treat it as an explicit experiment.

### Initial step estimates

On the first outer iteration, the named `step0` choices use the starting
gradient $g$ and direction $p$. Writing the directional derivative as
$\phi'(0) = g^T p$, the current formulas are:

- `"rasmussen"`: $1 / (1 - \phi'(0))$. For steepest descent, this becomes
  $1 / (1 + \lVert g \rVert_2^2)$.
- `"scipy"`: $-\lVert g \rVert_2 / \phi'(0)$. For steepest descent, where
  $p = -g$, this reduces to $1 / \lVert g \rVert_2$.
- `"schmidt"`: $1 / \lVert g \rVert_1$; there is no added 1 in the
  denominator.
- `"hz"`: Hager-Zhang scaling based first on the relative infinity norms of
  the parameters and gradient, then on the objective and squared gradient norm,
  with a safeguarded fallback.

A positive finite numeric `step0` bypasses these formulas. Hager-Zhang uses its
own first-step estimator by default; the other gradient searches default to
Rasmussen.

After the first iteration, `step_next_init` can use a slope ratio, quadratic
interpolation, the Hager-Zhang `"hz"` rule, or a fixed positive number.
Hager-Zhang's next-step rule may make one objective-only probe to fit a
quadratic. The probe is attempted only when there is a previous step and both
the local line-search budgets and the optimizer's remaining hard budgets leave
enough callbacks for the probe and a subsequent search. Otherwise the rule
falls back to an enlarged, safeguarded version of the previous step. When the
probe occurs, it is included in `ls_nf`, as in the diagnostic vocabulary above.

`try_newton_step` gives the calculated proposal a small upward adjustment, up
to the natural unit step of 1. Its current default is `TRUE` for `"NEWTON"`,
`"PHESS"`, `"BFGS"`, `"SR1"`, `"L-BFGS"`, and `"TN"`, and `FALSE` otherwise.
`"PHESS"` remains experimental/internal despite sharing this default. Set
`try_newton_step` explicitly when comparing initialization policies.

### Backtracking, constant, and bold-driver steps

Armijo backtracking repeatedly reduces the candidate until it provides the
decrease controlled by `c1`. With `step_down = NULL`, the current implementation
uses cubic interpolation and evaluates function and gradient at candidate
points. A numeric `step_down` instead multiplies the candidate by that fixed
factor and uses function-only candidate evaluations.

```{r armijo-and-fixed-steps}
backtracking_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Backtracking",
  step0 = 1,
  step_down = 0.5,
  max_iter = 10
)

constant_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Constant",
  norm_direction = TRUE,
  step0 = 0.01,
  max_iter = 10
)

bold_res <- mize(
  rb0,
  rb_fg,
  method = "SD",
  line_search = "Bold Driver",
  step_down = 0.5,
  step_up = 1.1,
  max_iter = 10
)
```

The constant strategy requires a numeric `step0` and performs no search. With
steepest descent, `norm_direction = TRUE` makes `step0` the parameter-update
length directly.

Bold driver accepts the first tested point with a lower objective, reduces
failed proposals by `step_down`, and initializes the next iteration by
multiplying the last accepted step by `step_up`. In the current implementation
the first bold-driver proposal is 1; `step0`, `c1`, `c2`, and `step_up_fun` do
not configure it. `ls_max_fn` is its local callback limit.

For Wolfe and backtracking searches, `ls_max_fn`, `ls_max_gr`, and `ls_max_fg`
limit callbacks within one outer line search. The optimizer budgets `max_fn`,
`max_gr`, and `max_fg` remain hard limits for the entire run. The default local
function limit is 20. See the
[line-search reference](../reference/mize.html#line-search) for the remaining
specialized safeguards.

## Delta-Bar-Delta learning rates

`method = "DBD"` changes a per-coordinate learning rate from the agreement in
sign between the current gradient and a weighted history. Because DBD chooses
both the direction scaling and the step, `line_search` is ignored.

The main controls are:

- `step0`, the initial per-coordinate learning rate (a positive number or a
  named first-step estimator);
- `step_down`, the multiplicative decrease when signs disagree;
- `step_up`, the increase when signs agree;
- `step_up_fun`, either `"*"` for a proportional increase (the default) or
  `"+"` for an additive increase; and
- `dbd_weight`, the gradient-history weight when no momentum schedule is used.

The following uses the additive update associated with the t-SNE-style
variant. It demonstrates the control semantics. Use problem-specific evidence
before adopting it for smooth optimization.

```{r additive-dbd}
dbd_res <- mize(
  rb0,
  rb_fg,
  method = "DBD",
  step0 = "rasmussen",
  step_down = 0.8,
  step_up = 0.2,
  step_up_fun = "+",
  dbd_weight = 0.5,
  max_iter = 10
)
```

If a momentum schedule is supplied with DBD, `dbd_weight` is ignored and the
momentum update supplies the history. DBD's step calculation itself is
gradient-based. Function-based convergence criteria still request objective
evaluations, and `mize()` may need an objective observation to report a best
result. Choose convergence controls according to the observations the run can
afford, and budget for those observations explicitly.

## Adaptive restart

[O'Donoghue and Candes](https://arxiv.org/abs/1204.3982) proposed restarting
accelerated schemes when momentum appears unhelpful. A restart discards the
stored momentum update and resets iteration-dependent momentum schedules. In
`mize`, it can be applied to any optimizer configuration that uses momentum.

| `restart` | Trigger |
|---|---|
| `"fn"` | The objective does not decrease on consecutive iterations |
| `"gr"` | The proposed update is not a descent direction |
| `"speed"` | The update vector does not grow in Euclidean norm |

Gradient restart is unavailable with `mom_type = "nesterov"` because its
gradient is evaluated at a different location. `restart_wait` is the number of
iterations to wait after a restart before another restart is allowed; its
default is 10. `restart_wait` governs restart policy; termination controls
govern convergence.

The following comparison uses classical momentum so these restart
configurations are valid. It holds the start, schedule, and 100-iteration
budget fixed, then varies restart type and wait. The no-restart result is
reused from the preceding comparison.

```{r adaptive-restart-comparison, fig.width = 7, fig.height = 4.5}
restart_runs <- list(
  None = momentum_runs[["Classical"]],
  `fn, wait = 10` = mize(
    rb0, rb_fg, method = "Momentum", mom_schedule = 0.9,
    restart = "fn", restart_wait = 10,
    max_iter = 100, store_progress = TRUE
  ),
  `gr, wait = 10` = mize(
    rb0, rb_fg, method = "Momentum", mom_schedule = 0.9,
    restart = "gr", restart_wait = 10,
    max_iter = 100, store_progress = TRUE
  ),
  `fn, wait = 1` = mize(
    rb0, rb_fg, method = "Momentum", mom_schedule = 0.9,
    restart = "fn", restart_wait = 1,
    max_iter = 100, store_progress = TRUE
  ),
  `gr, wait = 1` = mize(
    rb0, rb_fg, method = "Momentum", mom_schedule = 0.9,
    restart = "gr", restart_wait = 1,
    max_iter = 100, store_progress = TRUE
  )
)
assert_bounded_runs(restart_runs)
knitr::kable(summarize_runs(restart_runs), digits = 6)
plot_current_objective(restart_runs, "Adaptive restart")
```

```{r restart-order-check, include = FALSE}
fn_wait_10 <- restart_runs[["fn, wait = 10"]]
gr_wait_10 <- restart_runs[["gr, wait = 10"]]
fn_wait_1 <- restart_runs[["fn, wait = 1"]]
gr_wait_1 <- restart_runs[["gr, wait = 1"]]
stopifnot(
  fn_wait_1$best_f < fn_wait_10$best_f,
  fn_wait_1$last_f < fn_wait_10$last_f,
  gr_wait_1$best_f < gr_wait_10$best_f,
  gr_wait_1$last_f < gr_wait_10$last_f,
  gr_wait_1$best_f < fn_wait_1$best_f,
  gr_wait_1$last_f < fn_wait_1$last_f,
  gr_wait_1$nf > fn_wait_1$nf,
  gr_wait_1$ng > fn_wait_1$ng
)
```

Every row again reports `budget_exhausted` / `max_iter`. These runs demonstrate
bounded progress and leave convergence unestablished. Both wait-1 rows reach smaller best
and last objectives than their wait-10 counterparts. The gradient/wait-1 row
has slightly smaller objective values than the function/wait-1 row and uses
more function and gradient callbacks. These conclusions apply to the displayed
bounded run.

Function-based validation can add an objective callback when the needed value
is not cached; gradient validation likewise depends on where gradients are
already available. Compare both callback columns; cached observations determine
the cost of each trigger. A shorter `restart_wait` permits more frequent
restarts and can change both progress and callback demand, so tune it under the
budget and metric that matter for the application.

## See also

- [Getting started](mize.html) provides the shortest route to a first result.
- [Convergence](convergence.html) explains the status and termination fields
  used in every bounded comparison.
- [Nesterov accelerated gradient and momentum](nesterov.html) develops the
  schedule derivations and historical background.
- [Quasi-hyperbolic momentum](qhm.html) covers another momentum formulation.
