---
title: "Mize"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mize}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

`mize` performs unconstrained numerical optimization when you can provide an
objective function and its gradient. This guide follows one complete first-use
path: define the problem, check the gradient, run the optimizer, and interpret
the result. It ends with pointers to the specialist guides for convergence,
stateful use, and method tuning.

## Define the objective and gradient

`mize()` receives the objective and gradient together in a list. The `fn`
function must return one numeric value; `gr` must return a numeric vector with
the same length as its input. Here is the two-dimensional Rosenbrock function:

```{r define-rosenbrock}
rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  }
)
```

For a more involved problem, a factory function can return this list so that
`fn` and `gr` share data through a closure. The list may also contain an `fg`
function that computes both values together when they share expensive work.
The [Metric MDS](mmds.html) article develops that pattern in a realistic
example.

## Choose a starting point and check the gradient

The parameters are a numeric vector. Before optimizing, check an analytic
gradient at a representative, non-special starting point:

```{r check-gradient}
rb0 <- c(-1.2, 1)
gradient_check <- check_mize_gradient(rb_fg, rb0)

gradient_summary <- data.frame(
  max_abs_error = gradient_check$max_abs_error,
  max_rel_error = gradient_check$max_rel_error
)
gradient_summary
```

`check_mize_gradient()` compares `gr` with a finite-difference approximation
of `fn`. Small errors do not prove that a gradient is correct everywhere, but
large errors are a strong reason to fix it before trusting an optimization.
The full result also identifies the worst coordinate and, when an optional
combined `fg` function is present, checks it against the separate callbacks.

```{r assert-gradient, include=FALSE}
stopifnot(
  is.finite(gradient_check$max_rel_error),
  gradient_check$max_rel_error < 1e-6
)
```

## Run the optimization

The default method is limited-memory BFGS (`"L-BFGS"`), a useful general
starting point for a smooth problem. Set `store_progress = TRUE` here so that
we can inspect the trajectory later:

```{r run-mize}
res <- mize(rb0, rb_fg, store_progress = TRUE)
```

Start by reading the high-level outcome and the detailed termination record:

```{r result-status}
res[c("status", "converged", "message", "terminate")]
```

`status` is the quickest classification:

- `"converged"` means a configured tolerance was reached;
- `"budget_exhausted"` means an iteration or callback limit was reached;
- `"failed"` identifies failures such as a non-finite callback or line-search
  failure; and
- `"terminated"` covers another recognized stopping reason.

`converged` is `TRUE` only for the tolerance-based class, while `message` is a
human-readable summary. Use `terminate$what` when code needs the precise
reason. Its companion `terminate$val` is the value associated with that
stopping criterion. For example, after an absolute-function-tolerance exit it
is the observed change between consecutive objective values. It is not the
objective value itself: that is `res$f` (or, more explicitly, `res$best_f`).

### Best result and last iterate

By default, `mize()` tracks the best evaluated point. `par` and `f` are the
usual concise names for that result; `best_par` and `best_f` make the same
choice explicit. `last_par` and `last_f` describe the final optimizer state
before any earlier best point is restored. They can differ when a method's
trajectory is not monotone.

```{r result-points}
result_points <- data.frame(
  result = c("best", "last"),
  x1 = c(res$best_par[1], res$last_par[1]),
  x2 = c(res$best_par[2], res$last_par[2]),
  objective = c(res$best_f, res$last_f)
)
result_points

res[c("iter", "nf", "ng")]
```

Function values are conditional result components. At an exact hard callback
cap, `mize()` will not exceed the budget merely to evaluate the returned point,
so `f`, `best_f`, or `last_f` can be absent. Check with `is.null(res[["f"]])`
before using a value in budget-sensitive code. The
[Convergence](convergence.html) article explains the status classes, stopping
criteria, hard budgets, and best-versus-last behavior in detail.

The assertions below protect only the stable lessons of this example: the
objective is finite and improved, and a tolerance exit has a coherent summary.

```{r assert-result, include=FALSE}
initial_f <- rb_fg$fn(rb0)
convergence_reasons <- c(
  "abs_tol", "rel_tol", "grad_tol", "ginf_tol", "step_tol"
)

stopifnot(
  is.finite(res$best_f),
  res$best_f < initial_f,
  identical(res$status, "converged"),
  isTRUE(res$converged),
  res$terminate$what %in% convergence_reasons
)
```

## Inspect progress without printing everything

With `store_progress = TRUE`, `res$progress` contains the observations and
diagnostics available at stored iterations. Select the columns useful for the
question at hand and inspect only the beginning and end:

```{r compact-progress}
progress_columns <- intersect(
  c("f", "step", "nf", "ng", "alpha", "ls_outcome"),
  names(res$progress)
)
progress_view <- res$progress[, progress_columns, drop = FALSE]

head(progress_view, 3)
tail(progress_view, 3)
```

The initial row represents the starting point. Here `f` is the current
objective, `step` is the size of the outer parameter update, and `nf` and `ng`
are cumulative callback counts. Columns such as `alpha` and `ls_outcome` come
from a line search. Other methods and settings add different columns, and a
column is omitted when its owning calculation did not produce it. See the
[Progress section of the `mize()` reference](../reference/mize.html#progress)
for the complete dynamic schema.

For live console output, use `verbose = TRUE`; `log_every` controls how often
progress is printed or stored. Keeping every iteration can produce a large
result on a long run.

## Choose a method

The default is a sensible baseline, but problem size and available derivative
information may suggest another starting point:

| Situation | Method to consider | Main trade-off |
|---|---|---|
| General smooth problem | `L-BFGS` | Limited-memory quasi-Newton; the default |
| Smaller smooth problem | `BFGS` | Dense inverse-Hessian approximation uses quadratic storage |
| Very tight memory budget | `CG` | Low storage, but more sensitive to scaling and line-search choices |
| Extra gradient work is acceptable | `TN` | Approximates a Newton step without forming a Hessian |
| Hessian or inverse Hessian is available | `NEWTON` | Requires an `hs` or `hi` callback in the function list |

These are starting points, not a universal ranking. The
[Choosing methods and tuning](methods.html) guide covers the full method set,
compatibility, line searches, step sizes, and momentum.

## Where to go next

- [Convergence](convergence.html) explains tolerances, limits, failures, and
  result interpretation.
- [Stateful optimization](stateful.html) shows how to step, monitor,
  checkpoint, and resume an optimizer.
- [Metric MDS](mmds.html) is a complete case study using closures, a checked
  gradient, and combined callbacks.
- [Choosing methods and tuning](methods.html) covers algorithm selection and
  advanced controls.
