---
title: "Getting started"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

`mize` performs unconstrained numerical optimization when you can provide an
objective function and its gradient. This guide follows one complete first-use
path: define the problem, check the gradient, run the optimizer, and interpret
the result. It ends with pointers to the specialist guides for convergence,
stateful use, and method tuning.

## Define the objective and gradient

`mize()` receives the objective and gradient together in a list. The `fn`
function must return one numeric value; `gr` must return a numeric vector with
the same length as its input. Here is the two-dimensional Rosenbrock function:

```{r define-rosenbrock}
rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  }
)
```

For a more involved problem, a factory function can return this list so that
`fn` and `gr` share data through a closure. The list may also contain an `fg`
function that computes both values together when they share expensive work.
The [Metric MDS](mmds.html) article develops that pattern in a realistic
example.

## Choose a starting point and check the gradient

The parameters are a numeric vector. Before optimizing, check an analytic
gradient at a representative, non-special starting point:

```{r check-gradient}
rb0 <- c(-1.2, 1)
gradient_check <- check_mize_gradient(rb_fg, rb0)

gradient_summary <- data.frame(
  "Maximum absolute error" = formatC(
    gradient_check$max_abs_error,
    format = "e",
    digits = 2
  ),
  "Maximum relative error" = formatC(
    gradient_check$max_rel_error,
    format = "e",
    digits = 2
  ),
  check.names = FALSE
)
knitr::kable(gradient_summary, align = c("r", "r"))
```

`check_mize_gradient()` compares `gr` with a finite-difference approximation
of `fn`. Large errors are a strong reason to fix the gradient before trusting
an optimization. At this point, the analytic and finite-difference gradients
agree to approximately
`r formatC(gradient_check$max_rel_error, format = "e", digits = 1)` in relative
terms. The full result also identifies the worst coordinate and, when an
optional combined `fg` function is present, checks it against the separate
callbacks.

```{r assert-gradient, include=FALSE}
stopifnot(
  is.finite(gradient_check$max_rel_error),
  gradient_check$max_rel_error < 1e-6
)
```

## Run the optimization

The default method is limited-memory BFGS (`"L-BFGS"`), a useful general
starting point for a smooth problem. Set `store_progress = TRUE` here so that
we can inspect the trajectory later:

```{r run-mize}
res <- mize(rb0, rb_fg, store_progress = TRUE)
```

Start with the fitted parameters, objective, and high-level outcome:

```{r result-summary}
res[c("par", "f", "status", "message")]
```

The Rosenbrock function has a known minimum at `c(1, 1)`, where its objective is
zero. The returned parameters are close to that point, the objective is close
to zero, and `status = "converged"` says that a configured numerical tolerance
was reached.

### Read the termination record

`status` gives the quickest classification of any run:

- `"converged"` means a configured tolerance was reached;
- `"budget_exhausted"` means an iteration or callback limit was reached;
- `"failed"` identifies failures such as a non-finite callback or line-search
  failure; and
- `"terminated"` covers another recognized stopping reason.

`converged` is `TRUE` only for the tolerance-based class, while `message` is a
human-readable summary. Use the detailed record when code needs the precise
reason:

```{r result-termination}
res[c("converged", "terminate")]
```

```{r assert-result, include=FALSE}
initial_f <- rb_fg$fn(rb0)
convergence_reasons <- c(
  "abs_tol", "rel_tol", "grad_tol", "ginf_tol", "step_tol"
)

stopifnot(
  is.finite(res$best_f),
  res$best_f < initial_f,
  identical(res$status, "converged"),
  isTRUE(res$converged),
  res$terminate$what %in% convergence_reasons
)
```

## Inspect progress

With `store_progress = TRUE`, `res$progress` contains the observations and
diagnostics available at stored iterations. Plotting the objective against the
cumulative number of function evaluations gives a quick view of the run:

```{r progress-plot, echo=FALSE, fig.width=6.5, fig.height=4.2, fig.cap="Rosenbrock objective values over cumulative function evaluations. The vertical axis uses a logarithmic scale.", fig.alt="Line plot of the Rosenbrock objective decreasing from about 24 to near zero as cumulative function evaluations increase. The objective axis is logarithmic."}
progress_for_plot <- subset(res$progress, is.finite(f) & f > 0)

plot(
  progress_for_plot$nf,
  progress_for_plot$f,
  type = "o",
  pch = 16,
  cex = 0.6,
  log = "y",
  xlab = "Cumulative function evaluations",
  ylab = "Objective"
)

stopifnot(
  nrow(progress_for_plot) > 1,
  tail(progress_for_plot$f, 1) < progress_for_plot$f[1]
)
```

The stored objective falls by several orders of magnitude during this run.
For exact values, select the fields useful for the question and inspect a few
rows from the beginning and end:

```{r compact-progress}
progress_view <- res$progress[, c("f", "step", "nf", "ng"), drop = FALSE]
```

```{r progress-table, echo=FALSE}
progress_rows <- unique(c(
  seq_len(min(3L, nrow(progress_view))),
  seq.int(max(1L, nrow(progress_view) - 2L), nrow(progress_view))
))
progress_excerpt <- progress_view[progress_rows, , drop = FALSE]

progress_table <- data.frame(
  Iteration = as.integer(rownames(progress_excerpt)),
  Objective = trimws(formatC(progress_excerpt$f, format = "g", digits = 4)),
  Step = trimws(formatC(progress_excerpt$step, format = "g", digits = 4)),
  "Function calls" = progress_excerpt$nf,
  "Gradient calls" = progress_excerpt$ng,
  check.names = FALSE
)
knitr::kable(
  progress_table,
  align = rep("r", ncol(progress_table))
)
```

The initial row represents the starting point. Here `f` is the current
objective, `step` is the size of the outer parameter update, and `nf` and `ng`
are cumulative callback counts. Other methods and settings can add fields such
as `alpha` and `ls_outcome`; a field is omitted when its owning calculation did
not produce it. See the [Progress section of the `mize()`
reference](../reference/mize.html#progress) for the complete dynamic schema.

For live console output, use `verbose = TRUE`; `log_every` controls how often
progress is printed or stored. Keeping every iteration can produce a large
result on a long run.

> **Returned points and hard budgets.** `mize()` normally returns the best
> evaluated point. For nonmonotone runs, `best_*` and `last_*` distinguish that
> point from the final iterate. Under an exact callback budget, some objective
> fields may be unavailable. The [Convergence](convergence.html) article
> explains these cases, the status classes, and the stopping controls in
> detail.

## Where to go next

The default L-BFGS method is a useful baseline for a smooth problem. Problem
size and available derivative information may suggest another method:

| Situation | Method to consider | Main trade-off |
|---|---|---|
| General smooth problem | `L-BFGS` | Limited-memory quasi-Newton; the default |
| Smaller smooth problem | `BFGS` | Dense inverse-Hessian approximation uses quadratic storage |
| Very tight memory budget | `CG` | Low storage; sensitive to scaling and line-search choices |
| Extra gradient work is acceptable | `TN` | Approximates a Newton step without forming a Hessian |
| Hessian or inverse Hessian is available | `NEWTON` | Requires an `hs` or `hi` callback in the function list |

Treat these as starting points and benchmark promising methods on the problem
at hand. Continue with the guide that matches your next task:

- [Convergence](convergence.html) explains tolerances, limits, failures, and
  result interpretation.
- [Stateful optimization](stateful.html) shows how to step, monitor,
  checkpoint, and resume an optimizer.
- [Metric MDS](mmds.html) is a complete case study using closures, a checked
  gradient, and combined callbacks.
- [Choosing methods and tuning](methods.html) covers algorithm selection and
  advanced controls, including line searches, step sizes, and momentum.
