---
title: "Mize"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mize}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

`mize` is a package for doing unconstrained numerical optimization. This 
vignette describes the use of the `mize` function, which takes as input
a function to optimize, its gradient and a starting point and returns the 
optimized parameters.

First, I'll describe the required structure of the function and gradient
to optimize, then the various convergence options that apply no matter what
optimization methods are being used.

In the last section, I'll describe the methods which are available and any
options that apply to them specifically.

## The `fg` list

If you look at `stats::optim`, you'll see you pass the function to be optimized
and its gradient separately. By contrast, `mize` asks you to bundle them up
in a list, which I lazily refer to as an `fg` list. Here, for example, is the 
venerable Rosenbrock function:

```{r Defining a function and gradient to optimize}
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
```

For minimizing a simple function, there's not a lot of advantage to this. But
for more complex optimizations, there are some advantages:

* If `fn` and `gr` require extra parameterization, you can store the extra
state in a function that creates and returns the `fg` list, and then `fn`
and `gr` can access that state via the created closure. `stats::optim` takes
an alternative approach where extra data is passed as `...` to the `optim` 
function and then internally this is passed to `fn` and `gr` when needed.
* For some optimization methods (e.g. those using a strong Wolfe line search),
both `fn` and `gr` need to be evaluated for the same value of `par`. If there's
a lot of shared work between `fn` and `gr`, this can be quite inefficient.
If you provide an optional routine called `fg`, which calculates the function
and gradient simultaneously, this will be used by some parts of `mize` in 
preference to calling `fn` and `gr` separately, which can save a bit of time.

A version of `rb_fg` with an optional `fg` function could look like:

```{r A function list with an optional fg item}
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) },
   fg = function(x) {   
     a <- x[2] - x[1] * x[1]
     b <- 1 - x[1]
     list( 
       fn = 100 * a ^ 2 + b ^ 2,
       gr = c( -400 * x[1] * a - 2 * b,
                200 * a)
     )
   }
)
```

In the case of the Rosenbrock function, I'm not suggesting you should actually
do this. See the vignette on [Metric Multi-Dimensional Scaling](mmds.html) for 
the sort of scenario where it could make a difference.

For the rest of this vignette, we'll just assume you're only providing `fn`
and `gr`.

## The parameters to optimize

The parameters to optimize are passed as a numeric vector, `par`.

As well as using the `rb_fg` list we defined above to describe the Rosenbrock
function, we'll also start from the following starting point:

```{r Defining a starting point}
rb0 <- c(-1.2, 1)
```

## Basic options

If you don't particularly care about how your function gets optimized, the
default setting is to use the limited-memory Broyden-Fletcher-Goldfarb-Shanno
(L-BFGS) optimizer, which is a good choice for pretty much anything.

You may still need to tweak some options that are independent of the method,
like the maximum number of iterations to use, or other convergence criteria.

### Defaults

By default, `mize` will grind away on the function you give it for up to 100
iterations. But it will give up early if it senses that no further progress
is being made.

The return value provides both the optimized parameter as well as some 
information about how much work was done to get there.

```{r Defaults}
res <- mize(rb0, rb_fg)
# What were the final parameter values? (should be close to c(1, 1))
res$par

# What was the function value at that point (should be close to 0)
res$f

# How many iterations did it take?
res$iter

# How many function evaluations?
res$nf

# How many gradient evaluations?
res$ng

# Why did the optimization terminate?
res$terminate
```

As you can see from looking at `res$iter`, the L-BFGS didn't use all 100 
iterations it had available. The difference between consecutive function values,
`res$f` got sufficiently low, that the absolute tolerance criterion was passed: 
`res$terminate$what` will contain a string code that communicates what happened,
while `res$terminate$val` provides the specific value that causes the 
termination, although it's rarely interesting.

There is an entire vignette on the 
[different convergence criteria](convergence.html) available. The defaults 
should do a reasonable, albeit fairly conservative, job. 

### Logging progress to console

If you want to keep track of the progress of the optimization, set the `verbose`
option to `TRUE`:

```{r Verbose mode}
res <- mize(rb0, rb_fg, grad_tol = 1e-3, ginf_tol = 1e-3, max_iter = 10, 
            verbose = TRUE)
```

As you can see, information about the state of the optimization is logged to
the console every time the convergence criteria are checked. The values logged 
are:

* `f` - The current value of the function.
* `g2` - The current value of the gradient 2-norm (if you set a non-`NULL` 
`grad_tol`).
* `ginf` - The current value of the gradient infinity norm (if you set a 
non-`NULL` `ginf_tol`).
* `nf` - The total number of function evaluations so far.
* `ng` - The total number of gradient evaluations so far.
* `step` - The magnitude of the change between the previous version of `par` and
the current value.

Note that values are also provided for iteration 0, which represents the state
of `par` at its starting point. Because this requires a function evaluation that
might otherwise not be needed, you may get results where `nf` and `ng` are one
larger than you'd otherwise expect if `verbose` was `FALSE`.

Logging every iteration might be more output than you want. You can also set the 
`log_every` parameter to make logging slightly less noisy:

```{r Log every 10 iterations}
res <- mize(rb0, rb_fg, grad_tol = 1e-3, verbose = TRUE, log_every = 10)
```

On the assumption that you probably want to see information about the last
iteration, this is also reported, even if it wouldn't normally be reported
according to the setting for `log_every`.

### Storing progress

You may wish to do more than stare at the progress numbers logged to the console
during optimization when `verbose = TRUE`. If you would like access to these
values after optimization, set the `store_progress` parameter to true, and
a data frame called `progress` will be part of the return value:

```{r Returning stored progress}
res <- mize(rb0, rb_fg, store_progress = TRUE, log_every = 10)
res$progress
```

As can be seen, `store_progress` will also obey the `log_every` parameter and
not store every iteration of convergence information if requested.
