---
title: "Metric MDS"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric MDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

Metric multidimensional scaling (metric MDS) finds a low-dimensional
configuration whose pairwise distances resemble supplied dissimilarities. It
is also a useful example of an optimization problem where a closure keeps fixed
data out of the parameter vector.

We will reconstruct a two-dimensional configuration for the 21 European cities
in `datasets::eurodist`. The input contains road distances in kilometres, so a
two-dimensional Euclidean configuration can only approximate them.

## Stress and its gradient

Let $R = (r_{ij})$ be the target distances and let $D = (d_{ij})$ be the
Euclidean distances among candidate coordinates $Y$. We minimize raw stress:

$$
C(Y) = \sum_{i < j} (r_{ij} - d_{ij})^2.
$$

The condition $i < j$ counts each unordered pair once. The matrix
implementation below instead sums both symmetric triangles, so it multiplies
that sum by one half. With $n$ objects, the corresponding distance RMSE is
$\sqrt{C / {n \choose 2}}$ and has the same units as the input distances.

For distinct points, the derivative with respect to the coordinates of object
$i$ is

$$
\frac{\partial C}{\partial \mathbf{y}_i} =
-2 \sum_{j \ne i} \frac{r_{ij} - d_{ij}}{d_{ij}}
(\mathbf{y}_i - \mathbf{y}_j).
$$

At coincident points, $d_{ij} = 0$ and Euclidean distance is not
differentiable. The implementation uses a named `distance_epsilon` in the
denominator to keep the calculation finite. This numerical convention
slightly perturbs the displayed derivative. The derivative remains undefined
at a coincident configuration, so we validate the implementation at a seeded,
nondegenerate starting point.

The callback factory below captures the fixed target matrix in a closure. Its
`prepare()` helper performs the shared conversion from a parameter vector to
coordinates and pairwise distances. The counter will let us compare shared
work without relying on a noisy timing measurement.

```{r callback-factory}
distance_epsilon <- 1e-10

make_mmds_callbacks <- function(
  distances,
  combined = FALSE,
  epsilon = distance_epsilon
) {
  target <- as.matrix(distances)
  preparation_count <- 0L

  prepare <- function(par) {
    preparation_count <<- preparation_count + 1L
    coordinates <- matrix(par, ncol = 2, byrow = TRUE)
    list(
      coordinates = coordinates,
      distances = as.matrix(stats::dist(coordinates))
    )
  }

  stress <- function(prepared) {
    0.5 * sum((target - prepared$distances)^2)
  }

  gradient <- function(prepared) {
    weights <- (target - prepared$distances) /
      (prepared$distances + epsilon)
    coordinates <- prepared$coordinates
    result <- matrix(
      nrow = nrow(coordinates),
      ncol = ncol(coordinates)
    )

    for (i in seq_len(nrow(coordinates))) {
      differences <- sweep(-coordinates, 2, -coordinates[i, ])
      result[i, ] <- colSums(differences * weights[, i])
    }

    as.vector(t(result)) * -2
  }

  callbacks <- list(
    fn = function(par) stress(prepare(par)),
    gr = function(par) gradient(prepare(par))
  )
  if (combined) {
    callbacks$fg <- function(par) {
      prepared <- prepare(par)
      list(
        fn = stress(prepared),
        gr = gradient(prepared)
      )
    }
  }

  list(
    fg = callbacks,
    preparations = function() preparation_count,
    reset_preparations = function() {
      preparation_count <<- 0L
      invisible(NULL)
    }
  )
}
```

The separate variant supplies `fn` and `gr`. The combined variant also
supplies `fg`, which lets `mize` request an objective and gradient from one
prepared configuration.

```{r seeded-start}
target_distances <- datasets::eurodist
set.seed(42)
initial_par <- rnorm(attr(target_distances, "Size") * 2)

separate_callbacks <- make_mmds_callbacks(target_distances)
combined_callbacks <- make_mmds_callbacks(
  target_distances,
  combined = TRUE
)
```

## Validate the gradient

`check_mize_gradient()` compares the analytic gradient with central finite
differences. Because the combined callback is present, it also checks that
`fg` agrees with the separate `fn` and `gr` callbacks.

```{r gradient-check}
gradient_check <- check_mize_gradient(
  combined_callbacks$fg,
  initial_par
)

gradient_summary <- data.frame(
  max_abs_error = gradient_check$max_abs_error,
  max_rel_error = gradient_check$max_rel_error,
  fg_fn_abs_error = gradient_check$fg_consistency$fn$abs_error,
  fg_gr_max_abs_error =
    gradient_check$fg_consistency$gr$max_abs_error
)
signif(gradient_summary, digits = 4)
```

```{r assert-gradient-check, include = FALSE}
stopifnot(
  gradient_check$max_rel_error < 1e-3,
  isTRUE(gradient_check$fg_consistency$checked),
  gradient_check$fg_consistency$fn$abs_error < 1e-10,
  gradient_check$fg_consistency$gr$max_abs_error < 1e-10
)
```

The relative error is the more useful scale-free summary here. The exact zero
consistency errors show that the separate and combined paths implement the
same objective and gradient at the checked point.

## Optimize with separate and combined callbacks

This case study uses one explicit convergence criterion. Relative change in
stress is natural for this scale-dependent objective, so unrelated tolerances
are disabled. See the [Convergence](convergence.html) article when choosing a
criterion for another problem.

```{r optimization}
mmds_controls <- list(
  method = "L-BFGS",
  max_iter = 200,
  abs_tol = NULL,
  rel_tol = 1e-8,
  grad_tol = NULL,
  step_tol = NULL
)

separate_callbacks$reset_preparations()
combined_callbacks$reset_preparations()

separate_result <- do.call(
  mize,
  c(
    list(par = initial_par, fg = separate_callbacks$fg),
    mmds_controls
  )
)
combined_result <- do.call(
  mize,
  c(
    list(par = initial_par, fg = combined_callbacks$fg),
    mmds_controls
  )
)

results <- list(
  separate = separate_result,
  combined = combined_result
)
initial_stress <- gradient_check$fn
pair_count <- choose(attr(target_distances, "Size"), 2)

termination_table <- data.frame(
  callbacks = names(results),
  status = vapply(results, `[[`, character(1), "status"),
  termination = vapply(
    results,
    function(result) result$terminate$what,
    character(1)
  ),
  iterations = vapply(results, `[[`, numeric(1), "iter"),
  nf = vapply(results, `[[`, numeric(1), "nf"),
  ng = vapply(results, `[[`, numeric(1), "ng")
)
termination_table
```

```{r fit-table}
fit_table <- data.frame(
  callbacks = names(results),
  initial_stress = initial_stress,
  final_stress = vapply(results, `[[`, numeric(1), "f"),
  initial_rmse_km = sqrt(initial_stress / pair_count),
  final_rmse_km = sqrt(
    vapply(results, `[[`, numeric(1), "f") / pair_count
  )
)
fit_display <- fit_table
fit_display[-1] <- lapply(fit_display[-1], signif, digits = 7)
fit_display
```

Here, “final stress” is stress at the returned parameters; `mize()` returns
the best observed parameters. Both variants report status `converged` and
terminate on `rel_tol`; they reduce the distance RMSE from about
`r round(fit_table$initial_rmse_km[1])` km to
`r round(fit_table$final_rmse_km[1])` km. The tables show the observed
iteration and callback counts without turning an exact count into a lasting
promise.

```{r assert-optimization, include = FALSE}
stopifnot(
  all(vapply(results, `[[`, character(1), "status") == "converged"),
  all(vapply(results, `[[`, logical(1), "converged")),
  all(vapply(
    results,
    function(result) result$terminate$what,
    character(1)
  ) == "rel_tol"),
  all(is.finite(fit_table$final_stress)),
  all(fit_table$final_stress < fit_table$initial_stress),
  isTRUE(all.equal(
    separate_result$par,
    combined_result$par,
    tolerance = 1e-10
  )),
  isTRUE(all.equal(
    separate_result$f,
    combined_result$f,
    tolerance = 1e-10
  )),
  identical(separate_result$nf, combined_result$nf),
  identical(separate_result$ng, combined_result$ng)
)
```

The logical function and gradient counts are identical because both runs ask
for the same information. The underlying preparation counts expose the work
that `fg` shares:

```{r preparation-counts}
preparation_counts <- c(
  separate = separate_callbacks$preparations(),
  combined = combined_callbacks$preparations()
)
data.frame(
  callbacks = names(preparation_counts),
  shared_preparations = unname(preparation_counts)
)
```

```{r assert-preparation-counts, include = FALSE}
stopifnot(
  identical(separate_result$nf, combined_result$nf),
  identical(separate_result$ng, combined_result$ng),
  unname(preparation_counts["combined"]) <
    unname(preparation_counts["separate"])
)
```

This deterministic count shows that a combined callback avoids repeated work
when the objective and gradient share an expensive intermediate. The tiny
example provides no useful wall-clock estimate.

## Interpret the configuration

Distances leave absolute position and orientation unidentified. Adding one
vector to every point leaves all distances unchanged, as do rotations and
reflections. Centering removes the arbitrary translation for display. The plot
shows distance geometry; its axes and handedness carry no compass meaning.

```{r plot-results, fig.width=5.5, fig.height=5.5}
center_coordinates <- function(par) {
  coordinates <- matrix(par, ncol = 2, byrow = TRUE)
  sweep(coordinates, 2, colMeans(coordinates))
}

plot_mmds <- function(coordinates, distances, ...) {
  horizontal_cutoff <- 0.08 * diff(range(coordinates[, 1]))
  label_positions <- ifelse(
    coordinates[, 1] < -horizontal_cutoff,
    2,
    ifelse(
      coordinates[, 1] > horizontal_cutoff,
      4,
      rep(c(1, 3), length.out = nrow(coordinates))
    )
  )
  label_positions[which.min(coordinates[, 2])] <- 3
  label_positions[which.max(coordinates[, 2])] <- 1
  city_names <- labels(distances)
  label_positions[city_names == "Geneva"] <- 2
  label_positions[city_names == "Lyons"] <- 4
  graphics::plot(
    coordinates,
    type = "n",
    asp = 1,
    xlab = "Centered dimension 1",
    ylab = "Centered dimension 2",
    main = "Metric MDS configuration of European road distances"
  )
  graphics::points(coordinates, pch = 19, cex = 0.45)
  graphics::text(
    coordinates[, 1],
    coordinates[, 2],
    labels = city_names,
    pos = label_positions,
    offset = 0.25,
    ...
  )
}

centered_coordinates <- center_coordinates(combined_result$par)
plot_mmds(centered_coordinates, target_distances, cex = 0.55)
```

The invariance holds at the objective level. The hidden check below translates,
rotates, and reflects the returned configuration and verifies that stress is
unchanged to numerical tolerance.

```{r assert-stress-invariance, include = FALSE}
angle <- pi / 3
rotation <- matrix(
  c(cos(angle), -sin(angle), sin(angle), cos(angle)),
  nrow = 2,
  byrow = TRUE
)
translated <- sweep(
  centered_coordinates,
  2,
  c(1000, -500),
  "+"
)
rotated <- centered_coordinates %*% rotation
reflected <- centered_coordinates
reflected[, 1] <- -reflected[, 1]

stress_at <- function(coordinates) {
  combined_callbacks$fg$fn(as.vector(t(coordinates)))
}
invariant_stress <- vapply(
  list(translated, rotated, reflected),
  stress_at,
  numeric(1)
)
relative_invariance_error <- max(
  abs(invariant_stress - combined_result$f)
) / max(1, abs(combined_result$f))
stopifnot(relative_invariance_error < 1e-12)
```

The same closure pattern applies whenever callbacks need fixed data or share
an intermediate calculation. For the full callback contract, see `?mize`.

## See also

- [Getting started](mize.html) introduces callback lists and gradient checks.
- [Convergence](convergence.html) explains the relative tolerance used here.
- [Stateful optimization](stateful.html) shows step-by-step integration with
  retained optimizer state.
