---
title: "Metric MDS"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric MDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

Minimizing the Rosenbrock function is all very well, but there's often a more
complex relationship between the parameters we're looking to optimize and what
would be convenient to pass to a cost function. In many cases there are other
quantities that we need to define and store somewhere for the cost function
to provide a value, but which aren't under our direct control and hence are
not suitable for optimization. They just need to be available for evaluation
inside `fn` and `gr`.

This vignette is less to do with the running of `mize` itself, and more to do
with using it to solve an actual problem. It contains plenty of R code, but
not a lot of it is `mize`-specific. Hopefully it will demonstrate how it
can be used for non-trivial problems.

### Metric Multi-Dimensional Scaling

Metric multi-dimensional scaling (Metric MDS) is a way to provide a 
low-dimensional (usually two-dimensional) view of a high dimensional data set,
although all it needs is a distance matrix to work off.

Let's take the `eurodist` dataset as an example, which can be found in the
`datasets` package. The description says:

> The `eurodist` gives the road distances (in km) between 21 cities in Europe.

We should be able to reconstruct the relative locations of the 21 cities on
a 2D plot with this information, subject to some error because roads aren't
straight between the cities and the cities themselves lie on the non-flat
surface of the Earth.

### The Metric MDS Cost Function 

The parameters we're going to optimize are the 2D locations of each city. The
cost function, however, is more naturally expressed as being related to the 
distance matrices. What we want is for the distance matrix of the output, which
we'll call $D$, to resemble the input distance matrix, $R$, as much as possible.

One obvious way to express this as a cost function is to consider the square 
loss:

$$C = \sum_{i<j} \left(r_{ij} - d_{ij}\right)^2$$
where $r_{ij}$ is the distance between city $i$ and city $j$ in the input 
distance matrix, and $d_{ij}$ represents the distance in the output 
configuration. The $i<j$ bit indicates that we only want to count each $i,j$ 
pair twice, because distances are symmetric. Normally, you'd probably want
to divide this result by the number of pairs and then take the square root
to get the units of the error as a distance and not dependent on the size
of the data set, but we won't worry about that.

As R code, the cost function can be written as:

```{r cost function}
# R and D are the input and output distance matrices, respectively
cost_fun <- function(R, D) {
  diff2 <- (R - D) ^ 2
  sum(diff2) * 0.5
}
```

It's not good R style to use single character upper-case variables, but for 
matrices, we'll get away with it just this once. I've also taken advantage
of the vectorization of R code for matrices, rather than explicitly loop. As
the calculation takes place over the entire matrix, I've halved the result
to avoid the double counting of each distance.

Both $R$ and $D$ in the R code are of class `matrix` rather than the `dist`
type that `eurodist` is. The `matrix` class is more convenient for the
computations that are needed, but not as efficient in terms of storage. The
`eurodist` data can be converted using `as.matrix`:

```{r convert dist to matrix}
eurodist_mat <- as.matrix(eurodist)
```

### The Metric MDS Gradient

Writing out the cost function with respect to distances only makes it pretty
straightforward. However, we need the gradient with respect to the parameters
we are optimizing, which is the coordinates of each city. Indicating the 
(two-dimensional) vector that represents the coordintes of city $i$ as 
$\mathbf{y_i}$, the gradient of the cost function above is:

$$\frac{\partial C}{\partial \mathbf{y_i}} = 
  -2\sum_j \frac{r_{ij} - d_{ij}}{d_{ij}}
  \left(
   \mathbf{y_i - y_j}
  \right)
$$
where the sum $j$ is over all the other cities.

As R code, this is:

```{r MMDS gradient}
# R is the input distance matrix
# D is the output distance matrix
# y is a n x d matrix of output coordinates
cost_grad <- function(R, D, y) {
  K <- (R - D) / (D + 1.e-10)

  G <- matrix(nrow = nrow(y), ncol = ncol(y))
  
  for (i in 1:nrow(y)) {
    dyij <- sweep(-y, 2, -y[i, ])
    G[i, ] <- apply(dyij * K[, i], 2, sum)
  }

  as.vector(t(G)) * -2
}
```

That loop in the middle is a bit cryptic with its `sweep`s and `apply`s, but it
does the job of calculating the $\sum_j \mathbf{y_i} - \mathbf{y_j}$ part of the
gradient for each row of the gradient matrix.

### `mize` function and gradient

We can now write the function and gradient routines that have the interface that 
`mize` is expecting:

```{r mize function}
mmds_fn <- function(par) {
  R <- as.matrix(eurodist)
  y <- matrix(par, ncol = 2, byrow = TRUE)
  D <- as.matrix(stats::dist(y))
  
  cost_fun(R, D)
}
```

```{r mize gradient}
mmds_gr <- function(par) {
  R <- as.matrix(eurodist)
  y <- matrix(par, ncol = 2, byrow = TRUE)
  D <- as.matrix(stats::dist(y))

  cost_grad(R, D, y)
}
```

These routines are not a model of efficiency. Not a problem for this small 
dataset, but we'll have to revisit this below.

For now, let's choose a starting point and get optimizing. We may as well begin 
from a random location:

```{r optimization}
set.seed(42)
ed0 <- rnorm(attr(eurodist, 'Size') * 2)

res_euro <- mize(ed0, list(fn = mmds_fn, gr = mmds_gr), 
                method = "L-BFGS", verbose = TRUE, 
                grad_tol = 1e-5, check_conv_every = 10)
```

It takes 90 iterations to converge, but we get there. Now for the moment of 
truth: are the cities laid out in a way we'd expect? Here's a function that 
will plot the `par` results:

```{r plot results, fig.width=5.5, fig.height=5.5}
plot_mmds <- function(coords, dist, ...) {
  if (class(coords) == "numeric") {
    coords <- matrix(coords, ncol = 2, byrow = TRUE)
  }
  graphics::plot(coords, type = 'n')
  graphics::text(coords[, 1], coords[, 2], labels = labels(dist), ...)
}
plot_mmds(res_euro$par, eurodist, cex = 0.5)
```

Ah yes, Athens up in the top-right of a map of Europe, just as we expect. Ahem.
In fact, because we are only optimizing distances, not absolute positions, there's 
no reason that, starting from a random configuration, north will be "up" in the 
optimized results. Just to prove this works, let's rotate the coordinates by 90 
degrees clockwise:

```{r rotated plot, fig.width=5.5, fig.height=5.5}
rot90 <- matrix(c(0, -1, 1, 0), ncol = 2)
rotated <- t(rot90 %*% t(matrix(res_euro$par, ncol = 2, byrow = TRUE)))
plot_mmds(rotated, eurodist, cex = 0.5)
```

That's better.

### Improving the Efficiency of the Function and Gradient Routines

So the `mize`-powered MMDS routine is working. But it's un-necessarily inefficient. 
Every time the function or gradient is calculated, we convert `eurodist` into a
matrix. This only needs to be done once, as long as the resulting matrix is in
scope inside the function. Also, if the function and gradient is calculated for
the same value of `par`, `par` is converted to a distance matrix twice.

The first issue can be fixed without polluting global scope with the converted
`eurodist` matrix by making the function and gradient closures with access to the
input distance matrix. The second problem can be alleviated by adding a third item
in the list passed to `mize`. Add a function called `fg`. This should calculate the 
cost function and gradient in one call, returning the values in a list. Routines that 
need to calculate the function and gradient for the same value of `par` will look for 
`fg` and call that, in preference to calling `fn` and `gr` separately.

Putting it all together, a function to create a suitable `fg` list to pass
to `mize` is:

```{r more efficient fg}
make_fg <- function(distmat) {
  R <- as.matrix(distmat)
  cost_fun <- function(R, D) {
    diff2 <- (R - D) ^ 2
    sum(diff2) * 0.5
  }
  cost_grad <- function(R, D, y) {
    K <- (R - D) / (D + 1.e-10)

    G <- matrix(nrow = nrow(y), ncol = ncol(y))

    for (i in 1:nrow(y)) {
      dyij <- sweep(-y, 2, -y[i, ])
      G[i, ] <- apply(dyij * K[, i], 2, sum)
    }

    as.vector(t(G)) * -2
  }
  list(
    fn = function(par) {
      y <- matrix(par, ncol = 2, byrow = TRUE)
      D <- as.matrix(stats::dist(y))
      cost_fun(R, D)
    },
    gr = function(par) {
      y <- matrix(par, ncol = 2, byrow = TRUE)
      D <- as.matrix(stats::dist(y))
      cost_grad(R, D, y)
    },
    fg = function(par) {
      y <- matrix(par, ncol = 2, byrow = TRUE)
      D <- as.matrix(stats::dist(y))
      list(
        fn = cost_fun(R, D),
        gr = cost_grad(R, D, y)
      )
    }
  )
}
```

We can then run the MMDS optimization as before:
```{r optimization with improved fg}
res_euro <- mize(ed0, make_fg(eurodist), 
                method = "L-BFGS", verbose = TRUE, 
                grad_tol = 1e-5, check_conv_every = 10)
```
The same results are achieved, but it runs a tiny bit faster, although
unless you are using a very slow computer you will have to take my word for it.
However, as the work done to create the data that will be used for the gradient
and function calculation grows larger, you should consider making use of the \code{fg}
item.

