---
title: "Metric MDS"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric MDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

Can an optimizer recover a recognizable map of Europe from a table of road
distances? We will use that problem to explore three parts of the `mize`
callback interface: capturing fixed data in a closure, checking an analytic
gradient, and sharing intermediate calculations between the objective and
gradient.

`datasets::eurodist` contains road distances in kilometres among 21 European
cities. We will fit a two-dimensional configuration whose Euclidean distances
approximate those values. Roads need not follow straight lines, and the cities
lie on the curved surface of the Earth, so some disagreement is inevitable.

This example minimizes raw stress directly. `stats::cmdscale()` follows the
classical-scaling route and uses a different criterion.

## Stress and its gradient

Let $R = (r_{ij})$ be the target distances and let $D = (d_{ij})$ be the
Euclidean distances among candidate coordinates $Y$. We minimize raw stress:

$$
C(Y) = \sum_{i < j} (r_{ij} - d_{ij})^2.
$$

The condition $i < j$ counts each unordered pair once. The matrix
implementation below instead sums both symmetric triangles, so it multiplies
that sum by one half. With $n$ objects, the corresponding distance RMSE is
$\sqrt{C / {n \choose 2}}$ and has the same units as the input distances.

For distinct points, the derivative with respect to the coordinates of object
$i$ is

$$
\frac{\partial C}{\partial \mathbf{y}_i} =
-2 \sum_{j \ne i} \frac{r_{ij} - d_{ij}}{d_{ij}}
(\mathbf{y}_i - \mathbf{y}_j).
$$

> **Implementation note.** Euclidean distance is nondifferentiable when two
> candidate points coincide. `distance_epsilon` prevents division by zero in
> the code, although the mathematical derivative remains undefined at an
> exactly coincident configuration. The gradient check therefore uses a
> seeded, nondegenerate point.

The callback factory captures the fixed target matrix in a closure. Its
`prepare()` helper performs the shared conversion from a parameter vector to
coordinates and pairwise distances. A counter records how often that work is
performed.

```{r callback-factory}
distance_epsilon <- 1e-10

make_mmds_callbacks <- function(
  distances,
  combined = FALSE,
  epsilon = distance_epsilon
) {
  target <- as.matrix(distances)
  preparation_count <- 0L

  prepare <- function(par) {
    preparation_count <<- preparation_count + 1L
    coordinates <- matrix(par, ncol = 2, byrow = TRUE)
    list(
      coordinates = coordinates,
      distances = as.matrix(stats::dist(coordinates))
    )
  }

  stress <- function(prepared) {
    0.5 * sum((target - prepared$distances)^2)
  }

  gradient <- function(prepared) {
    weights <- (target - prepared$distances) /
      (prepared$distances + epsilon)
    coordinates <- prepared$coordinates
    result <- matrix(
      nrow = nrow(coordinates),
      ncol = ncol(coordinates)
    )

    for (i in seq_len(nrow(coordinates))) {
      differences <- sweep(-coordinates, 2, -coordinates[i, ])
      result[i, ] <- colSums(differences * weights[, i])
    }

    as.vector(t(result)) * -2
  }

  callbacks <- list(
    fn = function(par) stress(prepare(par)),
    gr = function(par) gradient(prepare(par))
  )
  if (combined) {
    callbacks$fg <- function(par) {
      prepared <- prepare(par)
      list(
        fn = stress(prepared),
        gr = gradient(prepared)
      )
    }
  }

  list(
    fg = callbacks,
    preparations = function() preparation_count,
    reset_preparations = function() {
      preparation_count <<- 0L
      invisible(NULL)
    }
  )
}
```

The separate variant supplies `fn` and `gr`. The combined variant also
supplies `fg`, which lets `mize` request an objective and gradient from one
prepared configuration.

```{r seeded-start}
target_distances <- datasets::eurodist
set.seed(42)
initial_par <- rnorm(attr(target_distances, "Size") * 2)

separate_callbacks <- make_mmds_callbacks(target_distances)
combined_callbacks <- make_mmds_callbacks(
  target_distances,
  combined = TRUE
)
```

## Validate the gradient

`check_mize_gradient()` compares the analytic gradient with central finite
differences. Because the combined callback is present, it also checks that
`fg` agrees with the separate `fn` and `gr` callbacks.

```{r gradient-check}
gradient_check <- check_mize_gradient(
  combined_callbacks$fg,
  initial_par
)

gradient_summary <- data.frame(
  max_abs_error = gradient_check$max_abs_error,
  max_rel_error = gradient_check$max_rel_error,
  fg_fn_abs_error = gradient_check$fg_consistency$fn$abs_error,
  fg_gr_max_abs_error =
    gradient_check$fg_consistency$gr$max_abs_error
)
knitr::kable(
  signif(gradient_summary, digits = 4),
  row.names = FALSE,
  col.names = c(
    "Maximum absolute error",
    "Maximum relative error",
    "fg/fn absolute error",
    "fg/gr maximum absolute error"
  )
)
```

```{r assert-gradient-check, include = FALSE}
stopifnot(
  gradient_check$max_rel_error < 1e-3,
  isTRUE(gradient_check$fg_consistency$checked),
  gradient_check$fg_consistency$fn$abs_error < 1e-10,
  gradient_check$fg_consistency$gr$max_abs_error < 1e-10
)
```

The analytic gradient agrees with the finite-difference approximation to about
`r format(gradient_check$max_rel_error, digits = 2, scientific = TRUE)` in
relative terms. At the checked point, the combined callback also reproduces
the separate objective and gradient to the displayed precision.

## Optimize with separate and combined callbacks

For this example, we stop when the relative change in stress falls below
$10^{-8}$. Stress depends on the scale of the supplied distances, which makes a
relative criterion convenient here. The [Convergence](convergence.html)
article discusses the other controls.

```{r optimization}
mmds_controls <- list(
  method = "L-BFGS",
  max_iter = 200,
  abs_tol = NULL,
  rel_tol = 1e-8,
  grad_tol = NULL,
  step_tol = NULL
)

separate_callbacks$reset_preparations()
combined_callbacks$reset_preparations()

separate_result <- do.call(
  mize,
  c(
    list(par = initial_par, fg = separate_callbacks$fg),
    mmds_controls
  )
)
combined_result <- do.call(
  mize,
  c(
    list(par = initial_par, fg = combined_callbacks$fg),
    mmds_controls
  )
)

results <- list(
  separate = separate_result,
  combined = combined_result
)
initial_stress <- gradient_check$fn
pair_count <- choose(attr(target_distances, "Size"), 2)
preparation_counts <- c(
  separate = separate_callbacks$preparations(),
  combined = combined_callbacks$preparations()
)

initial_rmse_km <- sqrt(initial_stress / pair_count)
final_rmse_km <- sqrt(combined_result$f / pair_count)

callback_comparison <- data.frame(
  callbacks = c("separate fn/gr", "combined fg"),
  iterations = vapply(results, `[[`, numeric(1), "iter"),
  objective_calls = vapply(results, `[[`, numeric(1), "nf"),
  gradient_calls = vapply(results, `[[`, numeric(1), "ng"),
  distance_preparations = unname(preparation_counts),
  check.names = FALSE
)
knitr::kable(
  callback_comparison,
  row.names = FALSE,
  col.names = c(
    "Callbacks",
    "Iterations",
    "Objective calls",
    "Gradient calls",
    "Distance preparations"
  )
)
```

Both callback forms converge on `rel_tol` from the fixed starting point and
return the same configuration. They take
`r combined_result$iter` iterations and reduce pairwise-distance RMSE from
`r format(round(initial_rmse_km), big.mark = ",")` km to
`r round(final_rmse_km)` km.

`mize` requests the same logical information in both runs. The combined
callback performs `r unname(preparation_counts["combined"])` distance-matrix
preparations instead of
`r unname(preparation_counts["separate"])` because a joint request can reuse
one prepared configuration. On this small problem, that deterministic count is
more informative than a wall-clock comparison. Exact callback totals may
change with line-search details.

Raw stress is nonconvex, so this seeded run demonstrates a reproducible fit
without claiming that it found the global minimum.

```{r assert-optimization, include = FALSE}
stopifnot(
  all(vapply(results, `[[`, character(1), "status") == "converged"),
  all(vapply(results, `[[`, logical(1), "converged")),
  all(vapply(
    results,
    function(result) result$terminate$what,
    character(1)
  ) == "rel_tol"),
  all(vapply(results, `[[`, numeric(1), "f") < initial_stress),
  isTRUE(all.equal(
    separate_result$par,
    combined_result$par,
    tolerance = 1e-10
  )),
  isTRUE(all.equal(
    separate_result$f,
    combined_result$f,
    tolerance = 1e-10
  )),
  identical(separate_result$nf, combined_result$nf),
  identical(separate_result$ng, combined_result$ng),
  unname(preparation_counts["combined"]) <
    unname(preparation_counts["separate"])
)
```

## Put the configuration on the map

The fitted configuration may initially look sideways or mirrored. Stress
depends only on pairwise distances, so translating, rotating, or reflecting
every point produces an equally good solution. Metric MDS recovers an
approximate distance geometry, while its origin, compass direction, and
handedness remain undetermined.

For a familiar frame of reference, we project approximate city-centre
locations onto a plane measured in kilometres, then find the translated,
rotated, or reflected copy of the fitted configuration that lies closest to
them. The alignment uses an orthogonal transformation and leaves the scale
fixed. Consequently, it preserves every fitted distance and the raw stress.

<details>
<summary>Show the projection and alignment code</summary>

```{r geographic-alignment}
city_names <- labels(target_distances)

# Approximate city-centre coordinates based on maps::world.cities (maps 3.4.3),
# with the historical eurodist names retained and Hook of Holland added.
city_locations <- data.frame(
  city = city_names,
  longitude = c(
    23.73, 2.17, 4.33, 1.86, -1.63, 6.97, 12.57,
    6.14, -5.35, 10.00, 4.13, -9.14, 4.83, -3.71,
    5.37, 9.19, 11.58, 2.34, 12.50, 18.07, 16.37
  ),
  latitude = c(
    37.98, 41.40, 50.83, 50.95, 49.64, 50.95, 55.68,
    46.21, 36.14, 53.55, 51.98, 38.72, 45.76, 40.42,
    43.31, 45.48, 48.14, 48.86, 41.89, 59.33, 48.22
  )
)

project_azimuthal_equidistant <- function(
  longitude,
  latitude,
  center_longitude,
  center_latitude,
  radius_km = 6371.0088
) {
  to_radians <- pi / 180
  lambda <- longitude * to_radians
  phi <- latitude * to_radians
  lambda0 <- center_longitude * to_radians
  phi0 <- center_latitude * to_radians
  delta_lambda <- lambda - lambda0

  cos_angle <-
    sin(phi0) * sin(phi) +
    cos(phi0) * cos(phi) * cos(delta_lambda)
  cos_angle <- pmax(-1, pmin(1, cos_angle))
  angle <- acos(cos_angle)

  projection_scale <- rep(1, length(angle))
  nonzero <- is.finite(angle) &
    abs(angle) > sqrt(.Machine$double.eps)
  projection_scale[nonzero] <-
    angle[nonzero] / sin(angle[nonzero])

  cbind(
    east_km =
      radius_km *
      projection_scale *
      cos(phi) *
      sin(delta_lambda),
    north_km =
      radius_km *
      projection_scale *
      (
        cos(phi0) * sin(phi) -
          sin(phi0) * cos(phi) * cos(delta_lambda)
      )
  )
}

rigid_align <- function(configuration, reference) {
  configuration_center <- colMeans(configuration)
  reference_center <- colMeans(reference)
  x <- sweep(configuration, 2, configuration_center)
  y <- sweep(reference, 2, reference_center)

  decomposition <- svd(crossprod(x, y))
  orthogonal <- decomposition$u %*% t(decomposition$v)

  list(
    coordinates = sweep(
      x %*% orthogonal,
      2,
      reference_center,
      "+"
    ),
    orthogonal = orthogonal
  )
}

projection_center <- c(
  mean(city_locations$longitude),
  mean(city_locations$latitude)
)
geographic_coordinates <- project_azimuthal_equidistant(
  city_locations$longitude,
  city_locations$latitude,
  center_longitude = projection_center[1],
  center_latitude = projection_center[2]
)

mmds_coordinates <- matrix(
  combined_result$par,
  ncol = 2,
  byrow = TRUE
)
alignment <- rigid_align(mmds_coordinates, geographic_coordinates)
aligned_coordinates <- alignment$coordinates

europe <- maps::map(
  "world",
  plot = FALSE,
  xlim = c(-12, 27),
  ylim = c(34, 62)
)
outside_europe <-
  europe$x < -12 | europe$x > 27 |
  europe$y < 34 | europe$y > 62
europe$x[outside_europe] <- NA_real_
europe$y[outside_europe] <- NA_real_
europe_coordinates <- project_azimuthal_equidistant(
  europe$x,
  europe$y,
  center_longitude = projection_center[1],
  center_latitude = projection_center[2]
)
```

</details>

```{r assert-rigid-alignment, include = FALSE}
stopifnot(
  identical(city_locations$city, city_names),
  max(abs(
    crossprod(alignment$orthogonal) - diag(2)
  )) < 1e-12,
  max(abs(
    as.vector(stats::dist(mmds_coordinates)) -
      as.vector(stats::dist(aligned_coordinates))
  )) < 1e-8
)
```

```{r aligned-map, echo=FALSE, fig.width=7.2, fig.height=6.2, fig.cap="Open circles mark projected geographic locations. Filled points show the metric-MDS configuration after translation, rotation, and possible reflection; grey segments join corresponding cities. No rescaling was applied.", fig.alt="Outline map of Europe with geographic city locations shown as open circles and rigidly aligned metric MDS positions shown as filled points, joined by grey line segments."}
plot_coordinates <- rbind(
  europe_coordinates,
  geographic_coordinates,
  aligned_coordinates
)
x_limits <- extendrange(
  plot_coordinates[, 1],
  f = 0.04
)
y_limits <- extendrange(
  plot_coordinates[, 2],
  f = 0.04
)

graphics::plot(
  europe_coordinates,
  type = "l",
  col = "grey82",
  lwd = 0.8,
  asp = 1,
  xlim = x_limits,
  ylim = y_limits,
  xlab = "Projected east-west position (km)",
  ylab = "Projected north-south position (km)"
)
graphics::segments(
  geographic_coordinates[, 1],
  geographic_coordinates[, 2],
  aligned_coordinates[, 1],
  aligned_coordinates[, 2],
  col = "grey65",
  lwd = 0.8
)
graphics::points(
  geographic_coordinates,
  pch = 1,
  cex = 0.8,
  lwd = 1
)
graphics::points(
  aligned_coordinates,
  pch = 19,
  cex = 0.55,
  col = "#1B6CA8"
)

horizontal_residual <-
  aligned_coordinates[, 1] - geographic_coordinates[, 1]
label_positions <- ifelse(
  abs(horizontal_residual) > 20,
  ifelse(horizontal_residual < 0, 2, 4),
  rep(c(2, 4), length.out = length(city_names))
)
label_positions[city_names %in% c("Athens", "Rome")] <- 4
label_positions[city_names %in% c("Lisbon", "Stockholm")] <- 2
label_positions[city_names == "Brussels"] <- 1
label_positions[city_names == "Calais"] <- 2
label_positions[city_names == "Cologne"] <- 2
label_positions[city_names == "Hook of Holland"] <- 4
label_positions[city_names == "Geneva"] <- 2
label_positions[city_names == "Milan"] <- 4
graphics::text(
  aligned_coordinates[, 1],
  aligned_coordinates[, 2],
  labels = city_names,
  pos = label_positions,
  offset = 0.25,
  cex = 0.62,
  col = "#143D59",
  xpd = NA
)
graphics::legend(
  "topright",
  inset = 0.02,
  legend = c("Geographic location", "Metric MDS"),
  pch = c(1, 19),
  col = c("black", "#1B6CA8"),
  pt.cex = c(0.8, 0.55),
  bty = "n",
  cex = 0.75
)
```

The `r round(final_rmse_km)` km pairwise-distance RMSE summarizes how well the
configuration fits the road-distance table. The segments in the map answer a
second question: how closely does that road-distance geometry resemble
projected geography? Their remaining lengths are expected because road routes,
coastlines, and the compromise required to fit all 210 city pairs affect the
target distances.

The same closure pattern applies whenever callbacks need fixed data or share
an intermediate calculation. For the full callback contract, see `?mize`.

## See also

- [Getting started](mize.html) introduces callback lists and gradient checks.
- [Convergence](convergence.html) explains the relative tolerance used here.
- [Stateful optimization](stateful.html) shows step-by-step integration with
  retained optimizer state.
