---
title: "Metric MDS"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric MDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

Can an optimizer recover a recognizable map of Europe from a table of road
distances? We will use that problem to explore three parts of the `mize`
callback interface: capturing fixed data in a closure, checking an analytic
gradient, and sharing intermediate calculations between the objective and
gradient.

`datasets::eurodist` contains road distances in kilometres among 21 European
cities. We will fit a two-dimensional configuration whose Euclidean distances
approximate those values. Roads need not follow straight lines, and the cities
lie on the curved surface of the Earth, so an exact two-dimensional Euclidean
fit should not be expected.

This example minimizes raw stress directly. `stats::cmdscale()` follows the
classical-scaling route and uses a different criterion.

## Stress and its gradient

Let $R = (r_{ij})$ be the target distances and let $D = (d_{ij})$ be the
Euclidean distances among candidate coordinates $Y$. We minimize raw stress:

$$
C(Y) = \sum_{i < j} (r_{ij} - d_{ij})^2.
$$

The condition $i < j$ counts each unordered pair once. The matrix
implementation below instead sums both symmetric triangles, so it multiplies
that sum by one half. With $n$ objects, the corresponding distance RMSE is
$\sqrt{C / {n \choose 2}}$ and has the same units as the input distances.

For distinct points, the derivative with respect to the coordinates of object
$i$ is

$$
\frac{\partial C}{\partial \mathbf{y}_i} =
-2 \sum_{j \ne i} \frac{r_{ij} - d_{ij}}{d_{ij}}
(\mathbf{y}_i - \mathbf{y}_j).
$$

> **Implementation note.** Euclidean distance is nondifferentiable when two
> candidate points coincide. `distance_epsilon` prevents division by zero in
> the code, although the mathematical derivative remains undefined at an
> exactly coincident configuration. The gradient check therefore uses a
> seeded, nondegenerate point.

The callback factory captures the fixed target matrix in a closure. Its
`prepare()` helper performs the shared conversion from a parameter vector to
coordinates and pairwise distances. A counter records how often that work is
performed.

```{r callback-factory}
distance_epsilon <- 1e-10

make_mmds_callbacks <- function(
  distances,
  combined = FALSE,
  epsilon = distance_epsilon
) {
  target <- as.matrix(distances)
  preparation_count <- 0L

  prepare <- function(par) {
    preparation_count <<- preparation_count + 1L
    coordinates <- matrix(par, ncol = 2, byrow = TRUE)
    list(
      coordinates = coordinates,
      distances = as.matrix(stats::dist(coordinates))
    )
  }

  stress <- function(prepared) {
    0.5 * sum((target - prepared$distances)^2)
  }

  gradient <- function(prepared) {
    weights <- (target - prepared$distances) /
      (prepared$distances + epsilon)
    coordinates <- prepared$coordinates
    result <- matrix(
      nrow = nrow(coordinates),
      ncol = ncol(coordinates)
    )

    for (i in seq_len(nrow(coordinates))) {
      differences <- sweep(-coordinates, 2, -coordinates[i, ])
      result[i, ] <- colSums(differences * weights[, i])
    }

    as.vector(t(result)) * -2
  }

  callbacks <- list(
    fn = function(par) stress(prepare(par)),
    gr = function(par) gradient(prepare(par))
  )
  if (combined) {
    callbacks$fg <- function(par) {
      prepared <- prepare(par)
      list(
        fn = stress(prepared),
        gr = gradient(prepared)
      )
    }
  }

  list(
    fg = callbacks,
    preparations = function() preparation_count,
    reset_preparations = function() {
      preparation_count <<- 0L
      invisible(NULL)
    }
  )
}
```

The separate variant supplies `fn` and `gr`. The combined variant also
supplies `fg`, which lets `mize` request an objective and gradient from one
prepared configuration.

```{r seeded-start}
target_distances <- datasets::eurodist
set.seed(42)
initial_par <- rnorm(attr(target_distances, "Size") * 2)

separate_callbacks <- make_mmds_callbacks(target_distances)
combined_callbacks <- make_mmds_callbacks(
  target_distances,
  combined = TRUE
)
```

## Validate the gradient

`check_mize_gradient()` compares the analytic gradient with central finite
differences. Because the combined callback is present, it also checks that
`fg` agrees with the separate `fn` and `gr` callbacks.

```{r gradient-check}
gradient_check <- check_mize_gradient(
  combined_callbacks$fg,
  initial_par
)

gradient_summary <- data.frame(
  max_abs_error = gradient_check$max_abs_error,
  max_rel_error = gradient_check$max_rel_error,
  fg_fn_abs_error = gradient_check$fg_consistency$fn$abs_error,
  fg_gr_max_abs_error =
    gradient_check$fg_consistency$gr$max_abs_error
)
knitr::kable(
  signif(gradient_summary, digits = 4),
  row.names = FALSE,
  col.names = c(
    "Maximum absolute error",
    "Maximum relative error",
    "fg/fn absolute error",
    "fg/gr maximum absolute error"
  )
)
```

```{r assert-gradient-check, include = FALSE}
stopifnot(
  gradient_check$max_rel_error < 1e-3,
  isTRUE(gradient_check$fg_consistency$checked),
  gradient_check$fg_consistency$fn$abs_error < 1e-10,
  gradient_check$fg_consistency$gr$max_abs_error < 1e-10
)
```

The analytic gradient agrees with the finite-difference approximation to about
`r format(gradient_check$max_rel_error, digits = 2, scientific = TRUE)` in
relative terms. At the checked point, the combined callback also reproduces
the separate objective and gradient to the displayed precision.

## Optimize with separate and combined callbacks

For this example, we stop when the relative change in stress falls below
$10^{-8}$. Stress depends on the scale of the supplied distances, which makes a
relative criterion convenient here. The [Convergence](convergence.html)
article discusses the other controls.

```{r optimization}
mmds_controls <- list(
  method = "L-BFGS",
  max_iter = 200,
  abs_tol = NULL,
  rel_tol = 1e-8,
  grad_tol = NULL,
  step_tol = NULL
)

separate_callbacks$reset_preparations()
combined_callbacks$reset_preparations()

separate_result <- do.call(
  mize,
  c(
    list(par = initial_par, fg = separate_callbacks$fg),
    mmds_controls
  )
)
combined_result <- do.call(
  mize,
  c(
    list(par = initial_par, fg = combined_callbacks$fg),
    mmds_controls
  )
)

results <- list(
  separate = separate_result,
  combined = combined_result
)
initial_stress <- gradient_check$fn
pair_count <- choose(attr(target_distances, "Size"), 2)
preparation_counts <- c(
  separate = separate_callbacks$preparations(),
  combined = combined_callbacks$preparations()
)

initial_rmse_km <- sqrt(initial_stress / pair_count)
final_rmse_km <- sqrt(combined_result$f / pair_count)

callback_comparison <- data.frame(
  callbacks = c("separate fn/gr", "combined fg"),
  iterations = vapply(results, `[[`, numeric(1), "iter"),
  objective_calls = vapply(results, `[[`, numeric(1), "nf"),
  gradient_calls = vapply(results, `[[`, numeric(1), "ng"),
  distance_preparations = unname(preparation_counts),
  check.names = FALSE
)
knitr::kable(
  callback_comparison,
  row.names = FALSE,
  col.names = c(
    "Callbacks",
    "Iterations",
    "Objective calls",
    "Gradient calls",
    "Distance preparations"
  )
)
```

Both callback forms converge on `rel_tol` from the fixed starting point and
return the same configuration. They take
`r combined_result$iter` iterations and reduce pairwise-distance RMSE from
`r format(round(initial_rmse_km), big.mark = ",")` km to
`r round(final_rmse_km)` km.

`mize` requests the same logical information in both runs. The combined
callback performs `r unname(preparation_counts["combined"])` distance-matrix
preparations instead of
`r unname(preparation_counts["separate"])` because a joint request can reuse
one prepared configuration. On this small problem, that deterministic count is
more informative than a wall-clock comparison. Exact callback totals may
change with line-search details.

Raw stress is nonconvex, so this seeded run demonstrates a reproducible fit
without claiming that it found the global minimum.

```{r assert-optimization, include = FALSE}
stopifnot(
  all(vapply(results, `[[`, character(1), "status") == "converged"),
  all(vapply(results, `[[`, logical(1), "converged")),
  all(vapply(
    results,
    function(result) result$terminate$what,
    character(1)
  ) == "rel_tol"),
  all(vapply(results, `[[`, numeric(1), "f") < initial_stress),
  isTRUE(all.equal(
    separate_result$par,
    combined_result$par,
    tolerance = 1e-10
  )),
  isTRUE(all.equal(
    separate_result$f,
    combined_result$f,
    tolerance = 1e-10
  )),
  identical(separate_result$nf, combined_result$nf),
  identical(separate_result$ng, combined_result$ng),
  unname(preparation_counts["combined"]) <
    unname(preparation_counts["separate"])
)
```

## Put the configuration on the map

Distances leave absolute position and orientation unidentified. Adding one
vector to every point leaves all distances unchanged, as do rotations and
reflections. A coordinate-only plot therefore cannot tell us where north is,
but it also leaves a more useful question unanswered: after accounting for
those arbitrary choices, where does the recovered shape disagree with the
actual geography?

To make that comparison, we use a small geographic reference bundled with the
article. The map outline and 20 city coordinates come from
[Natural Earth](https://www.naturalearthdata.com/); the Hook of Holland
coordinate comes from [GeoNames](https://www.geonames.org/). The source,
license, and reproduction details are recorded with the data files.

The helper below uses an equirectangular projection centred on Europe. Its
coordinates are approximate kilometres, which is sufficient for a regional
diagnostic rather than a precision map.

```{r geographic-reference}
geography_path <- system.file(
  "extdata",
  "eurodist-geography.csv",
  package = "mize"
)
map_path <- system.file(
  "extdata",
  "europe-map.csv",
  package = "mize"
)
city_geography <- utils::read.csv(geography_path)
europe_map <- utils::read.csv(map_path)

project_geography <- function(longitude, latitude) {
  earth_radius_km <- 6371.009
  reference_longitude <- 9
  reference_latitude <- 48
  degrees_to_radians <- pi / 180

  cbind(
    x = earth_radius_km * cos(reference_latitude * degrees_to_radians) *
      (longitude - reference_longitude) * degrees_to_radians,
    y = earth_radius_km *
      (latitude - reference_latitude) * degrees_to_radians
  )
}
```

The alignment is fitted rather than chosen by eye. It removes translation,
allows rotation or reflection because handedness is unidentified, and uses one
uniform scale factor. The scale separates the overall inflation of road
distances relative to straight-line geography from differences in the
configuration's shape. It is used only for this geographic comparison; the
optimizer's returned configuration and raw stress remain unchanged. The
singular value decomposition finds the least-squares similarity alignment.

```{r align-to-geography}
center_coordinates <- function(par) {
  coordinates <- matrix(par, ncol = 2, byrow = TRUE)
  sweep(coordinates, 2, colMeans(coordinates))
}

align_to_reference <- function(configuration, reference) {
  configuration <- scale(
    configuration,
    center = TRUE,
    scale = FALSE
  )
  reference_center <- colMeans(reference)
  centered_reference <- sweep(reference, 2, reference_center)
  decomposition <- svd(crossprod(configuration, centered_reference))
  orthogonal_transform <- decomposition$u %*% t(decomposition$v)
  scale_factor <- sum(decomposition$d) / sum(configuration^2)

  list(
    coordinates = sweep(
      scale_factor * configuration %*% orthogonal_transform,
      2,
      reference_center,
      "+"
    ),
    scale = scale_factor,
    determinant = det(orthogonal_transform)
  )
}

centered_coordinates <- center_coordinates(combined_result$par)
stopifnot(identical(city_geography$city, labels(target_distances)))
geographic_coordinates <- project_geography(
  city_geography$longitude,
  city_geography$latitude
)
alignment <- align_to_reference(
  centered_coordinates,
  geographic_coordinates
)
aligned_coordinates <- alignment$coordinates
geographic_residual_km <- sqrt(rowSums(
  (aligned_coordinates - geographic_coordinates)^2
))
geographic_rmse_km <- sqrt(mean(geographic_residual_km^2))
```

The fitted scale factor is `r round(alignment$scale, 3)`, and this fit includes
a reflection because the optimized configuration's handedness is arbitrary.
Each orange segment in the map joins a city's geographic location to its
aligned MDS location. The triangle pattern retains the recovered distance
geometry, while the segments make its local errors visible. Their RMSE is about
`r round(geographic_rmse_km)` km after the global similarity alignment. Long
segments at Athens and Stockholm, for example, are information that a polished
but unreferenced coordinate plot would hide.

```{r plot-geographic-alignment, echo = FALSE, fig.width = 7, fig.height = 7, fig.alt = sprintf("Map of Europe comparing the geographic and similarity-aligned metric MDS locations of 21 cities. Black open circles mark geographic locations, blue triangles mark aligned MDS locations, and orange segments show pointwise discrepancies; the residual RMSE is about %.0f kilometres.", geographic_rmse_km)}
map_coordinates <- project_geography(
  europe_map$longitude,
  europe_map$latitude
)
old_par <- graphics::par(
  mar = c(3.6, 3.6, 3.4, 1),
  mgp = c(2.1, 0.65, 0),
  tcl = -0.25
)

graphics::plot(
  range(map_coordinates[, 1]),
  range(map_coordinates[, 2]),
  type = "n",
  asp = 1,
  axes = FALSE,
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Recovered road-distance geometry versus geographic Europe",
  cex.main = 0.95
)
longitude_breaks <- seq(-10, 30, by = 10)
latitude_breaks <- seq(35, 60, by = 5)
grid_x <- project_geography(
  longitude_breaks,
  rep(48, length(longitude_breaks))
)[, 1]
grid_y <- project_geography(
  rep(9, length(latitude_breaks)),
  latitude_breaks
)[, 2]
graphics::abline(
  v = grid_x,
  h = grid_y,
  col = "#E5E7EB",
  lty = 3
)
for (indices in split(seq_len(nrow(europe_map)), europe_map$group)) {
  graphics::polygon(
    map_coordinates[indices, 1],
    map_coordinates[indices, 2],
    col = "#F3F4F6",
    border = "#B8C0C8",
    lwd = 0.55
  )
}
graphics::axis(
  1,
  at = grid_x,
  labels = c("10°W", "0°", "10°E", "20°E", "30°E"),
  cex.axis = 0.75
)
graphics::axis(
  2,
  at = grid_y,
  labels = paste0(latitude_breaks, "°N"),
  las = 1,
  cex.axis = 0.75
)
graphics::box(col = "#8B949E")
graphics::segments(
  geographic_coordinates[, 1],
  geographic_coordinates[, 2],
  aligned_coordinates[, 1],
  aligned_coordinates[, 2],
  col = grDevices::adjustcolor("#D55E00", alpha.f = 0.78),
  lwd = 1.35
)
graphics::points(
  aligned_coordinates,
  pch = 17,
  col = "#0072B2",
  cex = 0.8
)
graphics::points(
  geographic_coordinates,
  pch = 21,
  bg = "white",
  col = "#111827",
  cex = 0.75
)
label_positions <- c(
  4, 2, 1, 2, 2, 4, 4, 4, 2, 4, 2,
  2, 2, 2, 4, 4, 4, 2, 4, 4, 4
)
graphics::text(
  geographic_coordinates[, 1],
  geographic_coordinates[, 2],
  labels = city_geography$city,
  pos = label_positions,
  offset = 0.28,
  cex = 0.55,
  col = "#111827"
)
graphics::legend(
  "topleft",
  legend = c(
    "Geographic location",
    "Aligned MDS location",
    "Residual"
  ),
  pch = c(21, 17, NA),
  pt.bg = c("white", NA, NA),
  col = c("#111827", "#0072B2", "#D55E00"),
  lty = c(NA, NA, 1),
  lwd = c(NA, NA, 1.35),
  pt.cex = c(0.75, 0.8, NA),
  cex = 0.68,
  bg = grDevices::adjustcolor("white", alpha.f = 0.9),
  box.col = "#D1D5DB",
  inset = 0.015
)
graphics::mtext(
  sprintf(
    "Similarity alignment; pointwise residual RMSE = %.0f km",
    geographic_rmse_km
  ),
  side = 3,
  line = 0.25,
  cex = 0.72,
  col = "#4B5563"
)
graphics::par(old_par)
```

```{r assert-geographic-alignment, include = FALSE}
stopifnot(
  nrow(city_geography) == attr(target_distances, "Size"),
  all(is.finite(aligned_coordinates)),
  alignment$scale > 0,
  abs(abs(alignment$determinant) - 1) < 1e-12,
  geographic_rmse_km < 350
)
```

The invariance holds at the objective level. The hidden check below translates,
rotates, and reflects the returned configuration and verifies that stress is
unchanged to numerical tolerance.

```{r assert-stress-invariance, include = FALSE}
angle <- pi / 3
rotation <- matrix(
  c(cos(angle), -sin(angle), sin(angle), cos(angle)),
  nrow = 2,
  byrow = TRUE
)
translated <- sweep(
  centered_coordinates,
  2,
  c(1000, -500),
  "+"
)
rotated <- centered_coordinates %*% rotation
reflected <- centered_coordinates
reflected[, 1] <- -reflected[, 1]

stress_at <- function(coordinates) {
  combined_callbacks$fg$fn(as.vector(t(coordinates)))
}
invariant_stress <- vapply(
  list(translated, rotated, reflected),
  stress_at,
  numeric(1)
)
relative_invariance_error <- max(
  abs(invariant_stress - combined_result$f)
) / max(1, abs(combined_result$f))
stopifnot(relative_invariance_error < 1e-12)
```

Starting from road distances alone, the optimizer recovers a recognizable
geometry even though its coordinate frame is arbitrary. The closure and
combined-callback pattern used here also applies whenever an objective and
gradient share fixed data or an expensive intermediate calculation; `?mize`
describes the full callback contract.

## See also

- [Getting started](mize.html) introduces callback lists and gradient checks.
- [Convergence](convergence.html) explains the relative tolerance used here.
- [Stateful optimization](stateful.html) shows step-by-step integration with
  retained optimizer state.
