---
title: "Stateful optimization"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stateful optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

A stateful optimizer separates configuration from iteration. You keep the
current parameters and optimizer state, decide when to advance them, and can
log, visualize, intervene, or checkpoint between steps.

This is useful when a single blocking call to `mize()` does not leave enough
room for application-specific work. The price of that control is that your
loop owns the lifecycle: it must retain every updated optimizer, check terminal
state, and treat function and gradient observations as optional.

We will use the two-dimensional Rosenbrock function throughout:

```{r example-problem}
rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  }
)
rb0 <- c(-1.2, 1)
```

## Why retaining state matters

A tempting workaround with `stats::optim()` is to run a small batch, take its
parameters, and start another optimizer:

```{r reinitializing-optim}
first_batch <- stats::optim(
  par = rb0,
  fn = rb_fg$fn,
  gr = rb_fg$gr,
  method = "BFGS",
  control = list(maxit = 10)
)
second_batch <- stats::optim(
  par = first_batch$par,
  fn = rb_fg$fn,
  gr = rb_fg$gr,
  method = "BFGS",
  control = list(maxit = 10)
)
c(first = first_batch$value, second = second_batch$value)
```

The second call starts a new BFGS run. It receives the parameters and discards
the first run's inverse-Hessian approximation and other transient method state.
Shorter batches provide more interruptions while throwing away more of the
information that makes a stateful method effective.

## The three lifecycle operations

The stateful API has three main operations:

| Operation | Role |
|:--|:--|
| `make_mize()` | Configure an optimizer; optionally initialize it when `par` and `fg` are available. |
| `mize_init()` | Bind an optimizer to a starting point and callbacks, and initialize a new run. |
| `mize_step()` | Advance one iteration and return updated `opt` and `par`. |

Configuration and initialization can be separate:

```{r create-and-initialize}
opt <- make_mize(method = "BFGS")
opt <- mize_init(
  opt,
  par = rb0,
  fg = rb_fg,
  max_iter = 3,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
```

Or pass `par` and `fg` to `make_mize()` to initialize immediately. In
either form, always use the updated `opt` and `par` returned by each step.

## Optional observations

`mize_step()` always returns `opt`, `par`, `nf`, and `ng`. It returns
`f` or `g` only when that value was calculated at the returned parameters.
For example, constant-step steepest descent needs a gradient at the old
parameters but does not observe either value at the new parameters:

```{r optional-observation-witness}
observation_opt <- make_mize(
  method = "SD",
  line_search = "constant",
  step0 = 0.0001,
  par = rb0,
  fg = rb_fg,
  max_iter = 3,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
observation_step <- mize_step(observation_opt, rb0, rb_fg)
names(observation_step)
```

```{r assert-optional-observation-witness, include = FALSE}
stopifnot(
  is.null(observation_step$f),
  is.null(observation_step$g)
)
```

Code such as `if (step$f < best_f)` is therefore unsafe. An absent observation
is normal and leaves the step status unchanged.

## A minimal terminal-aware loop

A stateful run does not check its ordinary convergence controls until you
summarize a step and call `check_mize_convergence()`. A minimal safe loop
retains state and checks for termination after every call that can update the
optimizer:

```{r minimal-loop}
par <- rb0

while (!opt$is_terminated) {
  par_old <- par
  step <- mize_step(opt, par, rb_fg)
  opt <- step$opt
  par <- step$par
  if (opt$is_terminated) {
    break
  }

  step_info <- mize_step_summary(
    opt,
    par,
    rb_fg,
    par_old = par_old,
    calc_fn = FALSE,
    calc_gr = FALSE
  )
  opt <- step_info$opt
  if (opt$is_terminated) {
    break
  }

  opt <- check_mize_convergence(step_info)
}

opt[c("status", "converged", "message", "terminate")]
```

```{r assert-minimal-loop, include = FALSE}
stopifnot(
  isTRUE(opt$is_terminated),
  identical(opt$status, "budget_exhausted"),
  identical(opt$converged, FALSE),
  identical(opt$terminate$what, "max_iter")
)
```

This run deliberately ends at `max_iter = 3`: its status is
`budget_exhausted`, and `converged` is `FALSE`.

## Deliberate observations and diagnostics

`mize_step_summary()` is where a loop can request observations. Set
`calc_fn = TRUE` when logging or best-value tracking needs the objective, and
`calc_gr = TRUE` when gradient norms are needed. The returned optimizer must
be retained because a requested callback updates the lifetime counts and can
itself terminate at a hard budget.

The summary schema is dynamic. Common fields include `iter`, `f`, `nf`,
`ng`, `step`, and `alpha`; line-search methods can also expose
`alpha_init`, `slope_init`, `ls_reason`, `ls_outcome`, `ls_nf`, and
`ls_ng`. Some directions add `direction_reason`, and momentum methods can
add `mu`. See `?mize_step_summary` for the complete current schema.

Here is one canonical production loop. Its `state` list includes the required
`opt` and `par`, plus caller-owned best and progress records. The best value
is the best post-step objective this loop has actually observed; it is distinct
from the optimizer's current state.

```{r production-loop-definition}
value_or_na <- function(x) {
  if (is.null(x)) NA else x
}

advance_stateful <- function(state, fg, max_steps = Inf) {
  steps <- 0L

  while (!state$opt$is_terminated && steps < max_steps) {
    par_old <- state$par
    step <- mize_step(state$opt, state$par, fg)
    state$opt <- step$opt
    state$par <- step$par
    if (state$opt$is_terminated) {
      break
    }

    step_info <- mize_step_summary(
      state$opt,
      state$par,
      fg,
      par_old = par_old,
      calc_fn = TRUE
    )
    state$opt <- step_info$opt
    if (state$opt$is_terminated) {
      break
    }

    if (!is.null(step_info$f) && is.finite(step_info$f)) {
      if (step_info$f < state$best_f) {
        state$best_f <- step_info$f
        state$best_par <- state$par
      }

      state$progress <- rbind(
        state$progress,
        data.frame(
          iter = step_info$iter,
          f = step_info$f,
          nf = step_info$nf,
          ng = step_info$ng,
          step = step_info$step,
          alpha = value_or_na(step_info$alpha),
          ls_outcome = value_or_na(step_info$ls_outcome)
        )
      )
    }

    state$opt <- check_mize_convergence(step_info)
    steps <- steps + 1L
  }

  state
}
```

We will use a bounded BFGS run so the terminal classification is unambiguous:

```{r production-run}
controls <- list(
  max_iter = 30,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)

new_bfgs_state <- function() {
  opt <- do.call(
    make_mize,
    c(list(method = "BFGS", par = rb0, fg = rb_fg), controls)
  )
  list(
    opt = opt,
    par = rb0,
    best_par = NULL,
    best_f = Inf,
    progress = data.frame()
  )
}

uninterrupted <- advance_stateful(new_bfgs_state(), rb_fg)
tail(uninterrupted$progress, 3)
uninterrupted$opt[c("status", "converged", "message", "terminate")]
```

```{r assert-production-run, include = FALSE}
stopifnot(
  isTRUE(uninterrupted$opt$is_terminated),
  identical(uninterrupted$opt$status, "budget_exhausted"),
  identical(uninterrupted$opt$converged, FALSE),
  identical(uninterrupted$opt$terminate$what, "max_iter"),
  is.finite(uninterrupted$best_f),
  !is.null(uninterrupted$best_par)
)
```

The selected progress columns are useful for this run. Test method-specific
columns for presence before application code relies on them.

### Hard budgets can deny a requested observation

`max_fn` remains hard when `calc_fn = TRUE`. With no function evaluations
available, the summary returns no `f` and its updated
optimizer reports the exhausted budget:

```{r exact-function-budget}
budget_opt <- make_mize(
  method = "SD",
  line_search = "constant",
  step0 = 0.0001,
  par = rb0,
  fg = rb_fg,
  max_fn = 0,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
budget_info <- mize_step_summary(
  budget_opt,
  rb0,
  rb_fg,
  calc_fn = TRUE
)
list(
  f_available = !is.null(budget_info$f),
  nf = budget_info$nf,
  status = budget_info$opt$status,
  terminate = budget_info$opt$terminate
)
```

```{r assert-exact-function-budget, include = FALSE}
stopifnot(
  is.null(budget_info$f),
  identical(budget_info$nf, 0),
  identical(budget_info$opt$status, "budget_exhausted"),
  identical(budget_info$opt$terminate$what, "max_fn")
)
```

This is why the production loop both retains `step_info$opt` and guards
`step_info$f`. Calling `fg$fn()` directly bypasses `mize`'s counters and
limits. Keep direct callback calls outside hard-budget workflows.

## Checkpoint and resume

Checkpoint the complete caller-owned state needed to continue. Here that means
the optimizer and parameters, plus the best observation and progress log. The
callbacks are ordinary program code and are supplied again after restoration.

```{r checkpoint-and-resume}
partial <- advance_stateful(
  new_bfgs_state(),
  rb_fg,
  max_steps = 10
)

checkpoint_file <- tempfile(fileext = ".rds")
saveRDS(partial, checkpoint_file)
rm(partial)

resumed <- readRDS(checkpoint_file)
unlink(checkpoint_file)
resumed <- advance_stateful(resumed, rb_fg)
```

To show what preservation buys us, compare the resumed run with both an
uninterrupted stateful run and a one-shot `mize()` call using the same method
and termination controls:

```{r compare-lifecycles}
one_shot <- do.call(
  mize,
  c(list(par = rb0, fg = rb_fg, method = "BFGS"), controls)
)

comparison <- data.frame(
  run = c("one-shot", "stateful", "checkpoint/resume"),
  objective = c(one_shot$f, uninterrupted$best_f, resumed$best_f),
  nf = c(one_shot$nf, uninterrupted$opt$counts$fn, resumed$opt$counts$fn),
  ng = c(one_shot$ng, uninterrupted$opt$counts$gr, resumed$opt$counts$gr),
  status = c(
    one_shot$status,
    uninterrupted$opt$status,
    resumed$opt$status
  ),
  terminate = c(
    one_shot$terminate$what,
    uninterrupted$opt$terminate$what,
    resumed$opt$terminate$what
  )
)
comparison

comparison_tolerance <- 1e-10
equivalent <- c(
  one_shot_best = isTRUE(all.equal(
    one_shot$par,
    uninterrupted$best_par,
    tolerance = comparison_tolerance
  )),
  one_shot_objective = isTRUE(all.equal(
    one_shot$f,
    uninterrupted$best_f,
    tolerance = comparison_tolerance
  )),
  one_shot_last = isTRUE(all.equal(
    one_shot$last_par,
    uninterrupted$par,
    tolerance = comparison_tolerance
  )),
  resumed_current = isTRUE(all.equal(
    resumed$par,
    uninterrupted$par,
    tolerance = comparison_tolerance
  )),
  resumed_best_par = isTRUE(all.equal(
    resumed$best_par,
    uninterrupted$best_par,
    tolerance = comparison_tolerance
  )),
  resumed_best = isTRUE(all.equal(
    resumed$best_f,
    uninterrupted$best_f,
    tolerance = comparison_tolerance
  ))
)
equivalent
```

```{r assert-compare-lifecycles, include = FALSE}
stopifnot(
  all(equivalent),
  all(comparison$status == "budget_exhausted"),
  all(comparison$terminate == "max_iter"),
  !file.exists(checkpoint_file)
)
```

For this deterministic problem and implementation, all comparisons pass at
`1e-10`. The comparison shows that serializing and restoring the optimizer
preserves method state. Bit-for-bit identity across platforms remains outside
the contract.

## Reinitialization starts a new run

Calling `mize_init()` on an existing optimizer deliberately starts a new run.
It resets the iteration counter, terminal fields, caches, and transient
algorithm state. Function and gradient counts remain optimizer-lifetime totals:

```{r reinitialization-lifecycle}
counts_before <- uninterrupted$opt$counts
reinitialized <- mize_init(
  uninterrupted$opt,
  rb0,
  rb_fg
)

data.frame(
  iter_before = uninterrupted$opt$iter,
  iter_after = reinitialized$iter,
  counts_preserved = identical(counts_before, reinitialized$counts),
  terminated_after = reinitialized$is_terminated
)
```

```{r assert-reinitialization-lifecycle, include = FALSE}
stopifnot(
  identical(reinitialized$iter, 0),
  identical(reinitialized$counts, counts_before),
  identical(reinitialized$is_terminated, FALSE),
  is.null(reinitialized$status),
  is.null(reinitialized$terminate)
)
```

Use reinitialization when a new run is intended; use serialization when the
same run must continue. See `?mize_init`, `?mize_step`,
`?mize_step_summary`, and `?check_mize_convergence` for the complete API.

## See also

- [Getting started](mize.html) introduces the one-shot workflow and result
  contract.
- [Convergence](convergence.html) explains how to choose and interpret
  termination controls.
- [Metric MDS](mmds.html) develops a callback closure suitable for either
  one-shot or stateful use.
