---
title: "Stateful optimization"
author: "James Melville"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stateful optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

A stateful optimizer separates configuration from iteration. You keep the
current parameters and optimizer state, decide when to advance them, and can
log, visualize, apply a caller-owned stopping policy, or checkpoint between
steps.

This is useful when a single blocking call to `mize()` does not leave enough
room for application-specific work. The price of that control is that your
loop owns the lifecycle: it must retain every updated optimizer, check terminal
state, and treat function and gradient observations as optional.

We will use the two-dimensional Rosenbrock function throughout:

```{r example-problem}
rb_fg <- list(
  fn = function(x) {
    100 * (x[2] - x[1] * x[1])^2 + (1 - x[1])^2
  },
  gr = function(x) {
    c(
      -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
      200 * (x[2] - x[1] * x[1])
    )
  }
)
rb0 <- c(-1.2, 1)
```

## Why retaining state matters

A tempting workaround with `stats::optim()` is to run a small batch, take its
parameters, and start another optimizer:

```{r reinitializing-optim}
first_batch <- stats::optim(
  par = rb0,
  fn = rb_fg$fn,
  gr = rb_fg$gr,
  method = "BFGS",
  control = list(maxit = 10)
)
second_batch <- stats::optim(
  par = first_batch$par,
  fn = rb_fg$fn,
  gr = rb_fg$gr,
  method = "BFGS",
  control = list(maxit = 10)
)
c(first = first_batch$value, second = second_batch$value)
```

The second call starts a new BFGS run. It receives the parameters and discards
the first run's inverse-Hessian approximation and other transient method state.
Shorter batches provide more interruptions while throwing away more of the
information that makes a stateful method effective.

## The three lifecycle operations

The stateful API has three main operations:

| Operation | Role |
|:--|:--|
| `make_mize()` | Configure an optimizer; optionally initialize it when `par` and `fg` are available. |
| `mize_init()` | Bind an optimizer to a starting point and callbacks, and initialize a new run. |
| `mize_step()` | Advance one iteration and return updated `opt` and `par`. |

Configuration and initialization can be separate:

```{r create-and-initialize}
opt <- make_mize(method = "BFGS")
opt <- mize_init(
  opt,
  par = rb0,
  fg = rb_fg,
  max_iter = 3,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
```

Or pass `par` and `fg` to `make_mize()` to initialize immediately. In
either form, always use the updated `opt` and `par` returned by each step.

## Optional observations

`mize_step()` always returns `opt`, `par`, `nf`, and `ng`. It returns
`f` or `g` only when that value was calculated at the returned parameters.
For example, constant-step steepest descent needs a gradient at the old
parameters but does not observe either value at the new parameters:

```{r optional-observation-witness}
observation_opt <- make_mize(
  method = "SD",
  line_search = "constant",
  step0 = 0.0001,
  par = rb0,
  fg = rb_fg,
  max_iter = 3,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
observation_step <- mize_step(observation_opt, rb0, rb_fg)
names(observation_step)
```

```{r assert-optional-observation-witness, include = FALSE}
stopifnot(
  is.null(observation_step$f),
  is.null(observation_step$g)
)
```

Code such as `if (step$f < best_f)` is therefore unsafe. An absent observation
is normal and leaves the step status unchanged.

## A minimal terminal-aware loop

> **Important.** `mize_step()` advances the optimizer and may stop on a hard
> callback budget or an immediate method failure. A requested observation from
> `mize_step_summary()` can also stop at a hard callback budget. Ordinary
> numerical tolerances, `max_iter`, non-finite summary observations, and a
> completed line search that selected no usable step are handled by
> `check_mize_convergence()`. A caller-owned loop must retain the optimizer
> returned by each operation, stop if either of the first two operations
> terminates it, and otherwise call the convergence check after every step.

The run below has disabled every ordinary numerical tolerance and demonstrates
only `max_iter`. Its summary therefore sets `calc_fn = FALSE` and
`calc_gr = FALSE` to avoid requesting observations that the convergence check
does not need. A loop that enables function- or gradient-based tolerances
should let `mize_step_summary()` request the corresponding observations.

```{r minimal-loop}
par <- rb0

while (!opt$is_terminated) {
  par_old <- par
  step <- mize_step(opt, par, rb_fg)
  opt <- step$opt
  par <- step$par
  if (opt$is_terminated) {
    break
  }

  step_info <- mize_step_summary(
    opt,
    par,
    rb_fg,
    par_old = par_old,
    calc_fn = FALSE,
    calc_gr = FALSE
  )
  opt <- step_info$opt
  if (opt$is_terminated) {
    break
  }

  opt <- check_mize_convergence(step_info)
}

opt[c("status", "converged", "message", "terminate")]
```

```{r assert-minimal-loop, include = FALSE}
stopifnot(
  isTRUE(opt$is_terminated),
  identical(opt$status, "budget_exhausted"),
  identical(opt$converged, FALSE),
  identical(opt$terminate$what, "max_iter")
)
```

This run deliberately ends at `max_iter = 3`: its status is
`budget_exhausted`, and `converged` is `FALSE`.

## Work safely between steps

The optimizer's stored state describes the current problem and trajectory.
Reading that state and deciding whether to continue leaves it coherent.
Changing the problem, the current parameters, or controls that shape stored
algorithm history begins a new run and should go through `mize_init()`.

| Action | Examples | Treatment |
|:--|:--|:--|
| Observe or decide | Log or plot state, inspect diagnostics, save a checkpoint, update caller-owned records, apply an external stopping rule | Continue with the returned `opt` and `par`. |
| Change the problem | Replace callbacks or captured data, change dimension | Reinitialize. |
| Change the parameter state | Replace `par` independently, or change parameter meaning or scale | Reinitialize. |
| Change trajectory-owning controls | Change method; momentum or restart settings; L-BFGS memory; CG or TN preconditioning; line-search strategy or history-dependent step initialization | Reinitialize. |

Curvature pairs, cached values, momentum, and line-search history all refer to
the trajectory that produced them. Replacing part of that trajectory while
retaining the rest can make the state internally inconsistent. Supply stopping
controls through `make_mize()` or `mize_init()`; this guide does not rely on
direct mutation of `opt$convergence`.

## Deliberate observations and diagnostics

`mize_step_summary()` is where a loop can request observations. Set
`calc_fn = TRUE` when logging or best-value tracking needs the objective, and
`calc_gr = TRUE` when gradient norms are needed. The returned optimizer must
be retained because a requested callback updates the lifetime counts and can
itself terminate at a hard budget.

The summary schema is dynamic. A field is absent when its calculation was not
requested or its method did not produce it:

```{r summary-schema, echo=FALSE}
summary_schema <- data.frame(
  `Field group` = c(
    "Core",
    "Requested observations",
    "Line search",
    "Direction diagnostics",
    "Momentum"
  ),
  Examples = c(
    "`iter`, `nf`, `ng`, `step`",
    "`f`, `g2n`, `ginfn`",
    "`alpha`, `ls_reason`, `ls_outcome`, `ls_nf`, `ls_ng`",
    "`direction_reason`",
    "`mu`"
  ),
  Availability = c(
    "Generally available",
    "When calculated, cached, or requested",
    "Searches that report these diagnostics",
    "Methods that report direction provenance",
    "Methods with a momentum stage"
  ),
  check.names = FALSE
)
knitr::kable(summary_schema)
```

Line searches may also report initialization diagnostics such as `alpha_init`
and `slope_init`. See `?mize_step_summary` for the complete field definitions,
and test optional names before application code relies on them.

The monitored loop follows the same lifecycle on every pass:

1. Advance one step and retain its optimizer and parameters.
2. Stop immediately if the step terminated.
3. Request the observations needed by caller-owned records, and retain that
   optimizer too.
4. Update the best value and append one typed progress record.
5. Apply the ordinary convergence checks.

The starting objective is requested through the same budget-aware summary API
before the first step. The complete helper is available below for copying.

<details>
<summary>Complete executable state helpers</summary>

```{r production-loop-definition}
numeric_or_na <- function(x) {
  if (is.null(x)) NA_real_ else as.numeric(x)
}

character_or_na <- function(x) {
  if (is.null(x)) NA_character_ else as.character(x)
}

progress_record <- function(step_info) {
  data.frame(
    iter = as.integer(step_info$iter),
    f = as.numeric(step_info$f),
    nf = as.integer(step_info$nf),
    ng = as.integer(step_info$ng),
    step = as.numeric(step_info$step),
    alpha = numeric_or_na(step_info$alpha),
    ls_outcome = character_or_na(step_info$ls_outcome)
  )
}

bind_progress <- function(records) {
  if (length(records) == 0L) {
    return(data.frame(
      iter = integer(),
      f = double(),
      nf = integer(),
      ng = integer(),
      step = double(),
      alpha = double(),
      ls_outcome = character()
    ))
  }
  do.call(rbind, records)
}

initialize_stateful <- function(opt, par, fg) {
  initial_info <- mize_step_summary(
    opt,
    par,
    fg,
    calc_fn = TRUE,
    calc_gr = FALSE
  )
  state <- list(
    opt = initial_info$opt,
    par = par,
    best_par = NULL,
    best_f = Inf,
    progress = list()
  )

  if (state$opt$is_terminated || is.null(initial_info$f)) {
    return(state)
  }
  if (!is.finite(initial_info$f)) {
    state$opt <- check_mize_convergence(initial_info)
    return(state)
  }

  state$best_par <- par
  state$best_f <- initial_info$f
  state$progress[[1L]] <- progress_record(initial_info)
  state
}

advance_stateful <- function(state, fg, max_steps = Inf) {
  steps <- 0L

  while (!state$opt$is_terminated && steps < max_steps) {
    par_old <- state$par
    step <- mize_step(state$opt, state$par, fg)
    state$opt <- step$opt
    state$par <- step$par
    if (state$opt$is_terminated) {
      break
    }

    step_info <- mize_step_summary(
      state$opt,
      state$par,
      fg,
      par_old = par_old,
      calc_fn = TRUE
    )
    state$opt <- step_info$opt
    if (state$opt$is_terminated) {
      break
    }

    if (!is.null(step_info$f) && is.finite(step_info$f)) {
      if (step_info$f < state$best_f) {
        state$best_f <- step_info$f
        state$best_par <- state$par
      }

      next_record <- length(state$progress) + 1L
      state$progress[[next_record]] <- progress_record(step_info)
    }

    state$opt <- check_mize_convergence(step_info)
    steps <- steps + 1L
  }

  state
}
```

</details>

We will use a bounded BFGS run so the terminal classification is unambiguous:

```{r production-run}
controls <- list(
  max_iter = 30,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)

new_bfgs_state <- function() {
  opt <- do.call(
    make_mize,
    c(list(method = "BFGS", par = rb0, fg = rb_fg), controls)
  )
  initialize_stateful(opt, rb0, rb_fg)
}

initial_state <- new_bfgs_state()
initial_observation <- data.frame(
  `Starting objective` = initial_state$best_f,
  `Function calls` = initial_state$opt$counts$fn,
  `Progress records` = length(initial_state$progress),
  check.names = FALSE
)
knitr::kable(initial_observation)

uninterrupted <- advance_stateful(initial_state, rb_fg)
uninterrupted_progress <- bind_progress(uninterrupted$progress)
knitr::kable(
  tail(uninterrupted_progress, 3),
  digits = 6,
  row.names = FALSE
)
uninterrupted$opt[c("status", "converged", "message", "terminate")]
```

```{r assert-production-run, include = FALSE}
initial_recheck <- mize_step_summary(
  initial_state$opt,
  initial_state$par,
  rb_fg,
  calc_fn = TRUE,
  calc_gr = FALSE
)
stopifnot(
  identical(initial_state$best_par, rb0),
  identical(initial_state$best_f, rb_fg$fn(rb0)),
  identical(initial_state$opt$counts$fn, 1),
  identical(length(initial_state$progress), 1L),
  identical(initial_recheck$nf, initial_state$opt$counts$fn),
  identical(initial_recheck$f, initial_state$best_f),
  isTRUE(uninterrupted$opt$is_terminated),
  identical(uninterrupted$opt$status, "budget_exhausted"),
  identical(uninterrupted$opt$converged, FALSE),
  identical(uninterrupted$opt$terminate$what, "max_iter"),
  is.finite(uninterrupted$best_f),
  !is.null(uninterrupted$best_par)
)
```

For this BFGS configuration, initialization has not evaluated the objective.
The first summary therefore consumes one function call and records the
starting point as the caller-owned best. A repeated request at the same point
reuses the cached value. Progress remains a list of typed one-row data frames
until `bind_progress()` prepares it for display; method-specific columns should
still be tested for presence before application code relies on them.

### Hard budgets can deny a requested observation

`max_fn` remains hard when `calc_fn = TRUE`. With no function evaluations
available, initialization records neither a best value nor a progress row, and
its updated optimizer reports the exhausted budget:

```{r exact-function-budget}
budget_opt <- make_mize(
  method = "SD",
  line_search = "constant",
  step0 = 0.0001,
  par = rb0,
  fg = rb_fg,
  max_fn = 0,
  abs_tol = NULL,
  rel_tol = NULL,
  grad_tol = NULL,
  ginf_tol = NULL,
  step_tol = NULL
)
budget_state <- initialize_stateful(
  budget_opt,
  rb0,
  rb_fg
)
budget_result <- data.frame(
  `Best recorded` = !is.null(budget_state$best_par),
  `Function calls` = budget_state$opt$counts$fn,
  Status = budget_state$opt$status,
  Reason = budget_state$opt$terminate$what,
  check.names = FALSE
)
knitr::kable(budget_result)
```

```{r assert-exact-function-budget, include = FALSE}
stopifnot(
  is.null(budget_state$best_par),
  identical(budget_state$best_f, Inf),
  identical(length(budget_state$progress), 0L),
  identical(budget_state$opt$counts$fn, 0),
  identical(budget_state$opt$status, "budget_exhausted"),
  identical(budget_state$opt$terminate$what, "max_fn")
)
```

This is why both initialization and the production loop retain the optimizer
returned by a summary and guard the requested value. Calling `fg$fn()` directly
bypasses `mize`'s counters and limits. Keep direct callback calls outside
hard-budget workflows.

## Checkpoint and resume

Checkpoint the complete caller-owned state needed to continue. Here that means
the optimizer and parameters, plus the best observation and progress log. The
callbacks are ordinary program code and are supplied again after restoration.

```{r checkpoint-and-resume}
partial <- advance_stateful(
  new_bfgs_state(),
  rb_fg,
  max_steps = 10
)

checkpoint_file <- tempfile(fileext = ".rds")
saveRDS(partial, checkpoint_file)
rm(partial)

resumed <- readRDS(checkpoint_file)
unlink(checkpoint_file)
checkpoint_progress_is_list <- is.list(resumed$progress)
checkpoint_records_are_typed <- all(vapply(
  resumed$progress,
  is.data.frame,
  logical(1)
))
resumed <- advance_stateful(resumed, rb_fg)
```

To show what preservation buys us, compare the resumed run with both an
uninterrupted stateful run and a one-shot `mize()` call using the same method
and termination controls:

```{r compare-lifecycles}
one_shot <- do.call(
  mize,
  c(list(par = rb0, fg = rb_fg, method = "BFGS"), controls)
)

comparison <- data.frame(
  run = c("one-shot", "stateful", "checkpoint/resume"),
  objective = c(one_shot$f, uninterrupted$best_f, resumed$best_f),
  nf = c(one_shot$nf, uninterrupted$opt$counts$fn, resumed$opt$counts$fn),
  ng = c(one_shot$ng, uninterrupted$opt$counts$gr, resumed$opt$counts$gr),
  status = c(
    one_shot$status,
    uninterrupted$opt$status,
    resumed$opt$status
  ),
  terminate = c(
    one_shot$terminate$what,
    uninterrupted$opt$terminate$what,
    resumed$opt$terminate$what
  )
)
knitr::kable(
  comparison,
  digits = 8,
  col.names = c(
    "Run",
    "Objective",
    "Function calls",
    "Gradient calls",
    "Status",
    "Reason"
  )
)

comparison_tolerance <- 1e-10
equivalent <- c(
  one_shot_best = isTRUE(all.equal(
    one_shot$par,
    uninterrupted$best_par,
    tolerance = comparison_tolerance
  )),
  one_shot_objective = isTRUE(all.equal(
    one_shot$f,
    uninterrupted$best_f,
    tolerance = comparison_tolerance
  )),
  one_shot_last = isTRUE(all.equal(
    one_shot$last_par,
    uninterrupted$par,
    tolerance = comparison_tolerance
  )),
  resumed_current = isTRUE(all.equal(
    resumed$par,
    uninterrupted$par,
    tolerance = comparison_tolerance
  )),
  resumed_best_par = isTRUE(all.equal(
    resumed$best_par,
    uninterrupted$best_par,
    tolerance = comparison_tolerance
  )),
  resumed_best = isTRUE(all.equal(
    resumed$best_f,
    uninterrupted$best_f,
    tolerance = comparison_tolerance
  ))
)
```

```{r assert-compare-lifecycles, include = FALSE}
stopifnot(
  all(equivalent),
  checkpoint_progress_is_list,
  checkpoint_records_are_typed,
  all(comparison$status == "budget_exhausted"),
  all(comparison$terminate == "max_iter"),
  !file.exists(checkpoint_file)
)
```

The resumed run reaches the same current and best parameters as the
uninterrupted run to numerical tolerance, and both agree with the one-shot run.
This is the practical guarantee supplied by checkpointing the complete
optimizer state; exact byte-for-byte identity across different R or platform
builds is not required.

## Reinitialization starts a new run

Calling `mize_init()` on an existing optimizer deliberately starts a new run.
It resets the iteration counter, terminal fields, caches, and transient
algorithm state. Function and gradient counts remain optimizer-lifetime totals:

```{r reinitialization-lifecycle}
counts_before <- uninterrupted$opt$counts
reinitialized <- mize_init(
  uninterrupted$opt,
  rb0,
  rb_fg
)

data.frame(
  iter_before = uninterrupted$opt$iter,
  iter_after = reinitialized$iter,
  counts_preserved = identical(counts_before, reinitialized$counts),
  terminated_after = reinitialized$is_terminated
)
```

```{r assert-reinitialization-lifecycle, include = FALSE}
stopifnot(
  identical(reinitialized$iter, 0),
  identical(reinitialized$counts, counts_before),
  identical(reinitialized$is_terminated, FALSE),
  is.null(reinitialized$status),
  is.null(reinitialized$terminate)
)
```

Use reinitialization when a new run is intended; use serialization when the
same run must continue. See `?mize_init`, `?mize_step`,
`?mize_step_summary`, and `?check_mize_convergence` for the complete API.

## See also

- [Getting started](mize.html) introduces the one-shot workflow and result
  contract.
- [Convergence](convergence.html) explains how to choose and interpret
  termination controls.
- [Metric MDS](mmds.html) develops a callback closure suitable for either
  one-shot or stateful use.
