---
title: "Stateful Optimization"
author: "James Melville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stateful Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
library(mize)
```

By "Stateful" I mean what if we could create an optimizer independently of
the function it was operating on and be able to pass it around, store it, and
get full control over when we pass it data to continue the optimization.

This vignette is about using `mize` to manually control an optimization 
externally. Instead of passing `mize` a function to be optimized from a 
starting point, then waiting for `mize` to finish and get back the finished
results, you might want to tell `mize` to optimize for a few steps, then do
something with the intermediate results: log the cost, update some parameters,
test for some specific convergence criterion, checkpoint the current results,
or plot the current state of the result in some custom way. Then, if there's
still more optimization to be done, pass the results back off to `mize` and
get it to crank away for a few more iterations.

This was in fact the inspiration for creating `mize` in the first place: I
wanted access to the sort of optimization routines that the `stats::optim`
function provided, but the lack of control was a deal breaker. One way to
try and get around the problem is to only optimize for a few iterations at a 
time:

```{r optim example}
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
rb0 <- c(-1.2, 1)

par <- rb0
for (batch in 1:3) {
  optim_res <- stats::optim(par = par, fn = rb_fg$fn, gr = rb_fg$gr, 
                            method = "BFGS", control = list(maxit = 10))
  par <- optim_res$par
  message("batch ", batch, " f = ", formatC(optim_res$value))
}
```

but even this unsatisfactory work-around causes problems, because you are
reinitializing the optimization for with each batch, and losing all the 
information the optimizer has. In the case of methods like BFGS and CG, this
is important for their efficient use. The more control you want, the fewer 
iterations per batch, but that leads to behavior that approaches steepest
descent.

Instead, `mize` lets you create a stateful optimizer, that you pass to a
function, and an updated version of which is returned as part of the return 
value of the function. This gives you complete control over what to do in
between iterations, without sacrificing any of the information the optimizer
is using.

## Creating an Optimizer

To create an optimizer, use the `make_mize` function:

```{r Creating an optimizer}
opt <- make_mize(method = "BFGS")
```

## Initialize the Optimizer

Before starting the optimization, the optimizer needs to be initialized using
the function and starting point. Mainly this is to allow the various methods
to preallocate whatever storage they make use of (matrices and vectors) 
according to the size of the data, as specified by the starting location.

To continue the rosenbrock example from above:

```{r Initializing an optimizer}
opt <- mize_init(opt = opt, par = rb0, fg = rb_fg)
```

### A potential simplification

If you have both the starting point and the function to optimize to hand at
the point when the optimizer is created, you can provide that to `make_mize`
and it will do the initialization for you:

```{r Creating and initializing an optimizer}
opt <- make_mize(method = "BFGS", par = rb0, fg = rb_fg, max_iter = 30)
```

And there is no need to make a separate call to ```mize_init```. However, 
normally it's more convenient to handle configuring the optimizer earlier
than when the data shows up.

## Start optimizing

Using the batch of ten iteration approach we used with `optim` is very similar
with `mize`:

```{r Optimization}
par <- rb0
for (batch in 1:3) {
  for (i in 1:10) {
    mize_res <- mize_step(opt = opt, par = par, fg = rb_fg)
    par <- mize_res$par
    opt <- mize_res$opt
  }
  message("batch ", batch, " f = ", formatC(mize_res$f))
}
```

The difference here is that you have to do the iterating in batches of 10
manually yourself. The optimizer needs to be updated with the version that was
returned from the function.

### Return value of `mize_step`

As you can see, with the greater power of `mize_step` to control the iteration,
comes greater responsibility. You also need to decide when to stop iterating.
Apart from `par` and `opt`, there are some other components to the 
returned result list which might help:

* `f` - The function value, if it was calculated at `par`. For the few methods
which don't do this, you can of course generate it yourself via `rb_fg$fn(par)`.
* `g` - The gradient vector, if it was calculated at `par`. If it's not present, 
then obviously there's nothing to stop you calculating `rb_fg$gr(par)` yourself.
* `nf` - The number of function evaluations carried out over the optimizer's
lifetime. This count persists across `mize_init()` calls. `opt` coordinates it
with `mize_step`, so you don't need to update it manually between steps.
* `ng` - The optimizer-lifetime number of gradient evaluations, which also
persists across `mize_init()` calls.

You should treat the optimizer, `opt`, as a black box. After each step, check
`opt$is_terminated` before relying on optional function or gradient values. If
it is `TRUE`, `opt$terminate$what` gives the termination reason and
`opt$terminate$val` gives the observed criterion value. Termination can indicate
convergence, budget exhaustion, or a non-finite function or gradient.

Taking all that into account, here's a self-contained example, that removes the 
now un-necessary batching, does some minor error checking, and keeps track of
the best parameters seen so far (although with this combination of optimizer
and problem, you don't have to worry about it):

```{r Full example}
# Create the optimizer
opt <- make_mize(method = "BFGS")

# Pretend we don't have access to the function or starting point until later
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
rb0 <- c(-1.2, 1)

# Initialize
opt <- mize_init(opt = opt, par = rb0, fg = rb_fg)

# Store the best seen parameters in case something goes wrong
par <- rb0
par_best <- par
f_best <- rb_fg$fn(par_best)

for (i in 1:30) {
  mize_res <- mize_step(opt = opt, par = par, fg = rb_fg)
  par <- mize_res$par
  opt <- mize_res$opt
  
  # Do whatever you want with the data at each iteration
  
  if (opt$is_terminated) {
    # Stop before using values that may be unavailable after termination
    break
  }
  if (mize_res$f < f_best) {
    f_best <- mize_res$f
    par_best <- par
  }
}

# optimized result is in par_best
par_best
f_best
```


## Step information

The return value of `mize_step` provides the function and gradient values. If
you would like access to more information, the `mize_step_summary` function
can extract it conveniently:

```{r mize_step_summary}
# Create optimizer and do one step of optimization as usual
opt <- make_mize(method = "BFGS", par = rb0, fg = rb_fg)
par <- rb0
par_old <- par
mize_res <- mize_step(opt = opt, par = par, fg = rb_fg)
step_info <- mize_step_summary(
  mize_res$opt, mize_res$par, rb_fg, par_old = par_old
)

# info that's already available in mize_res
step_info$f
step_info$ng
step_info$nf
# and some extra
step_info$step
step_info$alpha
```

`mize_step_summary` takes `opt`, `par` and `fg` like `mize_step` does, but
also optionally wants a `par_old` argument. This is the value of `par`
from the previous iteration, from which it calculates the size of the step
taken in this iteration. 

Information available from the return value of `mize_step_summary` includes:

* `iter` The iteration number.
* `f` The function value, if it's available, or if you have set a convergence 
tolerance that requires its calculation (see below).
* `g2n` The gradient l2 (Euclidean) norm, if `grad_tol` is non-`NULL` (see the
Convergence section for more).
* `ginfn` The gradient infinity norm, if `ginf_tol` is non-`NULL` (also see the
Convergence section for more).
* `nf` The number of function evaluations over the optimizer's lifetime.
* `ng` The number of gradient evaluations over the optimizer's lifetime.
* `step` The step size of this iteration.
* `alpha` The size of the line search value found during the gradient descent
stage. This won't be the same as `step` even for optimizers that don't use an 
extra momentum stage because the total step size is normally the value of 
`alpha` multiplied by the magnitude of the gradient.
* `mu` If a momentum stage was used, the momentum coefficient.
* `opt` The optimizer with updated function and gradient counts, if `f`, 
`g2n`, `ginfn` was calculated.

In many cases, `f`, `g2n` and `ginfn` do not require any recalculation (or
aren't calculated), but to be on the safe side, always reassign `opt` to the
return value from `mize_step_summary`.

Here's a modified version of the previous example, where we log out information
from `mize_step_summary`. We're only going to go for 10 iterations to avoid
too much output.

```{r Example with step summary info}
# Create the optimizer
opt <- make_mize(method = "BFGS", par = rb0, fg = rb_fg)

par <- rb0
for (i in 1:10) {
  par_old <- par
  mize_res <- mize_step(opt = opt, par = par, fg = rb_fg)
  par <- mize_res$par
  opt <- mize_res$opt

  # step info
  step_info <- mize_step_summary(opt, par, rb_fg, par_old)
  opt <- step_info$opt
  message(paste(
    Map(function(x) { paste0(x, " = ", formatC(step_info[[x]])) }, 
        c("iter", "f", "nf", "ng", "step")), 
    collapse = ", "))
}
```

## Convergence

In the example up until now we have manually looped over 30 iterations and then
stopped. More sophisticated stopping criteria is available. Three changes are
needed:

1. When initializing the optimizer, when passing `par` and `fg` to either
`make_mize` or `mize_init`, also pass termination criteria:

```{r Optimizers with convergence info}
opt <- make_mize(method = "BFGS", par = rb0, fg = rb_fg, max_iter = 30)
# or
opt <- make_mize(method = "BFGS")
opt <- mize_init(opt = opt, par = rb0, fg = rb_fg, max_iter = 30)
```

2. At the end of the loop, after calling `mize_step_summary`, pass the return
value to the function `check_mize_convergence`. This returns an updated version
of `opt` which will indicate if optimization should stop by setting the 
`opt$is_terminated` boolean flag:

```{r Checking for convergence}
par <- rb0
par_old <- par
mize_res <- mize_step(opt, par, rb_fg)
par <- mize_res$par
step_info <- mize_step_summary(
  mize_res$opt, par, rb_fg, par_old = par_old
)
opt <- check_mize_convergence(step_info)
```

Note that you don't need to manually assign `opt` to the value that comes from
`mize_step_summary`, as `check_mize_convergence` handles that.

3. Instead of manually looping with a `for` loop you can use 
`while (!opt$is_terminated)`.

Once `opt$is_terminated` is `TRUE`, optimization should cease. Inspect
`opt$terminate$what` for the reason and `opt$terminate$val` for the observed
criterion value. Check the flag immediately after `mize_step()` before doing
anything that assumes a finite function or gradient value.

Apart from just maximum number of iterations, there are a variety of options 
that relate to convergence. There is a separate vignette which covers these
[convergence options](convergence.html), and all the parameters mentioned there
can be passed to `make_mize` and `mize_init`. Whatever options you use, 
setting `max_iter` is a good idea to avoid an infinite loop.

Here's the example repeated again, this time using `check_mize_convergence`
to control the number of iterations, rather than a `for` loop:
```{r Full example with convergence checking}
# Create the optimizer
opt <- make_mize(method = "BFGS")

rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
rb0 <- c(-1.2, 1)

# Initialize and set convergence criteria
opt <- mize_init(opt = opt, par = rb0, fg = rb_fg, max_iter = 30)

# Store the best seen parameters in case something goes wrong
par <- rb0
par_best <- par
f_best <- rb_fg$fn(par_best)

while (!opt$is_terminated) {
  par_old <- par
  mize_res <- mize_step(opt = opt, par = par, fg = rb_fg)
  par <- mize_res$par
  opt <- mize_res$opt
  
  # Do whatever you want with the data at each iteration
  
  if (opt$is_terminated) {
    # Stop before using values that may be unavailable after termination
    break
  }
  if (mize_res$f < f_best) {
    f_best <- mize_res$f
    par_best <- par
  }
  
  step_info <- mize_step_summary(opt, par, rb_fg, par_old)
  # Do something with the step info if you'd like
  # Check convergence
  opt <- check_mize_convergence(step_info)
}

# optimized result is in par_best
par_best
f_best
```
