---
title: "Stateful Optimization"
author: "James Melville"
date: "December 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mize)
```

By "Stateful" I mean what if we could create an optimizer indepedently of
the function it was operating on and be able to pass it around, store it, and
get full control over when we pass it data to continue the optimization.

This vignette is about using `mize` to manually control an optimization 
externally. Instead of passing `mize` a function to be optimized from a 
starting point, then waiting for `mize` to finish and get back the finished
results, you might want to tell `mize` to optimize for a few steps, then do
something with the intermediate results: log the cost, update some parameters,
test for some specific convergence criterion, checkpoint the current results,
or plot the current state of the result in some custom way. Then, if there's
still more optimization to be done, pass the results back off to `mize` and
get it to crank away for a few more iterations.

This was in fact the inspiration for creating `mize` in the first place: I
wanted access to the sort of optimization routines that the `stats::optim`
function provided, but the lack of control was a deal breaker. One way to
try and get around the problem is to only optimize for a few iterations at a 
time:

```{r}
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
rb0 <- c(-1.2, 1)

par <- rb0
for (batch in 1:3) {
  optim_res <- stats::optim(par = par, fn = rb_fg$fn, gr = rb_fg$gr, 
                            method = "BFGS", control = list(maxit = 10))
  par <- optim_res$par
  message("batch ", batch, " f = ", formatC(optim_res$value))
}
```

but even this unsatisfactory work-around causes problems, because you are
reinitializing the optimization for with each batch, and losing all the 
information the optimizer has. In the case of methods like BFGS and CG, this
is important for their efficient use. The more control you want, the fewer 
iterations per batch, but that leads to behavior that approaches steepest
descent.

Instead, `mize` lets you create a stateful optimizer, that you pass to a
function, and an updated version of which is returned as part of the return 
value of the function. This gives you complete control over what to do in
between iterations, without sacrificing any of the information the optimizer
is using.

## Creating an Optimizer

To create an optimizer, use the `make_mize` function:

```{r}
opt <- make_mize(method = "BFGS")
```

## Initialize the Optimizer

Before starting the optimization, the optimizer needs to be initialized using
the function and starting point. Mainly this is to allow the various methods
to preallocate whatever storage they make use of (matrices and vectors) 
according to the size of the data, as specified by the starting location.

To continue the rosenbrock example from above:

```{r}
opt <- mize_init(opt = opt, par = rb0, fg = rb_fg)
```

### A potential simplification

If you have both the starting point and the function to optimize to hand at
the point when the optimizer is created, you can provide that to `make_mize`
and it will do the initialization for you:

```{r}
opt <- make_mize(method = "BFGS", par = rb0, fg = rb_fg)
```

And there is no need to make a separate call to ```mize_init```. However, 
normally it's more convenient to handle configuring the optimizer earlier
than when the data shows up.

## Start optimizing

Using the batch of ten iteration approach we used with `optim` is very similar
with `mize`:

```{r}
par <- rb0
iter <- 0
for (batch in 1:3) {
  for (i in 1:10) {
    iter <- iter + 1
    mize_res <- mize_step(opt = opt, par = par, fg = rb_fg, iter = iter)
    par <- mize_res$par
    opt <- mize_res$opt
  }
  message("batch ", batch, " f = ", formatC(mize_res$f))
}
```

The difference here is that you have to do the iterating in batches of 10 
manualy yourself, remembering to increment the iteration counter and pass it
to `mize_step`. Plus, the optimizer needs to be updated with the version that
was returned from the function.

### Return value of `mize_step`

As you can see, with the greater power of `mize_step` to control the iteration,
comes greater responsibility. You also need to decide when to stop iterating.
Apart from `par` and `opt`, there are some other components to the 
returned result list which might help:

* `f` - The function value, if it was calculated at `par`. For the few methods
which don't do this, you can of course generate it yourself via `rb_fg$fn(par)`.
* `g2n` - The 2-norm (aka Euclidean norm) of the gradient vector, if it was
calculated at `par`. If it's not present, then obviously there's nothing to stop
you calculating `rb_fg$gr(par)` yourself.
* `nf` - The number of function evaluations carried out so far (i.e. since 
initialization). `opt` is also keeping track of this, and coordinates with 
`mize_step`, so you don't need to manually update this yourself between steps.
* `ng` - The number of gradient evaluations carried out so far.

You should treat the optimizer, `opt`, as a black box and not examine its 
horrific innards, except to check whether `opt$error` is non-`NULL`. If it's
anything other than `NULL`, then this means something really bad has happened
during the optimization, most likely a `NaN` or `Inf` was calculated in
the gradient. This can happen with a very poorly chosen starting point, and
a combination of descent method and line search which doesn't guarantee descent,
such as a very aggressive momentum scheme or more likely an adaptive learning
rate technique like delta-bar-delta. Monitoring the function value or the 
size of the change in `par` between iterations can help spot an imminent 
divergence.


Taking all that into account, here's a self-contained example, that removes the 
now un-necessary batching, does some minor error checking, and keeps track of
the best parameters seen so far (although with this combination of optimizer
and problem, you don't have to worry about it):

```{r}
# Create the optimizer
opt <- make_mize(method = "BFGS")

# Pretend we don't have access to the function or starting point until later
rb_fg <- list(
   fn = function(x) { 100 * (x[2] - x[1] * x[1]) ^ 2 + (1 - x[1]) ^ 2  },
   gr = function(x) { c( -400 * x[1] * (x[2] - x[1] * x[1]) - 2 * (1 - x[1]),
                          200 *        (x[2] - x[1] * x[1])) })
rb0 <- c(-1.2, 1)

# Initialize
opt <- mize_init(opt = opt, par = rb0, fg = rb_fg)

# Store the best seen parameters in case something goes wrong
par <- rb0
par_best <- par
f_best <- rb_fg$fn(par_best)

iter <- 0
for (i in 1:30) {
  iter <- iter + 1
  mize_res <- mize_step(opt = opt, par = par, fg = rb_fg, iter = iter)
  par <- mize_res$par
  opt <- mize_res$opt
  
  # Do whatever you want with the data at each iteration
  
  if (!is.null(opt$error)) {
    break
  }
  if (mize_res$f < f_best) {
    f_best <- mize_res$f
    par_best <- par
  }
}

# optimized result is in par_best
```


