---
title: "Nesterov Accelerated Gradient and Momentum"
description: >
  Background notes on Nesterov accelerated gradient and its relationship to
  momentum methods.
date: "December 31, 2016"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
---

*December 16 2021*: I have rewritten a large chunk of this document to make the
derivations a little clearer and added a brief section on the definition of
Nesterov momentum used in the NAdam paper.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

A way to express Nesterov Accelerated Gradient (NAG) in terms of a regular
momentum update was noted by [Sutskever and
co-workers](https://proceedings.mlr.press/v28/sutskever13.html), and
perhaps more importantly, when it came to training neural networks, it seemed to
work better than classical momentum schemes. This was further confirmed by
[Bengio and co-workers](https://arxiv.org/abs/1212.0901), who provided an
alternative formulation that might be easier to integrate into existing
software.

I implemented NAG as part of [mize](https://github.com/jlmelville/mize), but
found it was a bit more difficult than I'd anticipated to test that I'd got the
correct results.

It seems I am not alone in being confused by the exact details of implementing
NAG as a momentum scheme (e.g. comments and issues in the deep learning projects
[keras](https://github.com/fchollet/keras/issues/966) and
[mocha.jl](https://github.com/pluskid/Mocha.jl/pull/47), which I found from
random googling), so mainly for the benefit of future me, I am going to derive
the Sutskever and Bengio formulations from NAG in tedious detail. I am also
going to derive an alternative expression that I ended up using in mize.
Finally, I will write some R code to demonstrate their equivalence. This will
turn out to be trickier than I anticipated.

I originally tried to stick with the notation used by Sutskever but in
retrospect that was a mistake. It's too hard to keep everything straight between
the different papers. I will borrow some symbols and try and be very clear about
what they mean and what each expression applies to. Translating between this
document and the original papers may require mentally transposing some symbols
and shifting iteration indices from $t$ to $t+1$ or $t-1$.

Apart from the Sutskever and Bengio papers linked to above, it's worth checking
out the [appendix to the Sutskever paper
(PDF)](https://proceedings.mlr.press/v28/sutskever13-supp.pdf) or the
relevant part (chapter 7) of [Sutskever's
thesis](https://hdl.handle.net/1807/36012). For an introduction to NAG itself,
try the first part of this paper by [O'Donoghue and
Candès](https://arxiv.org/abs/1204.3982) (the rest of it is also good but not
germane to Nesterov momentum expressions).

## Definitions

The goal is to optimize some parameters, $\theta$. This is going to involve
gradient descent, so we will be evaluating the gradient of an objective
function of those parameters, $\nabla f\left(\theta\right)$, and moving a
certain distance in the direction of the negative of the gradient, the distance
being related to the learning rate, $\varepsilon$. There will also be a
momentum term involved, with momentum coefficient $\mu$. The value of the
parameters, learning rate and momentum at iteration $t$ will be indicated by a
subscript, e.g. $\theta_t$.

### The Update Vector

A definition which holds across all the methods discussed here is that the
parameters at iteration $t+1$ are related to the parameters at iteration $t$
by an update that involves the addition of a velocity vector, $v$:

$$\theta_{t+1} = \theta_t + v_{t+1} \implies v_{t+1} = \theta_{t+1} - \theta_t$$

This is pretty obvious, but the expressions we will be dealing with here use a
recursive definition of $v_t$, i.e. to usefully define these updates we need to
express $v_t$ in terms of $v_{t-1}$. The exact expression depends on whether we
are using classical momentum or NAG. In both cases, it would make sense to label
a particular vector as $v_t$, but due to the similarity in how they work, the
temptation to substitute the classical momentum recursive definition of $v_t$
into an expression for NAG is almost overwhelming, and also always wrong.

To confuse matters further, some papers (see the Bengio paper for example)
directly use $v_t$ in expressions that update parameters, but where the $v_t$
doesn't actually apply to the parameter being updated.

Finally, I'll try to be strict about names. When I'm referring to the original
Nesterov Accelerated Gradient, I'll call that NAG. When I'm referring to the
versions that recast it as a momentum scheme, I'll call that Nesterov momentum,
and refer to either the Sutskever or Bengio formulation where necessary. The
traditional momentum scheme is referred to as "standard" or "regular" momentum
by Bengio, and "classical" by Sutskever. Other authors call it "Polyak" or
"heavy ball" momentum. I'll refer to it as "classical" momentum.

## Classical Momentum

The classical momentum velocity vector is defined by Bengio as:

$$
v_t = \mu_{t-1}v_{t-1} - \varepsilon_{t-1}\nabla f\left(\theta_{t-1}\right)
$$

Sutskever gives the same definition but without any $t$ subscript on the
velocity vector or the momentum coefficient.

We can write out the full update for classical momentum as:

$$\theta_{t+1} = \theta_t + \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t\right)$$

with velocity:

$$v_{t+1} = \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t \right)$$

Later we will want to use the recursive definition of momentum, but it won't
be the full update, so we will not be able to define it as $v_t$. In those
circumstances, I will call the update $m_t$ to avoid confusion:

$$
m_{t+1} = \mu_t m_t - \varepsilon_t \nabla f\left(\theta_t \right)
$$

Remember: $v_{t+1} = m_{t+1}$ *only* in the case of classical momentum.

## NAG

The Nesterov Accelerated Gradient method consists of a gradient descent step,
followed by something that looks a lot like a momentum term, but isn't exactly
the same as that found in classical momentum. I'll call it a "momentum stage"
here. If you look at the Sutskever paper, It's important to note that the
parameters being minimized by NAG are given the symbol $y$, not $\theta$. You'll
see that $\theta$ is the symbol for the parameters after they've been updated by
the gradient descent stage, but before the momentum stage.

I'm going to stick with $\theta_t$ as the parameters being updated. But we do
need to define a symbol that accounts for the state of the parameters after
the gradient descent stage. I'll go with $\phi_{t+1}$. Whether you consider
the intermediate step to be part of iterate $t$ or $t+1$ doesn't really matter,
but different papers use different conventions, so you will need to keep track
of it all or risk getting very confused. In this case I am going with $t+1$ so
it's consistent with any new quantities calculated in an iteration (like
$v_{t+1}$).

Here's the gradient descent stage:

$$
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
$$

followed by the momentum-like stage:

$$
\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right)
$$

That concludes one iteration of NAG. The hard part is choosing the learning rate and momentum to get
the convergence guarantees that make the method attractive. Here I'll stick to comparing the update
expressions, using fixed values in the examples later on to keep things simple.

## Sutskever Nesterov Momentum

The key idea behind the Sutskever momentum derivation is to shift the
perspective about which of the parameters we want as the result of the
iteration, from $\theta$ to $\phi$. Rather than having the optimization
iterations proceed as "gradient descent, momentum (end of iteration 1), gradient
descent, momentum (end of iteration 2), gradient descent etc." move the boundary
of where the iterations end by a half-iteration to get "momentum, gradient
descent (end of iteration 1), momentum, gradient descent (end of iteration 2)
etc.". This leaves a phantom gradient descent step that used to be first stage
of the first iteration, now floating in the nether regions of iteration zero,
but you can just pretend that the starting position is the result of gradient
descent from some other arbitrary starting position.

Perhaps a visualization will help. Here are two full iterations of NAG in its
standard form, separated by boxes so you can see where one iteration ends and
the other begins.

$$
\begin{aligned}
\boxed{\begin{aligned}
\phi_{t+1} &= \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) \\
\theta_{t+1} &= \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right)
\end{aligned}} \\[0.75em]
\boxed{\begin{aligned}
\phi_{t+2} &= \theta_{t+1} - \varepsilon_{t+1} \nabla f\left(\theta_{t+1} \right) \\
\theta_{t+2} &= \phi_{t+2} + \mu_{t+1}\left(\phi_{t+2} - \phi_{t+1} \right)
\end{aligned}}
\end{aligned}
$$

The Sutskever derivation says shift the boundaries of the iteration so that it
starts one stage later. This is the same four equations as above, but with a
box showing the new definition of the iteration:

$$
\begin{gathered}
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) \\
\boxed{\begin{aligned}
\theta_{t+1} &= \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right) \\
\phi_{t+2} &= \theta_{t+1} - \varepsilon_{t+1} \nabla f\left(\theta_{t+1} \right)
\end{aligned}} \\
\theta_{t+2} = \phi_{t+2} + \mu_{t+1}\left(\phi_{t+2} - \phi_{t+1} \right)
\end{gathered}
$$

Now that the definition of where we start and finish each iteration has changed, let's shift all the
indices in the box back one step. That includes the momentum coefficient, which becomes $\mu_{t-1}$:

$$
\theta_t = \phi_t + \mu_{t-1}\left(\phi_t - \phi_{t-1} \right) \\
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
$$

Then because our update is now in terms of $\phi$ rather than $\theta$, we can label
$\phi_t - \phi_{t-1}$ as $v_t$. The lookahead point is then:

$$
\theta_t = \phi_t + \mu_{t-1} v_t
$$

And we can now substitute that definition of $\theta_t$ into the gradient stage, which lets us write
this half-shifted NAG iteration in one line:

$$
\phi_{t+1} = \phi_t + \mu_{t-1} v_t
- \varepsilon_t \nabla f\left(\phi_t + \mu_{t-1} v_t \right)
$$

If you are comparing this with a version that uses $\mu_t$, remember to shift the momentum schedule
along with the iteration numbers. With constant momentum, that's one less thing to worry about.

There you have it: this looks just like the classical momentum update, except that the gradient is
calculated at $\phi_t + \mu_{t-1}v_t$, after the momentum update. Classical momentum would evaluate
it at $\phi_t$. Hence, one can do NAG
by simply reversing the order in which the update is usually carried out: do the
momentum stage first, update the parameters, and then do the gradient descent
part. Just like Sutskever said you could.

## Bengio Nesterov Momentum

I found working through the Bengio version of Nesterov momentum a bit
brain-breaking. It starts from the Sutskever definition and then defines a new
variable, $\Theta$, which represents the parameters after the momentum update.
But that just moves us one half-step forward in the iteration, so we are back to
doing things in the original NAG way: gradient descent then momentum-style
update. So their paper's $\Theta$ is the same as our $\theta$. Its explanation
of "committing to the 'peekedahead' parameters" and then "backtracking by the
same amount before each update", also never landed with me (this is a reflection
on my mental capacity, not the paper). Finally, it expresses the update in terms
of $v$, but that's still the the update in terms of the difference in $\phi$,
not $\theta$. Unlike this document it also uses $t-1$ and $t$ to index the
parameters that are being updated, so you will have to remember to add or
subtract 1 from the subscripts when looking between here and the paper, but
that's the least of your problems honestly.

It's all too much. I can't get their explanation to stick in my brain, even
though I can follow the individual steps and get the right expression. So I will
show you how to arrange the standard NAG expressions into their form, rather
than starting from the Sutskever result.

First, let's remove the confusion around using $v$ as an update vector which
isn't the change in the parameters that we are updating each iteration. Instead
of $\phi_{t+1} - \phi_{t} = v_{t+1}$, define:

$$
\phi_{t+1} - \phi_{t} = b_{t+1}
$$

(the 'b' is for Bengio). Our goal now is to end up with an expression that is
defined using $\theta_t$ (no $\phi_t$ allowed) and $b_t$.

I'll use $s_t$ as shorthand for the gradient step:

$$
s_t = -\varepsilon_t \nabla f\left(\theta_t\right).
$$

The negative sign and the learning rate are both part of $s_t$.

Now we are going to go back to using the original NAG expression so to help you
scrub all the shifts in $t$ from your mind here is the expression again:

$$
\phi_{t+1} = \theta_t + s_t \\
\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right)
$$
Start by replacing all uses of $\phi_{t+1}$ in the second equation with the definition
in terms of $\theta_t$ given in the first equation:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \theta_t + s_t +
  \mu_t \left( \theta_t + s_t - \phi_{t} \right) \\
 & = \left(1 + \mu_t \right) \theta_t
 + \left(1 + \mu_t \right) s_t
  - \mu_t \phi_t
\end{split}
\end{equation}
$$

So far, so unpromising. Time to get $b_t$ involved. Remember how we defined:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right) \\
 & = \phi_{t+1} + \mu_t b_{t+1}
\end{split}
\end{equation}
$$

Let's shift everything back one iteration:

$$
\theta_{t} = \phi_{t} + \mu_{t-1} b_{t} \\
\implies  \phi_{t} = \theta_{t} - \mu_{t-1} b_{t}
$$

And now substitute for $\phi_t$ in the last term in our NAG update:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \left(1 + \mu_t \right) \theta_t
 + \left(1 + \mu_t \right) s_t
  - \mu_t \phi_t \\
 & = \left(1 + \mu_t \right) \theta_t
 + \left(1 + \mu_t \right) s_t
  - \mu_{t} \theta_t + \mu_{t-1} \mu_t b_t
\end{split}
\end{equation}
$$

The final step is to gather together the $\theta_t$ terms and the result is:

$$
\theta_{t+1} = \theta_t + \mu_{t-1} \mu_t b_t
 + \left(1 + \mu_t \right) s_t
$$

Ta da. There it is, just like in the paper, subject to the differences in
symbols and when the iteration ticks over from $t$ to $t+1$ and so on.

The advantage of this expression for the Nesterov momentum is that it doesn't
require calculating a gradient at a non-standard position, and only requires a
modification to the coefficients used to calculate the velocity, which is
probably an easier change to make to an existing codebase which already uses
classical momentum. However, because this expression is recursive on $b_t$
rather than the update vector $v_t$, you will need to remember to store a
different vector than you would with classical momentum. It's not really clear
to me whether that is an easy or hard thing to do with most optimization code
bases vs having to make code changes to calculate the gradient at a different
location than the current parameter vector, as you have to do with the Sutskever
version.

You should also remember that Bengio gives us the parameters after the momentum stage, while
Sutskever gives us the parameters after gradient descent. They're half a step out of sync. To get
matching results, we'll need to compare the same stage and get the starting vectors and momentum
coefficients lined up. This will occupy us for quite a while later on.

## An Alternative Expression for NAG

Let's return to the two stages of NAG, keeping our original notation: $\phi_{t+1}$ is the position
after gradient descent, and $\theta_{t+1}$ is the position after the momentum stage. This time we'll
write the update in terms of the actual displacement of $\theta$, rather than the Bengio buffer.

$$\phi_{t+1} = \theta_t + s_t$$

$$\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_t \right)$$


Now, let's just write out the momentum stage in terms of $\theta$, substituting
$\phi$ wherever we find it:

$$
\theta_{t+1} = \theta_t + s_t
+ \mu_t \left[\theta_t + s_t - \theta_{t-1} - s_{t-1}\right]
$$

Rearranging:

$$
\theta_{t+1} = \theta_t
+ \mu_t \left[\theta_t - \theta_{t-1} - s_{t-1} + s_t\right] + s_t
$$

Finally, we can substitute in $v_t$ for the first two terms in the square
brackets, to give:

$$
\theta_{t+1} = \theta_t + \mu_t \left[v_t - s_{t-1} + s_t\right] + s_t
$$

with velocity:

$$
v_{t+1} = \mu_t \left[v_t - s_{t-1} + s_t\right] + s_t
$$

This looks a lot like the classical momentum expression, but with the
velocity vector modified to first remove the contribution of the gradient
descent from the previous iteration, and replace it with the gradient descent
contribution from the *current* iteration. Gives an interesting insight into
the idea of the Nesterov momentum using a form of "lookahead" with the
gradient descent.

You could also choose to expand the velocity expression to make it look a bit
like the Bengio formulation:

$$
v_{t+1} = \mu_t \left[v_t - s_{t-1}\right] + \left(1 + \mu_t\right) s_t
$$

but as this version can't be expressed as the classical momentum form with
different coefficients, the way the Bengio formulation can be, it probably
doesn't gain you anything in terms of implementation, except you can expand it
and rearrange it further to give:

$$
v_{t+1} = \mu_t v_t + s_t + \mu_t\left[s_t - s_{t-1}\right]
$$
which now resembles classical momentum with an extra correction term. For a fixed learning rate,
this is the form in equation (1.4) of [Shi and
co-workers](https://link.springer.com/article/10.1007/s10107-021-01681-8), who call the extra term a
"gradient correction". User 'denis' also uses this expression in an
[answer on the Cross Validated Stack Exchange](https://stats.stackexchange.com/a/233430), calling it
"gradient momentum". I'll use "gradient correction" here.

*December 25 2025*: Here's a possible insight into the meaning of the gradient correction.

The correction is the change in the gradient step, scaled by $\mu_t$. When the steps are similar in
both direction and size, $s_t - s_{t-1} \approx 0$, the update looks a lot like classical momentum. But
consider a steep ravine where we are bouncing from one side to the other. If the step roughly
reverses each iteration, $s_t \approx -s_{t-1}$, the last term becomes approximately $2\mu_t s_t$.
This gives the current gradient step more say in where we go next. Remember that $s_t$ includes the
learning rate, so changing that will also affect this term.

### What happens on a quadratic?

Let's start with just one parameter, so our quadratic objective is an ordinary function:

$$
f(\theta) = \frac{1}{2}a\theta^2 + c\theta.
$$

Here $a$ and $c$ are constants. Taking the derivative gives:

$$
f'(\theta) = a\theta + c.
$$

The gradient symbol $\nabla f$ we've been using is just this derivative when there is only one
parameter. So the difference between the gradients at two successive positions is:

$$
\begin{aligned}
f'(\theta_t) - f'(\theta_{t-1})
&= (a\theta_t + c) - (a\theta_{t-1} + c) \\
&= a(\theta_t - \theta_{t-1}) \\
&= a v_t.
\end{aligned}
$$

The last line uses our definition of the velocity, $v_t = \theta_t - \theta_{t-1}$. With a fixed
learning rate, $s_t = -\varepsilon f'(\theta_t)$, so:

$$
s_t - s_{t-1} = -\varepsilon\left[f'(\theta_t) - f'(\theta_{t-1})\right]
= -\varepsilon a v_t.
$$

Now keep the momentum coefficient $\mu$ fixed too, and substitute into our NAG expression:

$$
\begin{aligned}
v_{t+1} &= \mu v_t + s_t + \mu(s_t - s_{t-1}) \\
&= \mu v_t - \varepsilon f'(\theta_t) - \mu\varepsilon a v_t \\
&= \mu(1 - \varepsilon a)v_t - \varepsilon f'(\theta_t).
\end{aligned}
$$

This looks like classical momentum with a momentum coefficient of
$\mu(1-\varepsilon a)$. And $a=f''(\theta)$ is the curvature of our quadratic. NAG keeps nearly all
its momentum where the curvature is close to zero, and reduces it as the positive curvature
increases, reaching zero at $\varepsilon a=1$. Beyond that, the momentum contribution reverses
direction. Negative curvature increases the momentum instead.

This result generalizes to quadratic objectives with multiple parameters; see section A.2 of the
[supplement to Sutskever et al](https://proceedings.mlr.press/v28/sutskever13-supp.pdf) for the
derivation.

We can also choose how much of this correction to apply, which brings us to unified momentum.

## Unified Momentum

The earliest example of unified momentum I've found is in a 2016 paper by
[Yang, Lin and Li](https://arxiv.org/abs/1604.03257).

The idea is to introduce an extra parameter, $\lambda_t$, that lets us choose how much of the
gradient step correction to use:

$$
v_{t+1} = \mu_t v_t + s_t + \lambda_t\mu_t(s_t-s_{t-1})
$$
Setting $\lambda_t = 0$ gives classical momentum and $\lambda_t = 1$ gives NAG.

For our one-parameter quadratic, the same substitution we used above gives:

$$
v_{t+1} = \mu(1 - \lambda\varepsilon a)v_t - \varepsilon f'(\theta_t).
$$

So $\lambda$ controls how much the curvature changes the momentum coefficient.

### Storing the NAG update

The gradient correction seems to require storing the previous gradient step as well as the
velocity. For NAG, where $\lambda_t=1$, we can combine them into one saved vector:

$$
r_t = v_t - s_{t-1}.
$$

This has a meaning in terms of the positions we've already defined. Since
$\phi_t=\theta_{t-1}+s_{t-1}$, after a completed update we have:

$$
\begin{aligned}
r_t &= (\theta_t-\theta_{t-1})-s_{t-1} \\
&= \theta_t-\phi_t \\
&= \mu_{t-1}b_t.
\end{aligned}
$$

So the saved vector is the gap between the gradient descent and momentum positions. Substituting
$v_t-s_{t-1}=r_t$ into the NAG update gives $v_{t+1}=s_t+\mu_t(r_t+s_t)$. Removing $s_t$ to prepare
the saved vector for the next iteration leaves us with:

$$
\begin{aligned}
s_t &= -\varepsilon_t\nabla f(\theta_t), \\
r_{t+1} &= \mu_t(r_t+s_t), \\
\theta_{t+1} &= \theta_t+s_t+r_{t+1}.
\end{aligned}
$$

Only $r_t$ needs to persist alongside the parameters; the actual displacement is
$v_{t+1}=s_t+r_{t+1}$. These equations also work when the learning rate or momentum changes.

Start with $r_0=0$, corresponding to $\theta_0=\phi_0$. The first displacement is then
$(1+\mu_0)s_0$. Setting $\mu_0=0$ gives an ordinary gradient step; using nonzero momentum gives a
longer first step. We'll return to that choice when we unroll the updates.

## Unrolling NAG and Classical Momentum

As a way to understand the difference between classical momentum and NAG,
let's write out the first few steps of the optimization, substituting in the
recursive definitions with the previous step, and see what emerges.

I'll assume that the momentum is constant, so we can write $\mu$ without a subscript.

Initial coordinates are at $\theta_0$.

### Classical Momentum

With constant momentum, the classical momentum update is:

$$
\theta_{t+1} = \theta_t + s_t + \mu v_t
$$

And the first four updates (including steepest descent on the first iteration)
are as follows:

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + s_0 \\
\\
\theta_2 & = \theta_1 + s_1 + \mu v_1 \\
 & = \theta_1 + s_1 + \mu s_0 \\
\\
\theta_3 & = \theta_2 + s_2 + \mu v_2 \\
 & = \theta_2 + s_2 + \mu \left( s_1 + \mu s_0 \right) \\
 & = \theta_2 + s_2 + \mu s_1 + \mu^2 s_0 \\
\\
\theta_4 & = \theta_3 + s_3 + \mu v_3 \\
 & = \theta_3 + s_3 + \mu \left( s_2 + \mu s_1 + \mu^2 s_0 \right)\\
 & = \theta_3 + s_3 + \mu s_2 + \mu^2 s_1 + \mu^3 s_0 \\
\end{split}
\end{equation}
$$

### NAG

We'll use this straight-forward expression for NAG:

$$
\theta_{t+1} = \theta_t + s_t
+ \mu \left(\theta_t + s_t
- \theta_{t-1} - s_{t-1}
\right)
$$

with just this bit of re-arranging so terms are all grouped together and the
$v_t$ is at the end, which makes it easier to follow the substitution of $v_t$:

$$
\theta_{t+1} = \theta_t + \left(1 + \mu \right) s_t - \mu s_{t-1} + \mu v_t
$$

#### The first step

*December 14 2021*: This section describing your choices for the first step is new.

Like classical momentum, we have no previous directions to draw on for the first step. So let's
start both sets of parameters at the same place, $\theta_0=\phi_0$, and see what happens. The
gradient stage gives us:

$$
\phi_1 = \theta_0 + s_0,
$$

and then the momentum stage gives us:

$$
\theta_1 = \phi_1 + \mu_0(\phi_1-\phi_0)
= \theta_0 + (1+\mu_0)s_0.
$$

There's the longer first step: with nonzero momentum, we move $(1+\mu_0)$ times as far along the
negative gradient. The momentum stage already has something to work with, because $\phi_1-\phi_0$ is
the gradient step we just took.

If we set $\mu_0=0$, the two stages finish at the same point and we get the same first step as
classical momentum. This is what the schedule in the [Sutskever
supplement](https://proceedings.mlr.press/v28/sutskever13-supp.pdf) does. We'll look at both
choices, because this apparently minor detail will turn up again when we try to get the code to
agree.

First, here's the "long" version, using the same nonzero $\mu$ from the start and taking our results
after the momentum stage:

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + \left( 1 + \mu \right) s_0 \\
\\
\theta_2 & = \theta_1 + \left( 1 + \mu \right) s_1 - \mu s_0 + \mu v_1 \\
 & = \theta_1 + \left(1 + \mu \right) s_1 - \mu s_0 + \mu \left( 1 + \mu \right) s_0 \\
 & = \theta_1 + \left(1 + \mu \right) s_1 + \mu^2 s_0 \\
\\
\theta_3 & = \theta_2 + \left( 1 + \mu \right) s_2 - \mu s_1 + \mu v_2 \\
& = \theta_2 + \left(1 + \mu \right) s_2 - \mu s_1 + \mu \left(1 + \mu \right) s_1 + \mu^3 s_0\\
& = \theta_2 + \left(1 + \mu \right) s_2 + \mu^2 s_1 + \mu^3 s_0 \\
\\
\theta_4 & = \theta_3 + \left(1 + \mu \right) s_3 - \mu s_2 + \mu v_3 \\
& = \theta_3 + \left(1 + \mu \right) s_3 - \mu s_2 + \mu \left[ \left(1 + \mu \right) s_2 + \mu^2 s_1 + \mu^3 s_0 \right] \\
& = \theta_3 + \left(1 + \mu \right) s_3 - \mu s_2 + \mu \left(1 + \mu \right) s_2 + \mu^3 s_1  + \mu^4 s_0 \\
& = \theta_3 + \left(1 + \mu \right) s_3 + \mu^2 s_2 + \mu^3 s_1 + \mu^4 s_0
\end{split}
\end{equation}
$$

You can see a pattern emerging here at this point, I hope. These are quite
similar in form compared to the classical momentum.

### NAG with a "short" first step

What happens if we set $\mu_0=0$ and then use the fixed value $\mu$ from the second iteration
onwards? We start with the same gradient step as classical momentum. Here's how that plays out:

$$
\begin{equation}
\begin{split}
\theta_{1} & = \theta_0 + s_0 \\
\\
\theta_2 & = \theta_1 + \left( 1 + \mu \right) s_1 - \mu s_0 + \mu v_1 \\
 & = \theta_1 + \left( 1 + \mu \right) s_1 - \mu s_0 + \mu s_0 \\
 & = \theta_1 + \left( 1 + \mu \right) s_1 \\
\\
\theta_3 & = \theta_2 + \left( 1 + \mu \right) s_2 - \mu s_1 + \mu v_2 \\
& = \theta_2 + \left(1 + \mu \right) s_2 - \mu s_1 + \mu \left(1 + \mu \right) s_1 \\
& = \theta_2 + \left(1 + \mu \right) s_2 + \mu^2 s_1 \\
\\
\theta_4 & = \theta_3 + \left( 1 + \mu \right) s_3 - \mu s_2 + \mu v_3 \\
& = \theta_3 + \left( 1 + \mu \right) s_3 - \mu s_2 + \mu \left[ \left( 1 + \mu \right) s_2 + \mu^2 s_1 \right] \\
& = \theta_3 + \left( 1 + \mu \right) s_3 - \mu s_2 + \mu \left( 1 + \mu \right) s_2 + \mu^3 s_1 \\
& = \theta_3 + \left( 1 + \mu \right) s_3 + \mu^2 s_2 + \mu^3 s_1
\end{split}
\end{equation}
$$

Overall the results aren't that different: nearly all the terms have the same coefficients. But the
$s_0$ terms cancel out of the second update, and once they're gone, they're gone for good. The first
gradient still got us to $\theta_1$, of course; we just stop adding a contribution from it in later
updates. And the second update is $(1+\mu)s_1$, so the "long" gradient step has turned up again, one
iteration later.

#### Relative Weights

Here's a table showing how, on the fourth update from the initial point, the
methods weight the current and previous gradient steps:

| Momentum Type              |  $s_3$     | $s_2$   | $s_1$   | $s_0$   |
:---------------------------:|:----------:|:-------:|:-------:|:--------:
Classical                    | 1          | $\mu$   | $\mu^2$ | $\mu^3$
NAG, fixed $\mu$ from step 0 | 1 + $\mu$  | $\mu^2$ | $\mu^3$ | $\mu^4$
NAG, $\mu_0=0$ startup       | 1 + $\mu$  | $\mu^2$ | $\mu^3$ | 0

With fixed $\mu$, NAG gives the current gradient step more weight and multiplies the coefficients of
all the older steps by an extra factor of $\mu$. After that, the weights decay at the same rate for
both methods: each older step gets another factor of $\mu$.

There's a nice bit of bookkeeping here too. For $0\leq\mu<1$, if we keep adding older terms, the NAG
weights sum to:

$$
1 + \mu + \frac{\mu^2}{1-\mu} = \frac{1}{1-\mu}.
$$

That's the same sum as classical momentum. NAG has shifted some weight from the older steps to the
current one, while keeping the total the same.

Dividing the newest coefficient by this sum gives a long-history weight of $1-\mu$ for classical
momentum and $(1+\mu)(1-\mu)=1-\mu^2$ for NAG. At $\mu=0.9$, that's 10% versus 19%. The four-step
examples below have much less history to share the weight with, so their percentages are larger.

The regularity of the difference between classical momentum and NAG weights
suggest that there should be a way to express the NAG update in terms of a
classical momentum update. See the section on 'Dozat Nesterov momentum' below
for more on that.

Let's write some R code to generate a table to compare how the relative weights
turn out between CM and NAG at different values of $\mu$. We'll normalize the
coefficients so they sum to 1.

```{r generate relative weights}
cm_weights <- function(mu) { c(1, mu, mu * mu, mu * mu * mu) }

nag_zero_start_weights <- function(mu) { c(1 + mu, mu * mu, mu * mu * mu, 0) }

nag_weights <- function(mu) { c(1 + mu, mu * mu, mu * mu * mu, mu * mu * mu * mu) }


mumat <- function(wfun, mus = c(0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) {
  rel_weights <- matrix(nrow = length(mus), ncol = 5)
  for (i in 1:length(mus)) {
    mu <- mus[i]
    weights <- wfun(mu)
    rel_weights[i, ] <- c(mu, weights / sum(weights))
  }
  colnames(rel_weights) <- c("mu", "s3", "s2", "s1", "s0")

  rel_weights
}
```

#### Classical momentum weights

```{r table of classical momentum relative weights}
knitr::kable(mumat(cm_weights), digits = 4)
```

#### NAG weights

```{r table of NAG relative weights}
knitr::kable(mumat(nag_weights), digits = 4)
```

#### NAG weights with $\mu_0=0$ startup

```{r table of zero-start NAG relative weights}
knitr::kable(mumat(nag_zero_start_weights), digits = 4)
```

For the four steps shown here, at high momentum CM puts around 25--30% of the weight on the current
step, compared with 41--46% for NAG and 50--55% for NAG with the short first step. That's quite a
shift in emphasis, even after just four iterations.

### Unrolling with a momentum schedule

I kept $\mu$ constant in the unrolling expressions so far to keep things neat
and to highlight some similarities between CM and NAG. But what if we want $\mu$
to vary with iteration? In our derivations of Nesterov momentum schemes we have
explicitly accounted for $\mu_t$ so let's take a look (and brace ourselves for
the expressions to lose a lot of readability):

#### Classical Momentum with a Schedule

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + s_0 \\
\\
\theta_2 & = \theta_1 + s_1 + \mu_1 v_1 \\
 & = \theta_1 + s_1 + \mu_1 s_0 \\
\\
\theta_3 & = \theta_2 + s_2 + \mu_2 v_2 \\
 & = \theta_2 + s_2 + \mu_2 \left( s_1 + \mu_1 s_0 \right) \\
 & = \theta_2 + s_2 + \mu_2 s_1 + \mu_2 \mu_1 s_0 \\
\\
\theta_4 & = \theta_3 + s_3 + \mu_3 v_3 \\
 & = \theta_3 + s_3 + \mu_3 \left( s_2 + \mu_2 s_1 + \mu_2 \mu_1 s_0 \right)\\
 & = \theta_3 + s_3 + \mu_3 s_2 + \mu_3 \mu_2 s_1 + \mu_3 \mu_2 \mu_1 s_0 \\
\end{split}
\end{equation}
$$
Already this is looking pretty gross.

#### NAG with a Momentum Schedule

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + \left( 1 + \mu_0 \right) s_0 \\
\\
\theta_2 & = \theta_1 + \left( 1 + \mu_1 \right) s_1 - \mu_1 s_0 + \mu_1 v_1 \\
 & = \theta_1 + \left(1 + \mu_1 \right) s_1 - \mu_1 s_0 + \mu_1 \left( 1 + \mu_0 \right) s_0 \\
 & = \theta_1 + \left(1 + \mu_1 \right) s_1 + \mu_1 \mu_0 s_0 \\
\\
\theta_3 & = \theta_2 + \left( 1 + \mu_2 \right) s_2 - \mu_2 s_1 + \mu_2 v_2 \\
& = \theta_2 + \left(1 + \mu_2 \right) s_2 - \mu_2 s_1 + \mu_2 \left(1 + \mu_1 \right) s_1 + \mu_2 \mu_1 \mu_0 s_0\\
& = \theta_2 + \left(1 + \mu_2 \right) s_2 + \mu_2 \mu_1 s_1 + \mu_2 \mu_1 \mu_0 s_0 \\
\\
\theta_4 & = \theta_3 + \left(1 + \mu_3 \right) s_3 - \mu_3 s_2 + \mu_3 v_3 \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 - \mu_3 s_2 + \mu_3 \left[ \left(1 + \mu_2 \right) s_2 + \mu_2 \mu_1 s_1 + \mu_2 \mu_1 \mu_0 s_0 \right] \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 - \mu_3 s_2 + \mu_3 \left(1 + \mu_2 \right) s_2 + \mu_3 \mu_2 \mu_1 s_1 + \mu_3 \mu_2 \mu_1 \mu_0 s_0 \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 + \mu_3 \mu_2 s_2 + \mu_3 \mu_2 \mu_1 s_1 + \mu_3 \mu_2 \mu_1 \mu_0 s_0
\end{split}
\end{equation}
$$

I can see why I was in no hurry to write all that out in full until now. With a
$\mu_0=0$ startup, the $s_0$ term vanishes after the first returned point, just
as in the constant-$\mu$ calculation above.

Here's a comparison of the weights applied to the different gradient steps at
$\theta_4$:

| Momentum Type |  $s_3$       | $s_2$         | $s_1$               | $s_0$                   |
:--------------:|:-------------|:--------------|:--------------------|:-------------------------
Classical       | 1            | $\mu_3$       | $\mu_3 \mu_2$       | $\mu_3 \mu_2 \mu_1$
NAG             | 1 + $\mu_3$  | $\mu_3 \mu_2$ | $\mu_3 \mu_2 \mu_1$ | $\mu_3 \mu_2 \mu_1 \mu_0$

Now the momentum is not constant at each iteration, we see a more complex
relationship between the CM and NAG weights. There definitely doesn't seem to
be a simple re-weighting that could map from CM to NAG here: each gradient
needs re-weighting by a different $\mu_t$.

## Dozat Nesterov Momentum

A paper by [Dozat integrating Nesterov
momentum](https://openreview.net/forum?id=OM0jvwB8jIp57ZJjtNEZ) into the
[Adam](https://arxiv.org/abs/1412.6980) gives a slightly different expression,
that is quite closely related to the Bengio form:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \phi_{t+1} + \mu_{t}\left(\phi_{t+1} - \phi_{t} \right) \\
 & = \phi_{t+1} + \mu_{t} b_{t+1} \\
 & = \theta_{t} + s_{t} + \mu_{t} b_{t+1}
\end{split}
\end{equation}
$$
Now let us express $b_{t+1}$ recursively in terms of the previous $b_{t}$:

$$
\begin{equation}
\begin{split}
b_{t+1} & = \phi_{t+1} - \phi_{t} \\
& = \left(\theta_{t} + s_{t} \right) - \phi_{t} \\
& = \phi_{t} + \mu_{t-1} \left(\phi_{t} - \phi_{t-1} \right) + s_{t} - \phi_{t} \\
& = \mu_{t-1} b_{t} + s_{t}
\end{split}
\end{equation}
$$
This looks just like the classical momentum update, with $b_t$ instead of $m_t$ or $v_t$ and the
momentum coefficient shifted back one iteration. Let's put the two forms next to each other.

Classical momentum looks like:

$$
\begin{equation}
\begin{split}
m_{t+1} = \mu_t m_{t} + s_{t} \\
\theta_{t+1} = \theta_{t} + m_{t+1}
\end{split}
\end{equation}
$$
and NAG looks like:

$$
\begin{equation}
\begin{split}
b_{t+1} = \mu_{t-1} b_{t} + s_{t} \\
\theta_{t+1} = \theta_{t} + s_{t} + \mu_{t} b_{t+1}
\end{split}
\end{equation}
$$

So a very similar recursive form for the momentum buffer for both CM and NAG, but using momentum
from a different time step, and then a different velocity vector for the update. The Dozat paper
shifts the momentum indices by one time step, so remember to shift the schedule too when translating
between the two. Constant momentum saves us some trouble here again.

### Relation to QHM

This way of combining a momentum buffer with the current gradient step also turns up in
[quasi-hyperbolic momentum (QHM)](https://arxiv.org/abs/1810.06801), introduced by Ma and Yarats in
2018. I wrote some extra [notes on QHM](https://jlmelville.github.io/mize/articles/qhm.html) for my
own benefit, but we can make the connection here with the expressions we already have.

With fixed momentum, the buffer $b_t$ above follows the classical momentum recurrence:

$$
b_{t+1} = \mu b_t + s_t.
$$

Now recall [unified momentum](#unified-momentum). Keeping $\mu$, $\varepsilon$ and $\lambda$ fixed,
with $0\leq\mu<1$, its update is:

$$
v_{t+1} = \mu v_t + s_t + \lambda\mu(s_t - s_{t-1})
$$

Starting with $v_0=b_0=s_{-1}=0$, a bit of substitution gives the following expression, which you
can check by induction:

$$
\boxed{
v_{t+1} = \lambda s_t
+ \left[1 - \lambda(1 - \mu)\right]b_{t+1}.
}
$$

Setting $\lambda=1$ gives $v_{t+1}=s_t+\mu b_{t+1}$, exactly the Dozat update above. Setting
$\lambda=0$ gives $v_{t+1}=b_{t+1}$, the classical momentum update.

QHM normalizes its momentum buffer by a factor of $1-\mu$, which means we also need to rescale the
learning rate. Translating into the QHM paper's symbols gives:

$$
\beta_{\mathrm{QHM}} = \mu, \qquad
\nu_{\mathrm{QHM}} = 1 - \lambda(1 - \mu), \qquad
\alpha_{\mathrm{QHM}} = \frac{\varepsilon}{1 - \mu}.
$$

With $\lambda=0$, we get $\nu_{\mathrm{QHM}}=1$, which is classical momentum. With $\lambda=1$, we
get $\nu_{\mathrm{QHM}}=\mu$, which is NAG. So letting $\lambda$ run from 0 to 1 takes us along the
part of QHM between classical momentum and NAG.

### Where the learning rate lives

There's another implementation detail that changes the weighting of history: does an old gradient
keep the learning rate it had when we calculated it, or does today's learning rate rescale its
contribution? To isolate that question, keep $\mu$ fixed and write $g_t=\nabla f(\theta_t)$.

Our buffer stores gradient steps, including their learning rates:

$$
\begin{aligned}
b_{t+1} &= \mu b_t-\varepsilon_t g_t, \\
\theta_{t+1} &= \theta_t-\varepsilon_t g_t+\mu b_{t+1}.
\end{aligned}
$$

Alternatively, we can store the gradients in a buffer $B_t$, and apply the learning rate when we
update the parameters:

$$
\begin{aligned}
B_{t+1} &= \mu B_t+g_t, \\
\theta_{t+1} &= \theta_t-\varepsilon_t(g_t+\mu B_{t+1}).
\end{aligned}
$$

With a constant learning rate and both buffers starting at zero, $b_t=-\varepsilon B_t$ makes the
updates equivalent. With a changing learning rate, the first version preserves the old learning
rates inside its accumulated steps, while the second applies the current rate to the whole gradient
history. They can therefore give different results under the same learning-rate schedule.

These are both used in practice: [Keras](https://keras.io/api/optimizers/sgd/) documents the first
convention, while [PyTorch](https://docs.pytorch.org/docs/stable/generated/torch.optim.SGD.html)
documents the second and notes the difference for Nesterov momentum too. We're comparing just the
momentum updates here, with no weight decay or dampening.

## NAG in practice

So that's how it's all supposed to work in principle. Here I'll demonstrate
the equivalence of these methods, by implementing them all in simple R code.

The biggest simplification I'll make is that I'll assume a constant learning
rate and a constant momentum coefficient.

### Classical Momentum

```{r Classical mometum}
#' Optimization by Classical Momentum
#'
#' @param par Starting point of vector of parameters to optimize.
#' @param fn Objective function to optimize. Takes vector with length of
#' \code{par} and returns a scalar.
#' @param gr Gradient of the objective function \code{fn}. Takes vector with
#' length of \code{par} and returns a vector of the same length.
#' @param lr Learning rate.
#' @param mu Momentum coefficient.
#' @param max_iter Maximum number of iterations to optimize for. First iteration
#' is always steepest descent.
#' @return list with components: \code{par} final set of parameters; \code{f}
#' value of \code{fn} evaluated at the returned set of parameters; \code{fs}
#' vector of function evaluated after each iteration.
cm <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- mu * v - lr * g
    par <- par + v

    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

This is a reference implementation of classical momentum. Nearly all the
parameters and the return values are the same for the other functions (except
where noted), so they're documented here once.

Onto the various NAG implementations. The `nag` routine below sets the momentum coefficient to zero
on the first iteration, giving us the "short" first step. The Bengio, Dozat and momentum-style
versions use the nonzero value straight away. Sutskever gives us its result after the gradient
stage. So we already have a few differences to keep an eye on. Classical momentum gets its initial
gradient step simply by starting the velocity vector at zero.

### NAG

```{r NAG}
# Optimization by Nesterov Accelerated Gradient
#
# Return list also contains gd_fs: function evaluated after gradient descent
# stage of each iteration; all: function evaluated after gradient descent
# stage and momentum stage, in order.
nag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)
  gd_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  x_old <- rep(0, length(par))
  for (i in 1:max_iter) {
    # gradient descent stage
    g <- gr(par)
    x <- par - (lr * g)

    # store gradient descent values
    f <- fn(x)
    gd_fs[i] <- f
    all[i * 2 - 1] <- f

    # momentum stage and update
    par <- x + ifelse(i == 1, 0, mu) * (x - x_old)
    x_old <- x

    # store momentum values
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, gd_fs = gd_fs, all = all)
}
```

In this routine, rather than store the velocity vector, we store the
previous gradient descent result, `x_old`. In order to ensure the
first iteration is gradient descent only, we also need to manually
set the momentum coefficient `mu` to zero on the first iteration,
which is what that `ifelse` expression does.

Also, there's some extra code to calculate and store the function values
after the gradient descent stage. These aren't needed for the optimization
to work, I just want to keep track of the values to demonstrate the
different methods are in fact equivalent.

### Sutskever Formulation

```{r Sutskever}
# Optimization by Sutskever Nesterov Momentum
#
# Return list also contains mu_fs: function evaluated after momentum
# stage of each iteration; all: function evaluated after gradient descent
# stage and momentum stage, in order.
snag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  v <- rep(0, length(par))

  fs <- rep(0, max_iter)
  mu_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  for (i in 1:max_iter) {
    # momentum stage and update parameters
    mu_step <- mu * v
    par <- par + mu_step

    # store momentum results
    f <- fn(par)
    mu_fs[i] <- f
    all[i * 2 - 1] <- f

    # gradient descent stage
    g <- gr(par)
    gd_step <- -lr * g

    # update and store velocity for next step
    par <- par + gd_step
    v <- mu_step + gd_step

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, mu_fs = mu_fs, all = all)
}
```

This one is pretty straight-forward.

### Bengio Formulation

```{r Bengio}
# Optimization by Bengio Nesterov Momentum
bnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    # here mu_{t-1} * mu_t equals mu^2 because mu is constant
    par <- par + mu * mu * v - (1 + mu) * lr * g
    v <- mu * v - lr * g

    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

Because this implementation uses constant momentum, we can replace $\mu_{t-1}\mu_t$ with $\mu^2$.
Easy enough, until we decide to start momentum at zero: on the next iteration, the product is still
zero, while $\mu^2$ isn't. That's the difference between some momentum and no momentum. We shall
return to this point later.

### Dozat Nesterov Momentum

```{r Dozat NAG}
# Optimization by Dozat's expression for Nesterov momentum
dnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  # mprime, the recursive variable
  mpt <- rep(0, length(par))
  mumu <- mu * mu

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)

    s <- lr * -g
    v <- (1 + mu) * s + mumu * mpt
    par <- par + v

    # update mprime for next step
    mpt <- mu * mpt + s
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

*18 December 2021*: Dozat's version of Nesterov momentum. We should expect this
to give the same result as NAG and Bengio's Nesterov momentum.

### Alternative momentum NAG Expression

```{r Alternative NAG}
# Optimization by Nesterov Accelerated Gradient, using a momentum-style
# update expression.
mnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  r <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    s <- -lr * g
    r <- mu * (r + s)
    par <- par + s + r

    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

Finally, here is your humble author's expression for NAG, written as a momentum-style update.
This is unified momentum with $\lambda=1$. To differentiate from the Sutskever and Bengio versions
of NAG, I'll refer to it as momentum-NAG or mNAG.

### Testing with Rosenbrock

Tradition dictates that I must demonstrate the use of these optimizers using the
2D Rosenbrock function, with a specific starting point:

```{r test function}
par <- c(-1.2, 1)

fn <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  100 * (x2 - x1 * x1) ^ 2 + (1 - x1) ^ 2
}
gr <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  c(
    -400 * x1 * (x2 - x1 * x1) - 2 * (1 - x1),
    200 *      (x2 - x1 * x1))
}
```

Let's run the optimizers for 100 iterations. The point here is not whether we
get really amazing optimization (we don't), but whether the outputs of the
Sutskever and Bengio algorithms are equivalent. It would be a bonus if my mNAG
result also worked. For comparison we'll throw in the classical momentum and as
a sanity check, the vanilla NAG routine.

```{r generating output}
lr <- 0.001
mu <- 0.95
max_iter <- 100
snag_opt <- snag(par, fn, gr, lr, mu, max_iter)
bnag_opt <- bnag(par, fn, gr, lr, mu, max_iter)
nag_opt <- nag(par, fn, gr, lr, mu, max_iter)
mnag_opt <- mnag(par, fn, gr, lr, mu, max_iter)
dnag_opt <- dnag(par, fn, gr, lr, mu, max_iter)

cm_opt <- cm(par, fn, gr, lr, mu, max_iter)

sbnag_df <- data.frame(Bengio = bnag_opt$fs,
                 Sutskever = snag_opt$fs,
                 "Suts Mom" = snag_opt$mu_fs,
                 "NAG" = nag_opt$fs,
                 "mNAG" = mnag_opt$fs,
                 "Dozat" = dnag_opt$fs,
                 "CM" = cm_opt$fs)
```

Let's have a look at the first few iteration results:

```{r table head}
knitr::kable(head(sbnag_df), caption = paste(
  "First few evaluations of NAG implementations, with lr = ",
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

The first two columns pit the Sutskever vs Bengio formulations directly. And, as
we would expect, they're not the same: the Sutskever iteration result is from
the gradient descent stage, and the Bengio result comes from the momentum stage.
But if we put the momentum results from the Sutskever formulation up, we can see
that they are the same as the Bengio result, but behind by one iteration, i.e.
the Bengio result at iteration $t$ matches the Sutskever momentum stage result
at $t+1$. The mNAG column uses the alternative derivation above and matches the Bengio results.
Hurrah.

*18 December 2021* The results for Dozat's version of Nesterov momentum have
also been added. As expected, it matches the Bengio and mNAG results. Not much
else to say about it so it won't appear in any further discussion.

However, none of these results match the vanilla `nag` implementation. A clue to what's going on is
in the first row: NAG and CM have the same result, because setting $\mu_0=0$ in `nag` gives us the
same initial gradient step. Sutskever also stops after gradient descent on its first iteration. The
Bengio, Dozat and mNAG routines take the "long" first step, and even Sutskever parts company with
`nag` on the next iteration.

We'll get back to this, but let's just make sure these observations hold up at
the end of the table too.

```{r table tail}
knitr::kable(tail(sbnag_df), caption = paste(
  "Last few evaluations of NAG implementations, with lr = ",
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

The Sutskever, Bengio and momentum NAG implementations all still match up in the same way they did.
CM and vanilla NAG are off doing their own thing.

### Behavior on first iteration

The reason for the differing behavior is down to two things:

* The `nag` implementation sets its initial momentum coefficient to zero. Bengio, Dozat and mNAG use
  the nonzero value from the start.
* Sutskever ends each iteration after gradient descent, while the others end after the momentum
  stage.

For the first point, the Bengio, Dozat and mNAG routines give us $\theta_0+(1+\mu)s_0$, even with
their momentum buffers starting at zero. This is the "long" first step we met when unrolling the
updates.

What about the Sutskever formulation? Let's think about the chain of parameter
updates that actually take place over the first few iterations:

For `nag`, the chain is: gradient descent stage, momentum stage, gradient descent stage, momentum
stage. Except the first momentum stage does nothing, because we've set its coefficient to zero. So
the first three iterations look like `g|gm|gm`, with the bar marking the end of an iteration.

For Sutskever, the chain is: momentum stage, gradient descent stage, momentum stage, gradient
descent stage. Its first momentum stage also does nothing, this time because the buffer is zero.
That gives us `g|mg|mg`.

There it is: our `nag` routine starts with two gradient steps in a row, while Sutskever puts a
momentum step between them. And then we ask for their results at different stages. No wonder the
numbers don't agree.

## Getting the NAG implementations to agree

We've spent quite a lot of effort showing that these expressions are equivalent, so it would be
nice to get the same numbers out of them. Let's give them all the same start as our `nag` routine,
with $\mu_0=0$, and compare the results at matching stages.

### Sutskever formulation with matched startup

```{r Sutskever with matched startup}
# Sutskever Nesterov momentum with matched startup
#
# Extra parameter wait: wait this number of extra iterations before applying
# momentum. Needed only to sync up with other implementations of Nesterov
# momentum: set wait to 1 to make mu_f at iter i match the output of the
# other implementations at iter i-1.
snagc <- function(par, fn, gr, lr, mu, max_iter = 10, wait = 0) {
  v <- rep(0, length(par))

  fs <- rep(0, max_iter)
  mu_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  for (i in 1:max_iter) {
    # momentum stage and update parameters
    mu_step <- ifelse(i > wait + 1, mu, 0) * v
    par <- par + mu_step

    # store momentum results
    f <- fn(par)
    mu_fs[i] <- f
    all[i * 2 - 1] <- f

    # gradient descent stage
    g <- gr(par)
    gd_step <- -lr * g

    # update and store velocity for next step
    par <- par + gd_step
    v <- mu_step + gd_step

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, mu_fs = mu_fs, all = all)
}
```

The Sutskever formulation already does gradient descent on its first iteration, so there's not a
huge change required. I've introduced a new parameter, `wait`, that controls how many extra
iterations to wait before applying momentum. Set it to `0` and you get the current behavior. Set it
to `1` and the momentum step will be zero on the second iteration too. That gives us the two
gradient descent stages in a row that should sync us up with `nag`.

### Bengio formulation with matched startup

```{r Bengio with matched startup}
# Bengio Nesterov momentum with matched startup
bnagc <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    if (i == 1) {
      par <- par - lr * g
    }
    else {
      par <- par + (mu * mu * v) - ((1 + mu) * lr * g)
      v <- (mu * v) - (lr * g)
    }

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

Getting the Bengio formulation to agree takes a bit more work than you might think. With $\mu_0=0$,
the product $\mu_0\mu_1$ is zero on the second iteration, so our convenient replacement by $\mu^2$
breaks down. We can't just set `mu` to zero for that whole iteration either, because we still need
the nonzero value in $1+\mu_1$.

We could start storing the previous momentum coefficient, even though I chose constant momentum to
keep things simple. Or we can leave the buffer at zero during the first iteration, which gets us
the same next update. I chose the latter.

### Momentum NAG with matched startup

```{r Momentum NAG with matched startup}
# momentum NAG with matched startup
mnagc <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  r <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    s <- -lr * g
    r <- ifelse(i == 1, 0, mu) * (r + s)
    par <- par + s + r

    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

And last, the rewritten momentum NAG update. This required the least
modification from the original routine: simply set the momentum coefficient to
zero on the first iteration.

### Results after matching the first steps

Time to look at some numbers again. We'll compare the new versions of the
Sutskever, Bengio and Momentum formulations of NAG with vanilla NAG. We'll also
once again pull out the momentum stage results for Sutskever so we can compare
directly to the Bengio result.

```{r optimization forcing gradient descent initially}
mnagc_opt <- mnagc(par, fn, gr, lr, mu, max_iter)
snagc_opt <- snagc(par, fn, gr, lr, mu, max_iter, wait = 1)
bnagc_opt <- bnagc(par, fn, gr, lr, mu, max_iter)

ncdf <- data.frame(cBengio = bnagc_opt$fs,
                 cSutskever = snagc_opt$fs,
                 "Suts Mom" = snagc_opt$mu_fs,
                 NAG = nag_opt$fs,
                 cmNAG = mnagc_opt$fs,
                 "NAG gd" = nag_opt$gd_fs)
```

Let's have a look at the first few iteration results:

```{r table of initial results}
knitr::kable(head(ncdf), caption = paste(
  "First few evaluations of matched-state NAG implementations, with lr = ",
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

These are in the same order as the previous table. The Bengio and Sutskever columns still give us
results from different stages. But the Sutskever momentum result in the third column *does* match
the Bengio result from the previous iteration. And now the NAG result in the fourth column matches
Bengio and mNAG too. Finally, as an extra check, the NAG gradient descent result in the last column
matches Sutskever. Everything is lining up at last.

And let's take a look at the final few iterations, just to make sure everything
still holds up:

```{r table of final results}
knitr::kable(tail(ncdf), caption = paste(
  "Last few evaluations of matched-state NAG implementations, with lr = ",
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

That's a relief.

```{r check-coordinate-trajectories, include=FALSE}
# Check the demonstrated functions at every coordinate, including the final one.
# Capturing gradient evaluation points also lets us align Sutskever's stages.
trace_gradients <- function(method, iterations = max_iter, ...) {
  points <- vector("list", iterations)
  n <- 0L
  traced_gr <- function(x) {
    n <<- n + 1L
    points[[n]] <<- x
    gr(x)
  }
  result <- method(par, fn, traced_gr, lr, mu, iterations, ...)
  stopifnot(n == iterations)
  list(points = do.call(rbind, points), final = result$par)
}

same_coordinates <- function(x, y) {
  stopifnot(isTRUE(all.equal(x, y, tolerance = 1e-10)))
}

long_trace <- trace_gradients(bnag)
long_points <- rbind(long_trace$points, long_trace$final)
for (method in list(dnag, mnag)) {
  traced <- trace_gradients(method)
  same_coordinates(rbind(traced$points, traced$final), long_points)
}
same_coordinates(trace_gradients(snag, max_iter + 1L)$points, long_points)

short_trace <- trace_gradients(nag)
short_points <- rbind(short_trace$points, short_trace$final)
for (method in list(bnagc, mnagc)) {
  traced <- trace_gradients(method)
  same_coordinates(rbind(traced$points, traced$final), short_points)
}
same_coordinates(
  trace_gradients(snagc, max_iter + 1L, wait = 1)$points,
  short_points
)

# Exercise the scheduled equations separately from the constant-coefficient
# demonstrations. The two-stage update supplies the reference trajectory.
scheduled_trajectory <- function(form, rates, momenta) {
  stopifnot(length(rates) == length(momenta))
  x <- phi <- par
  v <- previous_step <- b <- r <- B <- rep(0, length(par))
  previous_mu <- 0
  points <- matrix(NA_real_, nrow = length(rates) + 1L, ncol = length(par))
  points[1, ] <- x
  for (i in seq_along(rates)) {
    g <- gr(x)
    s <- -rates[i] * g
    momentum <- momenta[i]
    if (form == "two-stage") {
      next_phi <- x + s
      x <- next_phi + momentum * (next_phi - phi)
      phi <- next_phi
    } else if (form == "displacement") {
      v <- momentum * v + s + momentum * (s - previous_step)
      x <- x + v
    } else if (form == "Bengio") {
      b <- previous_mu * b + s
      x <- x + s + momentum * b
    } else if (form == "saved-gap") {
      r <- momentum * (r + s)
      x <- x + s + r
    } else if (form == "gradient-buffer") {
      # This comparison is used only with fixed momentum below.
      B <- momentum * B + g
      x <- x - rates[i] * (g + momentum * B)
    } else {
      stop("Unknown formulation")
    }
    previous_step <- s
    previous_mu <- momentum
    points[i + 1L, ] <- x
  }
  points
}

rates <- rep(c(0.001, 0.0003, 0.0008, 0.0005), 6)
momenta <- rep(c(0.7, 0.4, 0, 0.9, 0.2, 0.6), 4)
for (initial_mu in c(0, 0.7)) {
  momenta[1] <- initial_mu
  reference <- scheduled_trajectory("two-stage", rates, momenta)
  for (form in c("displacement", "Bengio", "saved-gap")) {
    same_coordinates(scheduled_trajectory(form, rates, momenta), reference)
  }
}

# The learning-rate conventions agree at a constant rate but can differ under
# a changing rate. Hold momentum and initialization fixed in both comparisons.
fixed_momenta <- rep(0.8, length(rates))
fixed_rates <- rep(0.001, length(rates))
same_coordinates(
  scheduled_trajectory("Bengio", fixed_rates, fixed_momenta),
  scheduled_trajectory("gradient-buffer", fixed_rates, fixed_momenta)
)
stopifnot(max(abs(
  scheduled_trajectory("Bengio", rates, fixed_momenta) -
    scheduled_trajectory("gradient-buffer", rates, fixed_momenta)
)) > 1e-5)
```

## Conclusions

Yes, NAG can be thought of as being like classical momentum where you do the momentum step first and
*then* the gradient step. We can also look at it as a weighted history of gradient steps: with fixed
momentum, NAG shifts some weight from the older steps to the current one. The weights still decay at
the same rate as classical momentum, and their total over an infinite history is the same.

On a quadratic, we can see exactly what that change does: the momentum coefficient depends on the
curvature. Flat directions keep their momentum, while increasing positive curvature reduces it
towards zero and eventually reverses
it. Personally, I find this view through the gradient history more appealing than the Sutskever
formulation, especially now that it leads us to unified momentum and QHM.

In terms of implementation, I still pity anyone tasked with implementing Nesterov momentum and
demonstrating that they actually got it right. The same objective, starting point, learning rate and
momentum can give you different outputs from each version, as we managed quite successfully above.
Getting them to agree means lining up the first momentum coefficients, the stored vectors and the
stage where we take the result. With a changing learning rate, we also need to check whether the
buffer stores gradients or scaled steps. Having done that, we can finally get the numbers to
confirm the algebra.
