---
title: "Nesterov Accelerated Gradient and Momentum"
author: "James Melville"
date: "December 31, 2016"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

A way to express Nesterov Accelerated Gradient in terms of a regular momentum 
update was noted by 
[Sutskever and co-workers](http://www.jmlr.org/proceedings/papers/v28/sutskever13.html),
and perhaps more importantly, when it came to training neural networks, it 
seemed to work better than classical momentum schemes. This was further
confirmed by [Bengio and co-workers](https://arxiv.org/abs/1212.0901), who
provided an alternative formulation that might be easier to integrate into
existing software.

I implemented NAG as part of [mize](https://github.com/jlmelville/mize), but
found it was a bit more difficult than I'd anticipated to test that I'd got the
correct results. 

It seems I am not alone in being confused by the exact details of implementing 
NAG as a momentum scheme (e.g. comments and issues in the deep learning projects [keras](https://github.com/fchollet/keras/issues/966) 
and [mocha.jl](https://github.com/pluskid/Mocha.jl/pull/47), which I found
from random googling), so mainly for the benefit of future me, I am
going to derive the Sutskever and Bengio formulations from NAG in tedious 
detail. I am also going to derive an alternative expression that I ended up
using in mize. Finally, I will write some R code to demonstrate their
equivalence. This will turn out to be trickier than I anticipated.

Wherever possible I will try and stick with the notation used by Sutskever.
Apart from the Sutskever and Bengio papers linked to above, it's worth checking
out the 
[appendix to the Sutskever paper (PDF)](http://www.jmlr.org/proceedings/papers/v28/sutskever13-supp.pdf) or the 
relevant part (chapter 7) of 
[Sutskever's thesis](http://hdl.handle.net/1807/36012). 
For an introduction to NAG itself, try the first part of this paper by 
[O'Donoghue and Candès](https://arxiv.org/abs/1204.3982) (the rest of it is
also good but not germane to Nesterov momentum expressions).

## Definitions

The goal is to optimize some parameters, $\theta$. This is going to involve
gradient descent, so we will be evaluating the gradient of an objective 
function of those parameters, $\nabla f\left(\theta\right)$, and moving a 
certain distance in the direction of the negative of the gradient, the distance
being related to the learning rate, $\varepsilon$. There will also be a 
momentum term involved, with momentum coefficient $\mu$. The value of the 
parameters, learning rate and momentum at iteration $t$ will be indicated by a 
subscript, e.g. $\theta_t$.

A definition which holds across all the methods discussed here is that the
parameters at iteration $t+1$ are related to the parameters at iteration $t$
by an update that involves the addition of a velocity vector, $v$:

$$\theta_{t+1} = \theta_t + v_{t+1} \implies v_{t+1} = \theta_{t+1} - \theta_t$$

This is pretty obvious, but different methods have different velocity 
defintions, and it's easy to get confused.

Finally, I'll try to be strict about names. When I'm referring to the original
Nesterov Accelerated Gradient, I'll call that NAG. When I'm referring to the
versions that recast it as a momentum scheme, I'll call that Nesterov momentum,
and refer to either the Sutskever or Bengio forumulation where necessary.
The traditional momentum scheme is referred to as "standard" or "regular" 
momentum by Bengio, and "classical" by Sutskever. I'll refer to it as 
"classical" momentum.

## Classical Momentum

The classical momentum velocity vector is defined by Bengio as:

$$v_t = \mu_{t-1}v_{t-1} - \varepsilon_{t-1}\nabla f\left(\theta_{t-1}\right)$$

Sutskever gives the same definition but without any $t$ subscript on the
velocity vector or the momentum coefficient.

We can write out the full update for classical momentum as:

$$\theta_{t+1} = \theta_t + \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t\right)$$
with velocity:

$$v_{t+1} = \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t \right)$$

## NAG

The Nesterov Accelerated Gradient method consists of a gradient descent step,
followed by something that looks a lot like a momentum term, but isn't exactly
the same as that found in classical momentum. I'll call it a "momentum stage"
here. It's important to note that the parameters being minimized by NAG are
given the symbol $y$ by Sutskever, not $\theta$. You'll see that $\theta$ is the
symbol for the parameters after they've been updated by the gradient descent
stage, but before the momentum stage.

Here's the gradient descent stage:
$$\theta_t = y_t - \varepsilon_t \nabla f\left(y_t\right)$$
And here's the momentum stage:
$$y_{t+1} = \theta_t + \mu_t\left(\theta_t - \theta_{t-1} \right)$$
That concludes one iteration of NAG. The hard part is actually finding the 
correct learning rate and momentum value in order to get the convergence
guarantees that make the method attractive, but that needn't concern us. 
Sutskever in his thesis suggests manual tuning to get an optimal result (at 
least for deep learning applications) so we are in heuristic territory here 
anyway.

## Sutskever Nesterov Momentum

The key idea behind the Sutskever momentum derivation is to shift the 
perspective about which of the parameters we want as the result of the 
iteration, from $y$ to $\theta$. Rather than having the optimization iterations 
proceed as "gradient descent, momentum (end of iteration 1), gradient descent, 
momentum (end of iteration 2), gradient descent etc." move the boundary of where
the iterations end by a half-iteration to get "momentum, gradient descent (end
of iteration 1), momentum, gradient descent (end of iteration 2) etc.". This
leaves a phantom gradient descent step that used to be first stage of the
first iteration, now floating in the nether regions of iteration zero, but you
can just pretend that the starting position is the result of gradient descent
from some other arbitrary starting position.

The Sutskever derivation then proceeds as follows. First, rewrite the
momentum stage in terms of $y_t$:

$$y_t = \theta_{t-1} + \mu_{t-1}\left(\theta_{t-1} - \theta_{t-2} \right)$$

Then note that the term in parentheses is just the definition of the velocity
vector, so we can write:

$$y_t = \theta_{t-1} + \mu_{t-1}v_{t-1}$$

We now substitute this expression for $y_t$ into the gradient descent stage:

$$\theta_t = \theta_{t-1} + \mu_{t-1}v_{t-1} - \varepsilon_t \nabla f\left(y_t\right)$$
and replace $y_t$ with an expression in terms of $\theta_t$:

$$\theta_t = \theta_{t-1} + \mu_{t-1}v_{t-1} - \varepsilon_t \nabla f\left(\theta_{t-1} + \mu_{t-1}v_{t-1}\right)$$

At this point, the expression differs from that given by Sutskever in one 
detail: the learning rate associated with the gradient descent is currently 
written as $\varepsilon_{t}$. But remember that we have now moved the boundaries
of the iteration: the gradient descent that would have been the first stage
of iteration $t+1$ is now the second stage of iteration $t$. So the above
expression is more correctly written as:

$$\theta_t = \theta_{t-1} + \mu_{t-1}v_{t-1} - \varepsilon_{t-1} \nabla f\left(\theta_{t-1} + \mu_{t-1}v_{t-1}\right)$$

just as given by Sutskever. And therefore the expression for the parameter 
update is trivially:

$$\theta_{t+1} = \theta_t + \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t + \mu_t v_t\right)$$

with the velocity vector defined as:

$$v_{t+1} = \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t + \mu_t v_t\right)$$

This looks just like the classical momentum update, except that the gradient is
calculated after the momentum update. Hence, one can do NAG by simply reversing
the order in which the update is usually carried out: do the momentum stage
first, update the parameters, and then do the gradient descent part.

### Bengio Nesterov Momentum

The Bengio formulation of Nesterov momentum starts from the Sutskever 
definition and then defines a new variable, $\Theta$, which represents 
$\theta$ after the momentum update:

$$\Theta_{t-1} \equiv \theta_{t-1} + \mu_{t-1} v_{t-1}$$

A re-arrangement that will come in handy is:
$$\theta_{t-1} = \Theta_{t-1} - \mu_{t-1} v_{t-1}$$

Also, we are going to do something a bit tricky, and that is to keep on using
the velocity vector definition from the Sutskever formulation, but with 
$\Theta$ substituted in:

$$v_t = \mu_{t-1} v_{t-1} - \varepsilon_{t-1} \nabla f\left(\Theta_{t-1}\right)$$
The reason I call this tricky is that the velocity vector still refers to
the $\theta$ update, but we are going to be writing the update in terms of
$\Theta$. You'll see what I mean.

Anyway, let's take the Sutskever Nesterov momentum update and substitute in
the expression above for $\theta$ in terms of $\Theta$:

$$\Theta_{t+1} - \mu_{t+1} v_{t+1} = \Theta_t - \mu_t v_t + \mu_t v_t - \varepsilon_t \nabla f\left(\Theta_t\right)$$

Noting that those two $\mu_t v_t$ terms cancel out and then rearranging to leave
$\Theta_{t+1}$ on the LHS, we get:

$$\Theta_{t+1} = \Theta_t + \mu_{t+1} v_{t+1} - \varepsilon_t \nabla f\left(\Theta_t\right)$$

We can now substitute in the Sutskever velocity expression, $v_{t+1}$:

$$\Theta_{t+1} = 
\Theta_t + \mu_{t+1}\left[\mu_t v_t - \varepsilon_t \nabla f\left(\Theta_t\right)\right] 
- \varepsilon_t \nabla f\left(\Theta_t\right)
$$
Expanding out the parentheses gives:

$$\Theta_{t+1} = 
\Theta_t + \mu_{t+1} \mu_t v_t - \mu_{t+1} \varepsilon_t \nabla f\left(\Theta_t\right) 
- \varepsilon_t \nabla f\left(\Theta_t\right)
$$
and, after grouping the gradient descent parts, finally gives:

$$\Theta_{t+1} = 
\Theta_t + \mu_{t+1} \mu_t v_t - \left(1 + \mu_{t+1} \right) \varepsilon_t \nabla f\left(\Theta_t\right) 
$$
At this point in the other derivations, I isolated the bits on the RHS that
wasn't the old parameter and defined them as the velocity vector. Well,
we can't do that here, as we're already using the Sutskever definition of
the velocity vector which is $\theta_t - \theta_{t-1}$ and *not*
$\Theta_t - \Theta_{t-1}$. I said it was a bit tricky.

The advantage of this expression for the Nesterov momentum is that it doesn't
require calculating a gradient at a non-standard position, and only requires
a modification to the coefficients used to calculate the velocity, which is
probably an easier change to make to an existing codebase which already uses 
classical momentum.

Before going any further, it's worth going back to the original expression
for the gradient descent part of NAG that Sutskever gives in his thesis (and 
appendix to the paper):

$$y_{t+1} = \theta_t + \mu_t\left(\theta_t - \theta_{t-1} \right) 
= \theta_t + \mu_t v_t
$$
Let's just compare that to the Bengio definition of $\Theta$:

$$\Theta_{t} \equiv \theta_{t} + \mu_{t} v_{t}
$$
Yep, that means:

$$\Theta_{t} \equiv y_{t+1}$$
So the Bengio formulation is really just another way to write the standard NAG update.

### An Alternative Expression for NAG

Let's go back to the original formulation of NAG and now instead of making
$\theta$ the variables after gradient descent, we'll put them where $y$ was 
originally used. The variables after gradient descent I'll refer to as $\phi$.

$$\phi_t = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)$$
$$\theta_{t+1} = \phi_t + \mu_t\left(\phi_t - \phi_{t-1} \right)$$

Now, let's just write out the momentum stage in terms of $\theta$, substituting
$\phi$ wherever we find it:

$$\theta_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
+ \mu_t \left[\theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
- \theta_{t-1} + \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
\right]
$$

Rearranging:

$$\theta_{t+1} = \theta_t 
+ \mu_t \left[\theta_t - \theta_{t-1}
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
- \varepsilon_t \nabla f\left(\theta_t \right)
\right]
- \varepsilon_t \nabla f\left(\theta_t \right)
$$

Finally, we can subtitute in $v_t$ for the first two terms in the square
brackets, to give:

$$\theta_{t+1} = \theta_t 
+ \mu_t \left[v_t
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
- \varepsilon_t \nabla f\left(\theta_t \right)
\right]
- \varepsilon_t \nabla f\left(\theta_t \right)
$$

with velocity:

$$v_{t+1} =  
 \mu_t \left[v_t
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
- \varepsilon_t \nabla f\left(\theta_t \right)
\right]
- \varepsilon_t \nabla f\left(\theta_t \right)
$$

This looks a lot like the classical momentum expression, but with the
velocity vector modified to first remove the contribution of the gradient 
descent from the previous iteration, and replace it with the gradient descent 
contribution from the *current* iteration. Gives an interesting insight into
the idea of the Nesterov momentum using a form of "lookahead" with the
gradient descent.

You could also choose to expand the velocity expression to make it look a bit 
like the Bengio formulation:

$$
v_{t+1} =  
\mu_t \left[v_t
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
\right]
- \left(1 + \mu_t\right) \varepsilon_t \nabla f\left(\theta_t \right)
$$

but as this version can't be expressed as the classical momentum form with
different coefficients, the way the Bengio formulation can be, it probably
doesn't gain you anything, except you can expand it and rearrange it further to
give:

$$
v_{t+1} =  
\mu_t v_t - \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
+
\mu_t 
\left[
\varepsilon_{t-1} \nabla f \left(\theta_{t-1}\right) 
-
\varepsilon_{t} \nabla f \left(\theta_{t}\right)
\right]
$$
which now resembles the classical momentum expression with an extra momentum 
term. I haven't found a definitive reference for this expression, or what, if 
any, extra insight it provides, but user 'denis' uses this expression in an
[answer on the Cross Validated Stack Exchange](https://stats.stackexchange.com/a/233430),
and refers to the third form as "gradient momentum".

We can create a "generalized" momentum update by introducing a parameter
$\beta$ to give:

$$
v_{t+1} =  
\mu_t v_t - \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
+
\beta_t \mu_t 
\left[
\varepsilon_{t-1} \nabla f \left(\theta_{t-1}\right) 
-
\varepsilon_{t} \nabla f \left(\theta_{t}\right)
\right]
$$
where setting $\beta_t = 0$ will give classical momentum and $\beta_t = 1$ gives
Nesterov. The Stack Exchange answer I mentioned above also suggests allowing
the two momentum terms to differ from each other.

From the look of the expression, it seems that a major downside of this 
expression is that calculating the parameters for iteration $t+1$ requires 
storage of information from not only iteration $t$ but from iteration $t-1$ too.
But you don't necessarily have to do any extra storage with an implementation
that used this version of NAG. At the end of an iteration, when saving the
velocity vector for the next iteration, you change:

$$v_{t-1} \leftarrow v_t$$
to:
$$v_{t-1} \leftarrow v_t + \varepsilon_{t} \nabla f\left(\theta_t \right)$$

and then when calculating the momentum term, change:

$$\mu_t v_t$$
to:
$$\mu_t \left[v_t - \varepsilon_{t} \nabla f\left(\theta_t \right)\right]$$

## NAG in practice

So that's how it's all supposed to work in principle. Here I'll demonstrate
the equivalence of these methods, by implementing them all in simple R code.

The biggest simplification I'll make is that I'll assume a constant learning
rate and a constant momentum coefficient.

### Classical Momentum

```{r Classical mometum}
#' Optimization by Classical Momentum
#'
#' @param par Starting point of vector of parameters to optimize.
#' @param fn Objective function to optimize. Takes vector with length of
#' \code{par} and returns a scalar.
#' @param gr Gradient of the objective function \code{fn}. Takes vector with
#' length of \code{par} and returns a vector of the same length.
#' @param lr Learning rate.
#' @param mu Momentum coefficient.
#' @param max_iter Maximum number of iterations to optimize for. First iteration
#' is always steepest descent.
#' @return list with components: \code{par} final set of parameters; \code{f}
#' value of \code{fn} evaluated at the returned set of parameters; \code{fs}
#' vector of function evaluated after each iteration.
cm <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- mu * v - lr * g
    par <- par + v
    
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```
This is a reference implementation of classical momentum. Nearly all the parameters and the
return values are the same for the other functions (except where noted), so they're 
documented here once. 

Onto the various NAG implementations. They all have an extra bit of code in them to ensure
that on the first iteration only the gradient descent stage is carried out. This is
needed to keep the outputs consistent between the different implementations. The `cm`
routine doesn't require any extra checks for steepest descent on the first iteration,
because initializing the velocity vector to zero results in steepest descent anyway
even if the momentum coefficient is non-zero. As we'll see below, some of the other
implementations either don't explicitly use a velocity vector or need the momentum
coefficient to be set to zero on the first iteration to get the same result.

### NAG

```{r NAG}
# Optimization by Nesterov Accelerated Gradient
#
# Return list also contains gd_fs: function evaluated after gradient descent
# stage of each iteration; all: function evaluated after gradient descent
# stage and momentum stage, in order.
nag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)
  gd_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  x_old <- rep(0, length(par))
  for (i in 1:max_iter) {
    # gradient descent stage
    g <- gr(par)
    x <- par - (lr * g)

    # store gradient descent values
    f <- fn(x)
    gd_fs[i] <- f
    all[i * 2 - 1] <- f

    # momentum stage and update
    par <- x + ifelse(i == 1, 0, mu) * (x - x_old)
    x_old <- x

    # store momentum values
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, gd_fs = gd_fs, all = all)
}
```
In this routine, rather than store the velocity vector, we store the
previous gradient descent result, `x_old`. In order to ensure the
first iteration is gradient descent only, we also need to manually
set the momentum coefficient `mu` to zero on the first iteration,
which is what that `ifelse` expression does.

Also, there's some extra code to calculate and store the function values
after the gradient descent stage. These aren't needed for the optimization
to work, I just want to keep track of the values to demonstrate the 
different methods are in fact equivalent.

### Sutskever Formulation

```{r Sutskever}
# Optimization by Sutskever Nesterov Momentum
#
# Return list also contains mu_fs: function evaluated after momentum
# stage of each iteration; all: function evaluated after gradient descent
# stage and momentum stage, in order.
snag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  v <- rep(0, length(par))

  fs <- rep(0, max_iter)
  mu_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  for (i in 1:max_iter) {
    # momentum stage and update parameters
    mu_step <- mu * v
    par <- par + mu_step

    # store momentum results
    f <- fn(par)
    mu_fs[i] <- f
    all[i * 2 - 1] <- f

    # gradient descent stage
    g <- gr(par)
    gd_step <- -lr * g

    # update and store velocity for next step
    par <- par + gd_step
    v <- mu_step + gd_step

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, mu_fs = mu_fs, all = all)
}
```

### Bengio Formulation
```{r Bengio}
# Optimization by Bengio Nesterov Momentum
bnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    par <- par + mu * mu * v - (1 + mu) * lr * g
    v <- mu * v - lr * g

    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```
Because this implementation only uses a constant momentum coefficient, we 
replace the $\mu_{t+1}\mu_{t}$ term with $\mu_{t}^2$. You would probably have to
using a pretty wacky non-constant momentum schedule where this approximation had any 
major effect anyway... except when the momentum is changing from zero to non-zero,
that is, in which case that's the difference between some momentum and no momentum.
We shall return to this point later.

### Alternative momentum NAG Expression
```{r Alternative NAG}
# Optimization by Nesterov Accelerated Gradient, using a momentum-style
# update expression.
mnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- mu * (v - lr * g) - lr * g
    par <- par + v
    
    # setup v for the next iteration by removing the old gradient contribution
    v <- v + lr * g
    
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

Finally, here is your humble author's expression for NAG, written as a momentum-style update.
To differentiate from the Sutskever and Bengio versions of NAG, I'll refer to it as momentum-NAG
or mNAG.

### Testing with Rosenbrock

Tradition dictates that I must demonstrate the use of these optimizers using the 2D Rosenbrock
function, with a specific starting point:
```{r test function}
par <- c(-1.2, 1)

fn <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  100 * (x2 - x1 * x1) ^ 2 + (1 - x1) ^ 2
}
gr <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  c(
    -400 * x1 * (x2 - x1 * x1) - 2 * (1 - x1),
    200 *      (x2 - x1 * x1))
}
```

Let's run the optimizers for 100 iterations. The point here is not whether we
get really amazing optimization (we don't), but whether the outputs of the Sutskever and
Bengio algorithms are equivalent. It would be a bonus if my mNAG result also worked.
For comparison we'll throw in the classical momentum and as a sanity check,
the vanilla NAG routine.

```{r Sutskever and Bengio formulations}
lr <- 0.001
mu <- 0.95
max_iter <- 100
snag_opt <- snag(par, fn, gr, lr, mu, max_iter)
bnag_opt <- bnag(par, fn, gr, lr, mu, max_iter)
nag_opt <- nag(par, fn, gr, lr, mu, max_iter)
mnag_opt <- mnag(par, fn, gr, lr, mu, max_iter)
cm_opt <- cm(par, fn, gr, lr, mu, max_iter)

sbnag_df <- data.frame(Bengio = bnag_opt$fs,
                 Sutskever = snag_opt$fs, 
                 "Suts Mom" = snag_opt$mu_fs,
                 "NAG" = nag_opt$fs,
                 "mNAG" = mnag_opt$fs,
                 "CM" = cm_opt$fs)
```

Let's have a look at the first few iteration results:

```{r Sutskever and Bengio table head}
knitr::kable(head(sbnag_df), caption = paste(
  "First few evaluations of NAG implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

The first two columns pit the Sutskever vs Bengio formulations directly. And, as
we would expect, they're not the same: the Sutskever iteration result is from
the gradient descent stage, and the Bengio result comes from the momentum stage.
But if we put the momentum results from the Sutskever formulation up, we
can see that they are the same as the Bengio result, but behind by one iteration,
i.e. the Bengio result at iteration $t$ matches the Sutskever momentum stage result
at $t+1$. The final column, indicates the results using the alternative derivation
I provided. It matches the Bengio results. Hurrah.

However, none of the results match the vanilla NAG implementation. A clue to what's
going on is from looking at the classical momentum result in the "CM" column. The
first iteration of classical momentum is always plain gradient descent, and that's also
how NAG works by construction. We can see that both NAG and CM have the same result
on the first iteration, which suggests they're working correctly. The Sutskever
version is also doing gradient descent on its first step. But the other
implementations clearly aren't doing gradient descent on their first iteration,
and on subsequent iterations even the Sutskever version diverges from the NAG
result.

We'll get back to this, but let's just make sure these observations holds up at 
the end of the table too.

```{r table tail}
knitr::kable(tail(sbnag_df), caption = paste(
  "Last few evaluations of NAG implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

The Sutskever, Bengio and momentum NAG implementations all still match up in
the same way they did. CM and NAG are off doing their own thing. I suppose it's
at least reassuring that for this arbitrarily chosen set of learning rate and
momentum coefficient values, the NAG results all do better than classical momentum.
And in fact it seems like not doing plain gradient descent on the first iteration
gives better results, at least for this set of parameters.

### Behavior on first iteration

The reason for the differing behaviors of the NAG implementations is down to
two things:

* For some implementations (the Bengio and momentum NAG formulations), simply 
setting the initial velocity vector, $v_0$, to zero isn't enough to get gradient 
descent on the first iteration.
* The Sutskever version of NAG does a different pattern of gradient descent and
momentum stages in its iterations, compared to vanilla NAG.

For the first point, let's look at the Bengio formulation again:

$$\Theta_{t+1} = 
\Theta_t + \mu_{t+1} \mu_t v_t - \left(1 + \mu_{t+1} \right) \varepsilon_t \nabla f\left(\Theta_t\right) 
$$
Let's see the result for the first iteration with our current strategy
of setting the initial velocity vector to zero:

$$\Theta_1 = 
\Theta_0 - \left(1 + \mu_1 \right) \varepsilon_t \nabla f\left(\Theta_0\right) 
$$
Without also specifying that the momentum coefficient should be zero on the first
iteration, we get an extra long gradient descent step. The same applies to the
momentum NAG expression.

That explains the Bengio formulation's difference from the vanilla NAG result.
What about the Sutskever formulation? Let's think about the chain of parameter 
updates that actually take place over the first few iterations:

For standard NAG, the chain is: gradient descent stage, momentum stage, gradient
descent stage, momentum stage. Except the momentum stage results in zero change 
on iteration one because of the zero velocity vector, so what actually happens 
is: gradient descent stage, gradient descent stage, momentum stage. Let's call 
that `g|gm` for short with the bar indicating where the end of an iteration occurs.
To make the pattern more obvious, the first three iterations look like `g|gm|gm`. 
This is the same pattern as classical momentum, althought the `m` stages are 
obviously different in content. 

What about the Sutskever formulation? Its pattern for the first two iterations is: 
momentum stage, gradient descent stage, momentum stage, gradient descent stage. 
Except, once again, the momentum stage doesn't happen on the first iteration, 
so you actually get: gradient descent stage, momentum stage, gradient descent 
stage, and so on. So for the Sutskever forumulation, the first three iterations 
look like `g|mg|mg`.

Therefore, as you can see, the initial set of gradient and momentum stages is 
different for the Sutskever formulation, which is able to strictly interleave
gradient descent and momentum stages, whereas NAG begins, like classical momentum,
with two gradient descent stages. This difference shouldn't be too surprising, 
because as we saw when deriving the Sutskever formulation, we conveniently
"forget" about an initial gradient descent stage compared to NAG.

Anyway, all this means that the parameters end up in different locations after
the first iteration.

## Consistent NAG momentum algorithms

It's not clear to me that there is any particular theoretical or practical 
benefit to trying to bend these different formulations to give either 
gradient descent on the first iteration or to having them give the same pattern
as vanilla NAG. But given the effort that's been made to show that NAG
is like classical momentum in some respects, we may as well see what it
would take.

### Sutskever Formulation consistent with NAG

```{r Sutskever consistent with NAG}
# Sutskever Nesterov Momentum consistent with NAG
#
# Extra parameter wait: wait this number of extra iterations before applying
# momentum. Needed only to sync up with other implementations of nesterov
# momentum: set wait to 1 to make mu_f at iter i match the output of the
# other implementations at iter i-1.
snagc <- function(par, fn, gr, lr, mu, max_iter = 10, wait = 0) {
  v <- rep(0, length(par))

  fs <- rep(0, max_iter)
  mu_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  for (i in 1:max_iter) {
    # momentum stage and update parameters
    mu_step <- ifelse(i > wait + 1, mu, 0) * v
    par <- par + mu_step

    # store momentum results
    f <- fn(par)
    mu_fs[i] <- f
    all[i * 2 - 1] <- f

    # gradient descent stage
    g <- gr(par)
    gd_step <- -lr * g

    # update and store velocity for next step
    par <- par + gd_step
    v <- mu_step + gd_step

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, mu_fs = mu_fs, all = all)
}
```

The Sutskever formulation already does gradient descent on its first iteration,
so there's not a huge change required for the algorithm. Instead, I've introduced
a new parameter `wait`, that indicates the number of iterations to wait before
applying momentum. If this is set to `0`, then you get the current behavior.
Set it to `1` and the momentum step will be zero on the second iteration, too.
That gives us the two gradient descent stages in a row that should sync us up
with NAG.

### Bengio Formulation consistent with NAG

```{r Bengio consistent with NAG}
# Bengio Nesterov Momentum consistent with NAG
bnagc <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    if (i == 1) {
      par <- par - lr * g
    }
    else {
      par <- par + (mu * mu * v) - ((1 + mu) * lr * g)
      v <- (mu * v) - (lr * g)
    }
    
    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

"Fixing" the Bengio formulation turns out to be a bit more work than
you might think. As we noticed earlier, to get gradient decent on the
first iteration, $\mu_1$, needs to be set to zero. However, as a result
this means the assumption we made that $\mu_{t}\mu_{t-1} = \mu_{t-1}^2$
isn't correct for iteration 2: $\mu_2\mu_1 = 0 \neq \mu_2^2$. However,
we can't just set `mu` to 0 for the second iteration, because the
non-zero $\mu_2$ is used on its own as part of the gradient descent 
expression. 

There are two ways to deal with this: start storing the
value of `mu` from the previous iteration, even though I purposely
chose a constant value of `mu` to keep things simple, or just force
$v_1$ to be zero by not updating it on the first iteration. I chose
the latter.

### Momentum NAG consistent with NAG
```{r Momentum NAG consistent with NAG}
# momentum NAG consistent with NAG
mnagc <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- (ifelse(i == 1, 0, mu) * (v - lr * g)) - lr * g
    par <- par + v
    
    # setup v for the next iteration by removing the old gradient contribution
    v <- v + lr * g
    
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```
And last, the rewritten momentum NAG update. This required the least modification from 
the original routine: simply set the momentum coefficient to zero on the first 
iteration.

### Results for the "NAG-consistent" routines

Time to look at some numbers again. We'll compare the new versions of
the Sutskever, Bengio and Momentum formulations of NAG with vanilla NAG.
We'll also once again pull out the momentum stage results for Sutskever
so we can compare directly to the Bengio result. 

```{r optimization forcing gradient descent initially}
mnagc_opt <- mnagc(par, fn, gr, lr, mu, max_iter)
snagc_opt <- snagc(par, fn, gr, lr, mu, max_iter, wait = 1)
bnagc_opt <- bnagc(par, fn, gr, lr, mu, max_iter)

ncdf <- data.frame( 
                 cBengio = bnagc_opt$fs,
                 cSutskever = snagc_opt$fs,
                 "Suts Mom" = snagc_opt$mu_fs,
                 NAG = nag_opt$fs, 
                 cmNAG = mnagc_opt$fs,
                 "NAG gd" = nag_opt$gd_fs)
```

Let's have a look at the first few iteration results:

```{r table of initial results}
knitr::kable(head(ncdf), caption = paste(
  "First few evaluations of NAG consistent implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

These are in the same order as the previous table. The Bengio and Sutskever
formulations don't match, as expected. But the Sutskever momentum result in the
third column *does* match the Bengio results from the previous iteration. 
So relative performance between Sutskever and Bengio has been maintained.
But in addition, the NAG result in the fourth column matches the Bengio
and Sutskever momentum results. The momentum NAG result has also been correctly 
updated. Finally, as an extra check, the NAG gradient desent result is shown 
in the final column, which also matches up with the Sutskever results.

And let's take a look at the final few iterations, just to make sure everything still
holds up:

```{r table of final results}
knitr::kable(tail(ncdf), caption = paste(
  "Last few evaluations of NAG consistent implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

That's a relief.

## Conclusions

I pity anyone tasked with implementing Nesterov momentum and demonstrating
that they actually got it right. Imagine naively implementing vanilla NAG,
the Sutskever formulation and the Bengio formulation. Using identical test functions,
starting points and parameters, their output are all different from each other
*but they're all correct*! 

