---
title: "Nesterov Accelerated Gradient and Momentum"
date: "December 31, 2016"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: false
---

*December 16 2021*: I have rewritten a large chunk of this document to make the
derivations a little clearer and added a brief section on the definition of
Nesterov momentum used in the NAdam paper.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

A way to express Nesterov Accelerated Gradient (NAG) in terms of a regular
momentum update was noted by [Sutskever and
co-workers](http://www.jmlr.org/proceedings/papers/v28/sutskever13.html), and
perhaps more importantly, when it came to training neural networks, it seemed to
work better than classical momentum schemes. This was further confirmed by
[Bengio and co-workers](https://arxiv.org/abs/1212.0901), who provided an
alternative formulation that might be easier to integrate into existing
software.

I implemented NAG as part of [mize](https://github.com/jlmelville/mize), but
found it was a bit more difficult than I'd anticipated to test that I'd got the
correct results. 

It seems I am not alone in being confused by the exact details of implementing
NAG as a momentum scheme (e.g. comments and issues in the deep learning projects
[keras](https://github.com/fchollet/keras/issues/966) and
[mocha.jl](https://github.com/pluskid/Mocha.jl/pull/47), which I found from
random googling), so mainly for the benefit of future me, I am going to derive
the Sutskever and Bengio formulations from NAG in tedious detail. I am also
going to derive an alternative expression that I ended up using in mize.
Finally, I will write some R code to demonstrate their equivalence. This will
turn out to be trickier than I anticipated.

I originally tried to stick with the notation used by Sutskever but in 
retrospect that was a mistake. It's too hard to keep everything straight between
the different papers. I will borrow some symbols and try and be very clear about
what they mean and what each expression applies to. Translating between this
document and the original papers may require mentally transposing some symbols
and shifting iteration indices from $t$ to $t+1$ or $t-1$.

Apart from the Sutskever and Bengio papers linked to above, it's worth checking
out the [appendix to the Sutskever paper
(PDF)](http://www.jmlr.org/proceedings/papers/v28/sutskever13-supp.pdf) or the
relevant part (chapter 7) of [Sutskever's
thesis](http://hdl.handle.net/1807/36012). For an introduction to NAG itself,
try the first part of this paper by [O'Donoghue and
Candès](https://arxiv.org/abs/1204.3982) (the rest of it is also good but not
germane to Nesterov momentum expressions).

## Definitions

The goal is to optimize some parameters, $\theta$. This is going to involve
gradient descent, so we will be evaluating the gradient of an objective 
function of those parameters, $\nabla f\left(\theta\right)$, and moving a 
certain distance in the direction of the negative of the gradient, the distance
being related to the learning rate, $\varepsilon$. There will also be a 
momentum term involved, with momentum coefficient $\mu$. The value of the 
parameters, learning rate and momentum at iteration $t$ will be indicated by a 
subscript, e.g. $\theta_t$.

### The Update Vector

A definition which holds across all the methods discussed here is that the
parameters at iteration $t+1$ are related to the parameters at iteration $t$
by an update that involves the addition of a velocity vector, $v$:

$$\theta_{t+1} = \theta_t + v_{t+1} \implies v_{t+1} = \theta_{t+1} - \theta_t$$

This is pretty obvious, but the expressions we will be dealing with here use a
recursive definition of $v_t$, i.e. to usefully define these updates we need to
express $v_t$ in terms of $v_{t-1}$. The exact expression depends on whether we
are using classical momentum or NAG. In both cases, it would make sense to label
a particular vector as $v_t$, but due to the similarity in how they work, the
temptation to substitute the classical momentum recursive definition of $v_t$
into an expression for NAG is almost overwhelming, and also always wrong.

To confuse matters further, some papers (see the Bengio paper for example) 
directly use $v_t$ in expressions that update parameters, but where the $v_t$
doesn't actually apply to the parameter being updated.

Finally, I'll try to be strict about names. When I'm referring to the original
Nesterov Accelerated Gradient, I'll call that NAG. When I'm referring to the
versions that recast it as a momentum scheme, I'll call that Nesterov momentum,
and refer to either the Sutskever or Bengio forumulation where necessary. The
traditional momentum scheme is referred to as "standard" or "regular" momentum
by Bengio, and "classical" by Sutskever. Other authors call it "Polyak" or
"heavy ball" momentum. I'll refer to it as "classical" momentum.

## Classical Momentum

The classical momentum velocity vector is defined by Bengio as:

$$
v_t = \mu_{t-1}v_{t-1} - \varepsilon_{t-1}\nabla f\left(\theta_{t-1}\right)
$$

Sutskever gives the same definition but without any $t$ subscript on the
velocity vector or the momentum coefficient.

We can write out the full update for classical momentum as:

$$\theta_{t+1} = \theta_t + \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t\right)$$

with velocity:

$$v_{t+1} = \mu_t v_t - \varepsilon_t \nabla f\left(\theta_t \right)$$

Later we will want to use the recursive definition of momentum, but it won't
be the full update, so we will not be able to define it as $v_t$. In those
circumstances, I will call the update $m_t$ to avoid confusion:

$$
m_{t+1} = \mu_t m_t - \varepsilon_t \nabla f\left(\theta_t \right)
$$

Remember: $v_{t+1} = m_{t+1}$ *only* in the case of classical momentum.

## NAG

The Nesterov Accelerated Gradient method consists of a gradient descent step,
followed by something that looks a lot like a momentum term, but isn't exactly
the same as that found in classical momentum. I'll call it a "momentum stage"
here. If you look at the Sutskever paper, It's important to note that the
parameters being minimized by NAG are given the symbol $y$, not $\theta$. You'll
see that $\theta$ is the symbol for the parameters after they've been updated by
the gradient descent stage, but before the momentum stage.

I'm going to stick with $\theta_t$ as the parameters being updated. But we do
need to define a symbol that accounts for the state of the parameters after
the gradient descent stage. I'll go with $\phi_{t+1}$. Whether you consider
the intermediate step to be part of iterate $t$ or $t+1$ doesn't really matter,
but different papers use different conventions, so you will need to keep track
of it all or risk getting very confused. In this case I am going with $t+1$ so
it's consistent with any new quantities calculated in an iteration (like
$v_{t+1}$).

Here's the gradient descent stage:

$$
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
$$

followed by the momentum-like stage:

$$
\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right)
$$

That concludes one iteration of NAG. The hard part is actually finding the 
correct learning rate and momentum value in order to get the convergence
guarantees that make the method attractive, but that needn't concern us. 
Sutskever in his thesis suggests manual tuning to get an optimal result (at 
least for deep learning applications) so we are in heuristic territory here 
anyway.

## Sutskever Nesterov Momentum

The key idea behind the Sutskever momentum derivation is to shift the
perspective about which of the parameters we want as the result of the
iteration, from $\theta$ to $\phi$. Rather than having the optimization
iterations proceed as "gradient descent, momentum (end of iteration 1), gradient
descent, momentum (end of iteration 2), gradient descent etc." move the boundary
of where the iterations end by a half-iteration to get "momentum, gradient
descent (end of iteration 1), momentum, gradient descent (end of iteration 2)
etc.". This leaves a phantom gradient descent step that used to be first stage
of the first iteration, now floating in the nether regions of iteration zero,
but you can just pretend that the starting position is the result of gradient
descent from some other arbitrary starting position.

Perhaps a visualization will help. Here are two full iterations of NAG in its
standard form, separated by boxes so you can see where one interation ends and
the other begins.

$$
\boxed{
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) \\
\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right)} \\
\boxed{
\phi_{t+2} = \theta_{t+1} - \varepsilon_{t+1} \nabla f\left(\theta_{t+1} \right) \\
\theta_{t+2} = \phi_{t+2} + \mu_{t+1}\left(\phi_{t+2} - \phi_{t+1} \right)
}
$$

The Sutskever derivation says shift the boundaries of the iteration so that it 
starts one stage later. This is the same four equations as above, but with a
box showing the new definition of the iteration:

$$
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) \\
\boxed
{
\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right) \\
\phi_{t+2} = \theta_{t+1} - \varepsilon_{t+1} \nabla f\left(\theta_{t+1} \right)
} \\
\theta_{t+2} = \phi_{t+2} + \mu_{t+1}\left(\phi_{t+2} - \phi_{+1} \right)
$$

Now that the definition of where we start and finish each iteration has changed
we need to re-label some variables: uses of $\phi$ and $\varepsilon$ need to be
shifted back one step of $t$ to give:

$$
\theta_{t+1} = \phi_{t} + \mu_t\left(\phi_{t} - \phi_{t-1} \right) \\
\phi_{t+1} = \theta_{t+1} - \varepsilon_{t} \nabla f\left(\theta_{t+1} \right)
$$

Then because our update is now in terms of $\phi$ rather than $\theta$, we can
label $\phi_{t} - \phi_{t-1}$ as $v_t$. The new definition of $\theta_{t+1}$
is then:

$$
\theta_{t+1} = \phi_{t} + \mu_t v_t
$$

And we can now substitute that definition for $\theta_{t+1}$ into the momentum
stage expression, which lets us write this half-shifted NAG iteration in one
line:

$$
\phi_{t+1} = \phi_{t} + \mu_t v_t - \varepsilon_{t} \nabla f\left(\phi_{t} + \mu_t v_t \right)
$$

There you have it: this looks just like the classical momentum update, except
that the gradient is calculated after the momentum update. Hence, one can do NAG
by simply reversing the order in which the update is usually carried out: do the
momentum stage first, update the parameters, and then do the gradient descent
part. Just like Sutskever said you could.

## Bengio Nesterov Momentum

I found working through the Bengio version of Nesterov momentum a bit
brain-breaking. It starts from the Sutskever definition and then defines a new
variable, $\Theta$, which represents the parameters after the momentum update.
But that just moves us one half-step forward in the iteration, so we are back to
doing things in the original NAG way: gradient descent then momentum-style
update. So their paper's $\Theta$ is the same as our $\theta$. Its explanation
of "committing to the 'peekedahead' parameters" and then "backtracking by the
same amount before each update", also never landed with me (this is a reflection
on my mental capacity, not the paper). Finally, it expresses the update in terms
of $v$, but that's still the the update in terms of the difference in $\phi$,
not $\theta$. Unlike this document it also uses $t-1$ and $t$ to index the
parameters that are being updated, so you will have to remember to add or
subtract 1 from the subscripts when looking between here and the paper, but
that's the least of your problems honestly.

It's all too much. I can't get their explanation to stick in my brain, even
though I can follow the individual steps and get the right expression. So I will
show you how to arrange the standard NAG expressions into their form, rather
than starting from the Sutskever result.

First, let's remove the confusion around using $v$ as an update vector which
isn't the change in the parameters that we are updating each iteration. Instead
of $\phi_{t+1} - \phi_{t} = v_{t+1}$, define:

$$
\phi_{t+1} - \phi_{t} = b_{t+1}
$$

(the 'b' is for Bengio). Our goal now is to end up with an expression that is
defined using $\theta_t$ (no $\phi_t$ allowed) and $b_t$.

Now we are going to go back to using the original NAG expression so to help you
scrub all the shifts in $t$ from your mind here is the expression again:

$$
\phi_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) \\
\theta_{t+1} = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right)
$$
Start by replacing all uses of $\phi_{t+1}$ in the second equation with the definition 
in terms of $\theta_t$ given in the first equation:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) + 
  \mu_t \left( \theta_t - \varepsilon_t \nabla f\left(\theta_t \right) - \phi_{t} \right) \\
 & = \left(1 + \mu \right) \theta_t 
 - \left(1 + \mu_t \right) \varepsilon_t \nabla f\left(\theta_t \right)
  - \mu_t \phi_t
\end{split}
\end{equation}  
$$

So far, so un-promising. Time to get $b_t$ involved. Remember how we defined:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \phi_{t+1} + \mu_t\left(\phi_{t+1} - \phi_{t} \right) \\
 & = \phi_{t+1} + \mu_t b_{t+1}
\end{split}
\end{equation}
$$

Let's shift everything back one iteration:

$$
\theta_{t} = \phi_{t} + \mu_{t-1} b_{t} \\
\implies  \phi_{t} = \theta_{t} - \mu_{t-1} b_{t}
$$

And now substitute for $\phi_t$ in the last term in our NAG update:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \left(1 + \mu \right) \theta_t 
 - \left(1 + \mu_t \right) \varepsilon_t \nabla f\left(\theta_t \right)
  - \mu_t \phi_t \\
 & = \left(1 + \mu \right) \theta_t 
 - \left(1 + \mu_t \right) \varepsilon_t \nabla f\left(\theta_t \right)
  - \mu_{t} \theta_t + \mu_{t-1} \mu_t b_t
\end{split}
\end{equation}  
$$

The final step is to gather together the $\theta_t$ terms and the result is:

$$
\theta_{t+1} = \theta_t + \mu_{t-1} \mu_t b_t
 - \left(1 + \mu_t \right) \varepsilon_t \nabla f\left(\theta_t \right)
$$

Ta da. There it is, just like in the paper, subject to the differences in
symbols and when the iteration ticks over from $t$ to $t+1$ and so on.

The advantage of this expression for the Nesterov momentum is that it doesn't
require calculating a gradient at a non-standard position, and only requires a
modification to the coefficients used to calculate the velocity, which is
probably an easier change to make to an existing codebase which already uses
classical momentum. However, because this expression is recursive on $b_t$
rather than the update vector $v_t$, you will need to remember to store the a
different vector than you would with classical momentum. It's not really clear
to me whether that is an easy or hard thing to do with most optimization code
bases vs having to make code changes to calculate the gradient at a different
location than the current parameter vector, as you have to do with the Sutskever
version.

You should also remember that while the parameters from this version and the
Sutskever version should converge to the same values from the same starting
positions, on an iteration-to-iteration version level, you will get different
values for the parameters, because they are half a step out of sync with each
other.

## An Alternative Expression for NAG

Let's go back to the original formulation of NAG and now instead of making
$\theta$ the variables after gradient descent, we'll put them where $y$ was 
originally used. The variables after gradient descent I'll refer to as $\phi$.

$$\phi_t = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)$$

$$\theta_{t+1} = \phi_t + \mu_t\left(\phi_t - \phi_{t-1} \right)$$


Now, let's just write out the momentum stage in terms of $\theta$, substituting
$\phi$ wherever we find it:

$$
\theta_{t+1} = \theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
+ \mu_t \left[\theta_t - \varepsilon_t \nabla f\left(\theta_t \right)
- \theta_{t-1} + \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
\right]
$$

Rearranging:

$$
\theta_{t+1} = \theta_t 
+ \mu_t \left[\theta_t - \theta_{t-1}
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
- \varepsilon_t \nabla f\left(\theta_t \right)
\right]
- \varepsilon_t \nabla f\left(\theta_t \right)
$$

Finally, we can subtitute in $v_t$ for the first two terms in the square
brackets, to give:

$$
\theta_{t+1} = \theta_t 
+ \mu_t \left[v_t
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
- \varepsilon_t \nabla f\left(\theta_t \right)
\right]
- \varepsilon_t \nabla f\left(\theta_t \right)
$$

with velocity:

$$
v_{t+1} =  
 \mu_t \left[v_t
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
- \varepsilon_t \nabla f\left(\theta_t \right)
\right]
- \varepsilon_t \nabla f\left(\theta_t \right)
$$

This looks a lot like the classical momentum expression, but with the
velocity vector modified to first remove the contribution of the gradient 
descent from the previous iteration, and replace it with the gradient descent 
contribution from the *current* iteration. Gives an interesting insight into
the idea of the Nesterov momentum using a form of "lookahead" with the
gradient descent.

You could also choose to expand the velocity expression to make it look a bit 
like the Bengio formulation:

$$
v_{t+1} =  
\mu_t \left[v_t
+ \varepsilon_{t-1} \nabla f\left(\theta_{t-1} \right)
\right]
- \left(1 + \mu_t\right) \varepsilon_t \nabla f\left(\theta_t \right)
$$

but as this version can't be expressed as the classical momentum form with
different coefficients, the way the Bengio formulation can be, it probably
doesn't gain you anything in terms of implementation, except you can expand it 
and rearrange it further to give:

$$
v_{t+1} =  
\mu_t v_t - \varepsilon_{t} \nabla f\left(\theta_{t} \right)
+
\mu_t 
\left[
\varepsilon_{t-1} \nabla f \left(\theta_{t-1}\right) 
-
\varepsilon_{t} \nabla f \left(\theta_{t}\right)
\right]
$$
which now resembles the classical momentum expression with an extra momentum 
term. I haven't found a definitive reference for this expression, or what, if 
any, extra insight it provides, but user 'denis' uses this expression in an
[answer on the Cross Validated Stack Exchange](https://stats.stackexchange.com/a/233430),
and refers to the third form as "gradient momentum". Another way to look at this
is to say that the third term reduces the contribution of the classical momentum 
component of the update and increases the contribution of the gradient descent, 
with the degree of weighting controlled by $\mu_{t}$. 

[Quasi-hyperbolic momentum](https://arxiv.org/abs/1810.06801) uses a similar 
formulation, but introduces an extra parameter that allows QHM to generalize
over classical momentum and NAG, subject to the restriction of a fixed learning
rate. The Stack Exchange answer given above also suggests that the "gradient
momentum" term can be set independently of $\mu$ and Denis Yarats is one of the
authors of the QHM paper, so perhaps we see the origins of QHM in this answer.

The setup and presentation of the update equations for QHM in that paper make it
a bit hard to see the connection between it and the form of NAG given here. The
supplemental material in 
[Understanding the Role of Momentum in Stochastic Gradient Methods](https://papers.nips.cc/paper/9158-understanding-the-role-of-momentum-in-stochastic-gradient-methods)
might be of interest, or some extra 
[notes on QHM](https://jlmelville.github.io/mize/qhm.html) I wrote for my own
benefit.

## Unified Momentum

*December 12 2021*: Previously in this section I wrote down a QHM-like
expression I called "generalized" momentum. This was a bad choice of words
because there is already a concept in physics called generalized momentum, which
is nothing to do with this. Fortunately, in 2018 [Zou and
co-workers](https://arxiv.org/abs/1808.03408) published a paper with the same
expression and called it "unified stochastic momentum". The stochastic bit is
because it is in the context of stochastic gradient descent but that has no
effect on the expression. So let's call it "unified momentum" instead:

Similarly to QHM, we can write a related ~~"generalized"~~ "unified" momentum 
update by introducing a parameter $\beta$ to give:

$$
v_{t+1} =  
\mu_t v_t - \varepsilon_{t} \nabla f\left(\theta_{t} \right)
+
\beta_t \mu_t 
\left[
\varepsilon_{t-1} \nabla f \left(\theta_{t-1}\right) 
-
\varepsilon_{t} \nabla f \left(\theta_{t}\right)
\right]
$$
where setting $\beta_t = 0$ will give classical momentum and $\beta_t = 1$ gives
Nesterov. The stochastic unified momentum paper calls $\beta$ the "interpolation
factor" (and use the symbol $\lambda$ instead of $\beta$).

In 2020, a similar expression was given by [Ziyin and
co-workers](https://arxiv.org/abs/2002.04839), but didn't give it a catchy name.
They don't cite the Zou paper, so presumably they discovered it independently.
I'd like to find the earliest use of this interpolation, but neither the Zou nor
the Ziyin paper give a reference to any earlier paper. Perhaps it's just
incredibly obvious?

From the look of the expression, it seems that a major downside of this 
expression is that calculating the parameters for iteration $t+1$ requires 
storage of information from not only iteration $t$ but from iteration $t-1$ too.
But you don't necessarily have to do any extra storage with an implementation
that used this version of NAG. At the end of an iteration, when saving the
velocity vector for the next iteration, you change:

$$v_{t-1} \leftarrow v_t$$
to:

$$v_{t-1} \leftarrow v_t + \varepsilon_{t} \nabla f\left(\theta_t \right)$$

and then when calculating the momentum term, change:

$$\mu_t v_t$$
to:
$$\mu_t \left[v_t - \varepsilon_{t} \nabla f\left(\theta_t \right)\right]$$

## Unrolling NAG and Classical Momentum

As a way to understand the difference between classical momentum and NAG, 
let's write out the first few steps of the optimization, substituting in the 
recursive definitions with the previous step, and see what emerges.

Because I have finally got tired of writing out 
$\varepsilon_t \nabla f\left(\theta_t \right)$ all over the place (and it's 
visually distracting), let's define:

$$
-\varepsilon_t \nabla f\left(\theta_t \right) \equiv s_t
$$

to represent the steepest descent direction. Note that this includes the
negative sign. Also, I am going to assume that the momentum is constant, to save
on some more notational clutter.

Initial coordinates are at $\theta_0$.

### Classical Momentum

Using the slightly briefer symbols, the classical momentum update is:

$$
\theta_{t+1} = \theta_t + s_t + \mu v_t
$$

And the first four updates (including steepest descent on the first iteration)
are as follows:

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + s_0 \\
\\
\theta_2 & = \theta_1 + s_1 + \mu v_1 \\
 & = \theta_1 + s_1 + \mu s_0 \\
\\
\theta_3 & = \theta_2 + s_2 + \mu v_2 \\
 & = \theta_2 + s_2 + \mu \left( s_1 + \mu s_0 \right) \\
 & = \theta_2 + s_2 + \mu s_1 + \mu^2 s_0 \\
\\
\theta_4 & = \theta_3 + s_3 + \mu v_3 \\
 & = \theta_3 + s_3 + \mu \left( s_2 + \mu s_1 + \mu^2 s_0 \right)\\
 & = \theta_3 + s_3 + \mu s_2 + \mu^2 s_1 + \mu^3 s_0 \\
\end{split}
\end{equation}
$$

### NAG

We'll use this straight-forward expression for NAG:

$$
\theta_{t+1} = \theta_t + s_t
+ \mu \left(\theta_t + s_t
- \theta_{t-1} - s_{t-1}
\right)
$$

with just this bit of re-arranging so terms are all grouped together and the 
$v_t$ is at the end, which makes it easier to follow the substitution of $v_t$:

$$
\theta_{t+1} = \theta_t + \left(1 + \mu \right) s_t - \mu s_{t-1} + \mu v_t
$$

#### The first step

*December 14 2021* (this section describing your choices for the first step is
new).

Like classical momentum, the only sensible direction to move in on the first
step is the steepest descent direction because we have no "memory" of any
previous directions to use. But a strict application of the NAG update would set
$v_0 = 0$ and $s_{-1} = 0$ and leave us with:

$$
\theta_{1} = \theta_0 + \left(1 + \mu \right) s_0
$$

This is still moving in the direction of the negative gradient, but it takes a
longer step size than classical momentum. This doesn't make a lot of sense to me
as a good first step without any contributions from previous updates or
gradients. Why have a momentum term involved at all when most of what that $\mu$
gets applied to in subsequent iterations isn't around?

Also, as we will see later, taking the long step makes reconciling the different
versions of NAG a bit harder than it needs to be . Plus, I see the appeal of
both classical momentum and NAG taking the step on the first iteration when
there is no history of updates to work on. So I won't blame you if you decide to
go with steepest descent with step size of length $\varepsilon_1$ rather than
$\left(1 + \mu \right) \varepsilon_1$.

Unfortunately, as you will discover, your efforts to avoid taking the "long"
steepest descent step will be in vain. If you don't do it on the first
iteration, it just shows up again in the second iteration. 

We'll look at the results with both choices, and it's honestly not that
consequential, but probably going with the $\left(1 + \mu \right)$ step size
on the first iteration is the right move.

Having decided on the "long" version of the first steepest descent step, here
it is along with the subsequent three steps of NAG:

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + \left( 1 + \mu \right) s_0 \\
\\
\theta_2 & = \theta_1 + \left( 1 + \mu \right) s_1 - \mu s_0 + \mu v_1 \\
 & = \theta_1 + \left(1 + \mu \right) s_1 - \mu s_0 + \mu \left( 1 + \mu \right) s_0 \\
 & = \theta_1 + \left(1 + \mu \right) s_1 + \mu^2 s_0 \\
\\ 
\theta_3 & = \theta_2 + \left( 1 + \mu \right) s_2 - \mu s_1 + \mu v_2 \\
& = \theta_2 + \left(1 + \mu \right) s_2 - \mu s_1 + \mu \left(1 + \mu \right) s_1 + \mu^3 s_0\\
& = \theta_2 + \left(1 + \mu \right) s_2 + \mu^2 s_1 + \mu^3 s_0 \\
\\
\theta_4 & = \theta_3 + \left(1 + \mu \right) s_3 - \mu s_2 + \mu v_3 \\
& = \theta_3 + \left(1 + \mu \right) s_3 - \mu s_2 + \mu \left[ \left(1 + \mu \right) s_2 + \mu^2 s_1 + \mu^3 s_0 \right] \\
& = \theta_3 + \left(1 + \mu \right) s_3 - \mu s_2 + \mu \left(1 + \mu \right) s_2 + \mu^3 s_1  + \mu^4 s_0 \\
& = \theta_3 + \left(1 + \mu \right) s_3 + \mu^2 s_2 + \mu^3 s_1 + \mu^4 s_0
\end{split}
\end{equation}
$$

You can see a pattern emerging here at this point, I hope. These are quite
similar in form compared to the classical momentum.

### NAG with "short" steepest descent

What happens if we decide to go with the shorter steepest descent step and make
the same first step as we do with classical momentum? Here's how that plays out:

$$
\begin{equation}
\begin{split}
\theta_{1} & = \theta_0 + s_0 \\
\\
\theta_2 & = \theta_1 + \left( 1 + \mu \right) s_1 - \mu s_0 + \mu v_1 \\
 & = \theta_1 + \left( 1 + \mu \right) s_1 - \mu s_0 + \mu s_0 \\
 & = \theta_1 + \left( 1 + \mu \right) s_1 \\
\\ 
\theta_3 & = \theta_2 + \left( 1 + \mu \right) s_2 - \mu s_1 + \mu v_2 \\
& = \theta_2 + \left(1 + \mu \right) s_2 - \mu s_1 + \mu \left(1 + \mu \right) s_1 \\
& = \theta_2 + \left(1 + \mu \right) s_2 + \mu^2 s_1 \\
\\
\theta_4 & = \theta_3 + \left( 1 + \mu \right) s_3 - \mu s_2 + \mu v_3 \\
& = \theta_3 + \left( 1 + \mu \right) s_3 - \mu s_2 + \mu \left[ \left( 1 + \mu \right) s_2 + \mu^2 s_1 \right] \\
& = \theta_3 + \left( 1 + \mu \right) s_3 - \mu s_2 + \mu \left( 1 + \mu \right) s_2 + \mu^3 s_1 \\
& = \theta_3 + \left( 1 + \mu \right) s_3 + \mu^2 s_2 + \mu^3 s_1
\end{split}
\end{equation}
$$

Overall the results aren't that different: nearly all of the terms are present
with the same coefficients. However we can see that, as I mentioned above, the
second step ends up being steepest descent as well, because the contribution of
the $s_1$ terms cancel out. And once it's gone, it's gone for good: all the
subsequent iterations lack the contribution of the $s_1$ term. And the second
iteration is the "long" steepest descent step, the thing we were trying to avoid
in the first place.

So that's two disadvantages to deal with: first, the long steepest descent is
going to happen anyway. Second, the disappearance of the contribution of the
first gradient after the first iteration. None of the other gradients behave
like that and it also slightly messes with the relationship between the weights
we get with NAG and classical momentum.

On balance, I recommend using the "long" step for the first iteration of NAG.
If I need to distinguish between the two methods I will call the version where
we use the "short" step "NAGs".

#### Relative Weights

Here's a table showing how on the 4th step, the two methods weight the current
and previous gradients in deciding on the next direction:

| Momentum Type |  $s_3$     | $s_2$   | $s_1$   | $s_0$   |
:--------------:|:----------:|:-------:|:-------:|:--------:
Classical       | 1          | $\mu$   | $\mu^2$ | $\mu^3$ 
NAG             | 1 + $\mu$  | $\mu^2$ | $\mu^3$ | $\mu^4$
NAGs            | 1 + $\mu$  | $\mu^2$ | $\mu^3$ | 0

From this it's clear that the difference between NAG and classical momentum is
that NAG puts more weight on recent gradients: another way of looking at it is
to say that NAG forgets old gradients more quickly.

The regularity of the difference between classical momentum and NAG weights
suggest that there should be a way to express the NAG update in terms of a
classical momentum update. See the section on 'Dozat Nesterov momentum' below
for more on that.

Let's write some R code to generate a table to compare how the relative weights
turn out between CM and NAG at different values of $\mu$. We'll normalize the 
contributions so they sum to 1.

```{r generate relative weights}
cm_weights <- function(mu) { c(1, mu, mu * mu, mu * mu * mu) }

nags_weights <- function(mu) { c(1 + mu, mu * mu, mu * mu * mu, 0) }

nag_weights <- function(mu) { c(1 + mu, mu * mu, mu * mu * mu, mu * mu * mu * mu) }


mumat <- function(wfun, mus = c(0.1, 0.25, 0.5, 0.75, 0.9, 0.99)) {
  rel_weights <- matrix(nrow = length(mus), ncol = 5)
  for (i in 1:length(mus)) {
    mu <- mus[i]
    weights <- wfun(mu)
    rel_weights[i, ] <- c(mu, weights / sum(weights))
  }
  colnames(rel_weights) <- c("mu", "s3", "s2", "s1", "s0")

  rel_weights
}
```

#### Classical momentum weights

```{r table of classical momentum relative weights}
knitr::kable(mumat(cm_weights), digits = 4)
```

#### NAG weights

```{r table of NAG relative weights}
knitr::kable(mumat(nag_weights), digits = 4)
```

#### NAGs weights

```{r table of NAGs relative weights}
knitr::kable(mumat(nags_weights), digits = 4)
```

At high momentum, classical momentum is only placing around 25-30% of the weight
on the current gradient, whereas for NAG, it's 45-50%. For NAGs, the weight is
even higher at 50-55%.

### Unrolling with a momentum schedule

I kept $\mu$ constant in the unrolling expressions so far to keep things neat
and to highlight some similarities between CM and NAG. But what if we want $\mu$
to vary with iteration? In our derivations of Nesterov momentum schemes we have
explicitly accounted for $\mu_t$ so let's take a look (and brace ourselves for
the expressions to lose a lot of readability):

#### Classical Momentum with a Schedule

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + s_0 \\
\\
\theta_2 & = \theta_1 + s_1 + \mu_1 v_1 \\
 & = \theta_1 + s_1 + \mu_1 s_0 \\
\\
\theta_3 & = \theta_2 + s_2 + \mu_2 v_2 \\
 & = \theta_2 + s_2 + \mu_2 \left( s_1 + \mu_1 s_0 \right) \\
 & = \theta_2 + s_2 + \mu_2 s_1 + \mu_2 \mu_1 s_0 \\
\\
\theta_4 & = \theta_3 + s_3 + \mu_3 v_3 \\
 & = \theta_3 + s_3 + \mu_3 \left( s_2 + \mu_2 s_1 + \mu_2 \mu_1 s_0 \right)\\
 & = \theta_3 + s_3 + \mu_3 s_2 + \mu_3 \mu_2 s_1 + \mu_3 \mu_2 \mu_1 s_0 \\
\end{split}
\end{equation}
$$
Already this is looking pretty gross.

#### NAG with a Momentum Schedule

$$
\begin{equation}
\begin{split}
\theta_1 & = \theta_0 + \left( 1 + \mu_0 \right) s_0 \\
\\
\theta_2 & = \theta_1 + \left( 1 + \mu_1 \right) s_1 - \mu_1 s_0 + \mu_1 v_1 \\
 & = \theta_1 + \left(1 + \mu_1 \right) s_1 - \mu_1 s_0 + \mu_1 \left( 1 + \mu_0 \right) s_0 \\
 & = \theta_1 + \left(1 + \mu_1 \right) s_1 + \mu_1 \mu_0 s_0 \\
\\ 
\theta_3 & = \theta_2 + \left( 1 + \mu_2 \right) s_2 - \mu_2 s_1 + \mu_2 v_2 \\
& = \theta_2 + \left(1 + \mu_2 \right) s_2 - \mu_2 s_1 + \mu_2 \left(1 + \mu_1 \right) s_1 + \mu_2 \mu_1 \mu_0 s_0\\
& = \theta_2 + \left(1 + \mu_2 \right) s_2 + \mu_2 \mu_1 s_1 + \mu_2 \mu_1 \mu_0 s_0 \\
\\
\theta_4 & = \theta_3 + \left(1 + \mu_3 \right) s_3 - \mu_3 s_2 + \mu_3 v_3 \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 - \mu_3 s_2 + \mu_3 \left[ \left(1 + \mu_2 \right) s_2 + \mu_2 \mu_1 s_1 + \mu_2 \mu_1 \mu_0 s_0 \right] \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 - \mu_3 s_2 + \mu_3 \left(1 + \mu_2 \right) s_2 + \mu_3 \mu_2 \mu_1 s_1 + \mu_3 \mu_2 \mu_1 \mu_0 s_0 \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 + \mu_3 \mu_2 s_2 + \mu_3 \mu_2 \mu_1 s_1 + \mu_3 \mu_2 \mu_1 \mu_0 s_0
\end{split}
\end{equation}
$$

I can see why I was in no hurry to write all that out in full until now. No
point showing the "NAGs" version for this, it's the same as with the constant
$\mu$ case: the $s_0$ term vanishes after the first iteration.

Here's a comparison of the weights applied to the different gradients at 
$\theta_4$:

| Momentum Type |  $s_3$       | $s_2$         | $s_1$               | $s_0$                   |
:--------------:|:-------------|:--------------|:--------------------|:-------------------------
Classical       | 1            | $\mu_3$       | $\mu_3 \mu_2$       | $\mu_3 \mu_2 \mu_1$ 
NAG             | 1 + $\mu_3$  | $\mu_3 \mu_2$ | $\mu_3 \mu_2 \mu_1$ | $\mu_3 \mu_2 \mu_1 \mu_0$

Now the momentum is not constant at each iteration, we see a more complex
relationship between the CM and NAG weights. There definitely doesn't seem to
be a simple re-weighting that could map from CM to NAG here: each gradient
needs re-weighting by a different $\mu_t$.

## Dozat Nesterov Momentum

*18 December 2021*: A new section (currently in a serviceable but somewhat
unsatisfactory state) on the expression for Nesterov momentum given by Timothy
Dozat in the NAdam paper.

A paper by [Dozat integrating Nesterov
momentum](https://openreview.net/forum?id=OM0jvwB8jIp57ZJjtNEZ) into the
[Adam](https://arxiv.org/abs/1412.6980) stochastic gradient descent method shows
how to get from the CM to NAG by re-weighting the CM update. As we just saw,
allowing $\mu$ to vary at each iterate has complicated the relationship between
CM and NAG, so we need to define a new CM-like update.

Back when I defined CM, I said it would be useful to define the recursive
momentum equation as $m_t$, identifying as separate from the parameter update, 
$v_{t}$, even though they coincide in CM. Here's where it comes in useful to have
separated those two concepts. We now need a CM-like definition, which differs
from $m_{t}$ only in that we use the momentum coefficient from the previous
iterate. This new definition I will label as $m^{\prime}_{t}$:

$$
\begin{equation}
\begin{split}
m_{t+1} & = \mu_{t} m_t + s_t\\
m^{\prime}_{t+1} & = \mu_{t-1} m^{\prime}_t + s_t\\
\end{split}
\end{equation}
$$
For NAG, the update vector $v_{t+1}$ can now be defined in terms of $m^{\prime}_{t+1}$,
using the current iterate's momentum coefficient:

$$
v_{t+1} = \mu_{t} m^{\prime}_{t+1} + s_t
$$

This is like doing momentum twice per iteration (I'd love to say "double
momentum" but that's [already a thing](https://arxiv.org/abs/2102.07367)). If we
expand the update to be in terms of $m^{\prime}_t$ rather than
$m^{\prime}_{t+1}$ we get:

$$
\begin{equation}
\begin{split}
v_{t+1} & = \mu_{t} \left( \mu_{t-1} m_t^{\prime} + s_t \right) + s_t \\
& = \mu_{t-1} \mu_{t} m_t^{\prime} + \left(1 + \mu_t \right) s_t
\end{split}
\end{equation}
$$

So here's the overall Dozat Nesterov momentum update, rearranged a bit to make
the forthcoming un-rolling easier to write out:

$$
\theta_{t+1} = \theta_t + \left(1 + \mu_t \right) s_t + \mu_{t-1} \mu_{t} m_t^{\prime}
$$

At this point, I would love to say that I can get to this expression for NAG
starting from the expression we have been using so far. But I admit that it's
not obvious to me how to get from the general recursive-in-$v_t$ expression to
one which is recursive in $m^{\prime}_t$. The key maye be to notice the
connection between this expression and two others: the Bengio Nesterov momentum
expression and the expanded and rearranged NAG expression:

$$
\begin{equation}
\begin{split}
\theta_{t+1} & = \theta_t + \left(1 + \mu_t \right) s_t + \mu_{t-1} \mu_{t} m_t^{\prime} \\
\theta_{t+1} & = \theta_t + \left(1 + \mu_t \right) s_t + \mu_{t-1} \mu_t b_t \\
\theta_{t+1} & = \theta_t + \left(1 + \mu_t \right) s_t + \mu_t \left(v_t - s_{t-1} \right)
\end{split}
\end{equation}
$$

This means that:

$$
m_t^{\prime} = b_t = \frac{\left(v_t - s_{t-1} \right)}{\mu_{t-1}}
$$

Perhaps one day I will return to this section and fill in the obvious details.
For now, let's just unroll the first few iterations of the Dozat expression:

$$
\begin{equation}
\begin{split}
m^{\prime}_{0} & = 0\\
\theta_1 & = \theta_0 + \left( 1 + \mu_0 \right) s_0 + \mu_0 \mu_{-1} m^{\prime}_{0} \\
\theta_1 & = \theta_0 + \left( 1 + \mu_0 \right) s_0 \\
\\
m^{\prime}_{1} & = \mu_{-1} m^{\prime}_0 + s_0 \\
& = s_0 \\
\theta_2 & = \theta_1 + \left( 1 + \mu_1 \right) s_1 + \mu_1 \mu_0 m^{\prime}_{1} \\
 & = \theta_1 + \left(1 + \mu_1 \right) s_1 + \mu_1 \mu_0 s_0 \\
\\
m^{\prime}_{2} & = \mu_{0} m^{\prime}_1 + s_1 \\
& = \mu_0 s_0 + s_1 \\
\theta_3 & = \theta_2 + \left( 1 + \mu_2 \right) s_2 + \mu_2 \mu_1 m^{\prime}_{2} \\
& = \theta_2 + \left(1 + \mu_2 \right) s_2 + \mu_2 \mu_1 s_1 + \mu_2 \mu_1 \mu_0 s_0 \\
\\
m^{\prime}_{3} & = \mu_{1} m^{\prime}_2 + s_2 \\
& = \mu_1 \mu_0 s_0 + \mu_1 s_1 + s_2 \\
\theta_4 & = \theta_3 + \left(1 + \mu_3 \right) s_3 + \mu_3 \mu_2 m^{\prime}_{3} \\
& = \theta_3 + \left(1 + \mu_3 \right) s_3 + \mu_3 \mu_2 s_2 + \mu_3 \mu_2 \mu_1 s_1 + \mu_3 \mu_2 \mu_1 \mu_0 s_0
\end{split}
\end{equation}
$$

Yep, that's the NAG update alright.

Note that for this implementation, you *don't* store the previous update, $v_t$,
but the momentum update $m^{\prime}_t$. Also, you need to store the previous
iterate's momentum coefficient $\mu_{t-1}$.

### The QHM-Dozat Nesterov Momentum Connection

Quasi-hyperbolic momentum and the Dozat version of Nesterov momentum have a
very similar form: QHM can be thought of as a generalization of Dozat Nesterov
momentum, by replacing $\mu_{t-1}$ in the definition of $m^{\prime}_{t+1}$ 
with a new free parameter, $\nu_{t}$. To avoid any confusion let's call this
new update $q_{t+1}$:

$$
\begin{equation}
\begin{split}
q_{t+1} & = \nu_t q_t + s_t \\
v_{t+1} & = \mu_{t} q_{t+1} + s_t \\
& = \mu_{t} \nu_t q_t + \left(1 + \mu_t \right) s_t
\end{split}
\end{equation}
$$

Setting:

* $\nu_t = 0$ and you get steepest descent (nearly).
* $\nu_t = \mu_{t-1}$ and you get Nesterov momentum.
* $\nu_t = 1$ and you get classical momentum (nearly).

So $\nu_t$ gives you an extra parameter along with $\mu_t$ to tune how much
weight should go on previous gradients when constructing the update vector.

At this point I should note that this isn't *exactly* the QHM update, although
it is quite close. Also, you only get "nearly" steepest decent and CM from this
equation: the step size for steepest descent is the "long" version we have grown
all too accustomed to. The QHM paper doesn't have the same issue mainly because
it takes care to both separate the *direction* of the update vector versus its
magnitude. This takes us further away from how NAG and classical momentum are
often defined, however.

## NAG in practice

So that's how it's all supposed to work in principle. Here I'll demonstrate
the equivalence of these methods, by implementing them all in simple R code.

The biggest simplification I'll make is that I'll assume a constant learning
rate and a constant momentum coefficient.

### Classical Momentum

```{r Classical mometum}
#' Optimization by Classical Momentum
#'
#' @param par Starting point of vector of parameters to optimize.
#' @param fn Objective function to optimize. Takes vector with length of
#' \code{par} and returns a scalar.
#' @param gr Gradient of the objective function \code{fn}. Takes vector with
#' length of \code{par} and returns a vector of the same length.
#' @param lr Learning rate.
#' @param mu Momentum coefficient.
#' @param max_iter Maximum number of iterations to optimize for. First iteration
#' is always steepest descent.
#' @return list with components: \code{par} final set of parameters; \code{f}
#' value of \code{fn} evaluated at the returned set of parameters; \code{fs}
#' vector of function evaluated after each iteration.
cm <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- mu * v - lr * g
    par <- par + v
    
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

This is a reference implementation of classical momentum. Nearly all the
parameters and the return values are the same for the other functions (except
where noted), so they're documented here once.

Onto the various NAG implementations. They all have an extra bit of code in them
to ensure that on the first iteration only the gradient descent stage is carried
out. This is needed to keep the outputs consistent between the different
implementations. The `cm` routine doesn't require any extra checks for steepest
descent on the first iteration, because initializing the velocity vector to zero
results in steepest descent anyway even if the momentum coefficient is non-zero.
As we'll see below, some of the other implementations either don't explicitly
use a velocity vector or need the momentum coefficient to be set to zero on the
first iteration to get the same result.

### NAG

```{r NAG}
# Optimization by Nesterov Accelerated Gradient
#
# Return list also contains gd_fs: function evaluated after gradient descent
# stage of each iteration; all: function evaluated after gradient descent
# stage and momentum stage, in order.
nag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)
  gd_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  x_old <- rep(0, length(par))
  for (i in 1:max_iter) {
    # gradient descent stage
    g <- gr(par)
    x <- par - (lr * g)

    # store gradient descent values
    f <- fn(x)
    gd_fs[i] <- f
    all[i * 2 - 1] <- f

    # momentum stage and update
    par <- x + ifelse(i == 1, 0, mu) * (x - x_old)
    x_old <- x

    # store momentum values
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, gd_fs = gd_fs, all = all)
}
```

In this routine, rather than store the velocity vector, we store the
previous gradient descent result, `x_old`. In order to ensure the
first iteration is gradient descent only, we also need to manually
set the momentum coefficient `mu` to zero on the first iteration,
which is what that `ifelse` expression does.

Also, there's some extra code to calculate and store the function values
after the gradient descent stage. These aren't needed for the optimization
to work, I just want to keep track of the values to demonstrate the 
different methods are in fact equivalent.

### Sutskever Formulation

```{r Sutskever}
# Optimization by Sutskever Nesterov Momentum
#
# Return list also contains mu_fs: function evaluated after momentum
# stage of each iteration; all: function evaluated after gradient descent
# stage and momentum stage, in order.
snag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  v <- rep(0, length(par))

  fs <- rep(0, max_iter)
  mu_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  for (i in 1:max_iter) {
    # momentum stage and update parameters
    mu_step <- mu * v
    par <- par + mu_step

    # store momentum results
    f <- fn(par)
    mu_fs[i] <- f
    all[i * 2 - 1] <- f

    # gradient descent stage
    g <- gr(par)
    gd_step <- -lr * g

    # update and store velocity for next step
    par <- par + gd_step
    v <- mu_step + gd_step

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, mu_fs = mu_fs, all = all)
}
```

This one is pretty straight-forward.

### Bengio Formulation

```{r Bengio}
# Optimization by Bengio Nesterov Momentum
bnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    par <- par + mu * mu * v - (1 + mu) * lr * g
    v <- mu * v - lr * g

    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

Because this implementation only uses a constant momentum coefficient, we
replace the $\mu_{t+1}\mu_{t}$ term with $\mu_{t}^2$. You would probably have to
using a pretty wacky non-constant momentum schedule where this approximation had
any major effect anyway... except when the momentum is changing from zero to
non-zero, that is, in which case that's the difference between some momentum and
no momentum. We shall return to this point later.

### Dozat Nesterov Momentum

```{r Dozat NAG}
# Optimization by Dozat's expression for Nesterov momentum
dnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)
  
  # mprime, the recursive variable
  mpt <- 0
  mumu <- mu * mu
  
  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    
    s <- lr * -g
    v <- (1 + mu) * s + mumu * mpt
    par <- par + v
    
    # update mprime for next step
    mpt <- mu * mpt + s
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

*18 December 2021*: Dozat's version of Nesterov momentum. We should expect this
to give the same result as NAG and Bengio's Nesterov momentum.

### Alternative momentum NAG Expression

```{r Alternative NAG}
# Optimization by Nesterov Accelerated Gradient, using a momentum-style
# update expression.
mnag <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- mu * (v - lr * g) - lr * g
    par <- par + v
    
    # setup v for the next iteration by removing the old gradient contribution
    v <- v + lr * g
    
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

Finally, here is your humble author's expression for NAG, written as a
momentum-style update. Effectively this is Zou and co-worker's unified momentum,
with the interpolation factor set to 1 to give NAG. To differentiate from the
Sutskever and Bengio versions of NAG, I'll refer to it as momentum-NAG or mNAG.

### Testing with Rosenbrock

Tradition dictates that I must demonstrate the use of these optimizers using the
2D Rosenbrock function, with a specific starting point:

```{r test function}
par <- c(-1.2, 1)

fn <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  100 * (x2 - x1 * x1) ^ 2 + (1 - x1) ^ 2
}
gr <- function(x) {
  x1 <- x[1]
  x2 <- x[2]
  c(
    -400 * x1 * (x2 - x1 * x1) - 2 * (1 - x1),
    200 *      (x2 - x1 * x1))
}
```

Let's run the optimizers for 100 iterations. The point here is not whether we
get really amazing optimization (we don't), but whether the outputs of the
Sutskever and Bengio algorithms are equivalent. It would be a bonus if my mNAG
result also worked. For comparison we'll throw in the classical momentum and as
a sanity check, the vanilla NAG routine.

```{r generating output}
lr <- 0.001
mu <- 0.95
max_iter <- 100
snag_opt <- snag(par, fn, gr, lr, mu, max_iter)
bnag_opt <- bnag(par, fn, gr, lr, mu, max_iter)
nag_opt <- nag(par, fn, gr, lr, mu, max_iter)
mnag_opt <- mnag(par, fn, gr, lr, mu, max_iter)
dnag_opt <- dnag(par, fn, gr, lr, mu, max_iter)

cm_opt <- cm(par, fn, gr, lr, mu, max_iter)

sbnag_df <- data.frame(Bengio = bnag_opt$fs,
                 Sutskever = snag_opt$fs, 
                 "Suts Mom" = snag_opt$mu_fs,
                 "NAG" = nag_opt$fs,
                 "mNAG" = mnag_opt$fs,
                 "Dozat" = dnag_opt$fs,
                 "CM" = cm_opt$fs)
```

Let's have a look at the first few iteration results:

```{r table head}
knitr::kable(head(sbnag_df), caption = paste(
  "First few evaluations of NAG implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

The first two columns pit the Sutskever vs Bengio formulations directly. And, as
we would expect, they're not the same: the Sutskever iteration result is from
the gradient descent stage, and the Bengio result comes from the momentum stage.
But if we put the momentum results from the Sutskever formulation up, we can see
that they are the same as the Bengio result, but behind by one iteration, i.e.
the Bengio result at iteration $t$ matches the Sutskever momentum stage result
at $t+1$. The final column indicates the results using the alternative
derivation I provided. It matches the Bengio results. Hurrah.

*18 December 2021* The results for Dozat's version of Nesterov momentum have
also been added. As expected, it matches the Bengio and mNAG results. Not much
elso to say about it so it won't appear in any further discussion.

However, none of the results match the vanilla NAG implementation. A clue to
what's going on is from looking at the classical momentum result in the "CM"
column. The first iteration of classical momentum is always plain gradient
descent, and that's also how NAG works by construction. We can see that both NAG
and CM have the same result on the first iteration, which suggests they're
working correctly. The Sutskever version is also doing gradient descent on its
first step. But the other implementations clearly aren't doing gradient descent
on their first iteration, and on subsequent iterations even the Sutskever
version diverges from the NAG result.

We'll get back to this, but let's just make sure these observations holds up at 
the end of the table too.

```{r table tail}
knitr::kable(tail(sbnag_df), caption = paste(
  "Last few evaluations of NAG implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

The Sutskever, Bengio and momentum NAG implementations all still match up in the
same way they did. CM and NAG are off doing their own thing. I suppose it's at
least reassuring that for this arbitrarily chosen set of learning rate and
momentum coefficient values, the NAG results all do better than classical
momentum. And in fact it seems like not doing plain gradient descent on the
first iteration gives better results, at least for this set of parameters.

### Behavior on first iteration

The reason for the differing behaviors of the NAG implementations is down to
two things:

* For some implementations (the Bengio and momentum NAG formulations), simply 
setting the initial velocity vector, $v_0$, to zero isn't enough to get gradient 
descent on the first iteration.
* The Sutskever version of NAG does a different pattern of gradient descent and
momentum stages in its iterations, compared to vanilla NAG.

For the first point, the implementation of Bengio Nesterov momentum and momentum
NAG are both giving the "long" version of the steepest descent on the first
iteration. In the discussion on unrolling the updates where we first came across
this point of difference with CM, I decided that the "long" version of the
first iteration was the right decision, but you can see that it causes some
extra issues in our comparisons.

What about the Sutskever formulation? Let's think about the chain of parameter 
updates that actually take place over the first few iterations:

For standard NAG, the chain is: gradient descent stage, momentum stage, gradient
descent stage, momentum stage. Except the momentum stage results in zero change
on iteration one because of the zero velocity vector, so what actually happens
is: gradient descent stage, gradient descent stage, momentum stage. Let's call
that `g|gm` for short with the bar indicating where the end of an iteration
occurs. To make the pattern more obvious, the first three iterations look like
`g|gm|gm`. This is the same pattern as classical momentum, although the `m`
stages are obviously different in content.

What about the Sutskever formulation? Its pattern for the first two iterations
is: momentum stage, gradient descent stage, momentum stage, gradient descent
stage. Except, once again, the momentum stage doesn't happen on the first
iteration, so you actually get: gradient descent stage, momentum stage, gradient
descent stage, and so on. So for Sutskever momentum, the first three iterations
look like `g|mg|mg`.

Therefore, as you can see, the initial set of gradient and momentum stages is
different for the Sutskever formulation, which is able to strictly interleave
gradient descent and momentum stages, whereas NAG begins, like classical
momentum, with two gradient descent stages. This difference shouldn't be too
surprising, because as we saw when deriving the Sutskever formulation, we
conveniently "forget" about an initial gradient descent stage compared to NAG.
Anyway, all this means that the parameters end up in different locations after
the first iteration.

## Consistent NAG momentum algorithms

It's not clear to me that there is any particular theoretical or practical
benefit to trying to bend these different formulations to give either the same
gradient descent on the first iteration as CM, or to having them give the same
pattern as vanilla NAG. But given the effort that's been made to show that NAG
is like classical momentum in some respects, we may as well see what it would
take.

### Sutskever Formulation consistent with NAG

```{r Sutskever consistent with NAG}
# Sutskever Nesterov Momentum consistent with NAG
#
# Extra parameter wait: wait this number of extra iterations before applying
# momentum. Needed only to sync up with other implementations of nesterov
# momentum: set wait to 1 to make mu_f at iter i match the output of the
# other implementations at iter i-1.
snagc <- function(par, fn, gr, lr, mu, max_iter = 10, wait = 0) {
  v <- rep(0, length(par))

  fs <- rep(0, max_iter)
  mu_fs <- rep(0, max_iter)
  all <- rep(0, max_iter * 2)

  for (i in 1:max_iter) {
    # momentum stage and update parameters
    mu_step <- ifelse(i > wait + 1, mu, 0) * v
    par <- par + mu_step

    # store momentum results
    f <- fn(par)
    mu_fs[i] <- f
    all[i * 2 - 1] <- f

    # gradient descent stage
    g <- gr(par)
    gd_step <- -lr * g

    # update and store velocity for next step
    par <- par + gd_step
    v <- mu_step + gd_step

    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
    all[i * 2] <- f
  }

  list(par = par, f = f, fs = fs, mu_fs = mu_fs, all = all)
}
```

The Sutskever formulation already does gradient descent on its first iteration,
so there's not a huge change required for the algorithm. Instead, I've introduced
a new parameter `wait`, that indicates the number of iterations to wait before
applying momentum. If this is set to `0`, then you get the current behavior.
Set it to `1` and the momentum step will be zero on the second iteration, too.
That gives us the two gradient descent stages in a row that should sync us up
with NAG.

### Bengio Formulation consistent with NAG

```{r Bengio consistent with NAG}
# Bengio Nesterov Momentum consistent with NAG
bnagc <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    if (i == 1) {
      par <- par - lr * g
    }
    else {
      par <- par + (mu * mu * v) - ((1 + mu) * lr * g)
      v <- (mu * v) - (lr * g)
    }
    
    # store gradient descent results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

"Fixing" the Bengio formulation turns out to be a bit more work than you might
think. As we noticed earlier, to get gradient decent on the first iteration,
$\mu_1$, needs to be set to zero. However, as a result this means the assumption
we made that $\mu_{t}\mu_{t-1} = \mu_{t-1}^2$ isn't correct for iteration 2:
$\mu_2\mu_1 = 0 \neq \mu_2^2$. However, we can't just set `mu` to 0 for the
second iteration, because the non-zero $\mu_2$ is used on its own as part of the
gradient descent expression.

There are two ways to deal with this: start storing the value of `mu` from the
previous iteration, even though I purposely chose a constant value of `mu` to
keep things simple, or just force $v_1$ to be zero by not updating it on the
first iteration. I chose the latter.

### Momentum NAG consistent with NAG

```{r Momentum NAG consistent with NAG}
# momentum NAG consistent with NAG
mnagc <- function(par, fn, gr, lr, mu, max_iter = 10) {
  fs <- rep(0, max_iter)

  v <- rep(0, length(par))
  for (i in 1:max_iter) {
    g <- gr(par)
    v <- (ifelse(i == 1, 0, mu) * (v - lr * g)) - lr * g
    par <- par + v
    
    # setup v for the next iteration by removing the old gradient contribution
    v <- v + lr * g
    
    # store results
    f <- fn(par)
    fs[i] <- f
  }

  list(par = par, f = f, fs = fs)
}
```

And last, the rewritten momentum NAG update. This required the least
modification from the original routine: simply set the momentum coefficient to
zero on the first iteration.

### Results for the "NAG-consistent" routines

Time to look at some numbers again. We'll compare the new versions of the
Sutskever, Bengio and Momentum formulations of NAG with vanilla NAG. We'll also
once again pull out the momentum stage results for Sutskever so we can compare
directly to the Bengio result.

```{r optimization forcing gradient descent initially}
mnagc_opt <- mnagc(par, fn, gr, lr, mu, max_iter)
snagc_opt <- snagc(par, fn, gr, lr, mu, max_iter, wait = 1)
bnagc_opt <- bnagc(par, fn, gr, lr, mu, max_iter)

ncdf <- data.frame(cBengio = bnagc_opt$fs,
                 cSutskever = snagc_opt$fs,
                 "Suts Mom" = snagc_opt$mu_fs,
                 NAG = nag_opt$fs, 
                 cmNAG = mnagc_opt$fs,
                 "NAG gd" = nag_opt$gd_fs)
```

Let's have a look at the first few iteration results:

```{r table of initial results}
knitr::kable(head(ncdf), caption = paste(
  "First few evaluations of NAG consistent implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

These are in the same order as the previous table. The Bengio and Sutskever
formulations don't match, as expected. But the Sutskever momentum result in the
third column *does* match the Bengio results from the previous iteration. So
relative performance between Sutskever and Bengio has been maintained. But in
addition, the NAG result in the fourth column matches the Bengio and Sutskever
momentum results. The momentum NAG result has also been correctly updated.
Finally, as an extra check, the NAG gradient desent result is shown in the final
column, which also matches up with the Sutskever results.

And let's take a look at the final few iterations, just to make sure everything
still holds up:

```{r table of final results}
knitr::kable(tail(ncdf), caption = paste(
  "Last few evaluations of NAG consistent implementations, with lr = ", 
  formatC(lr), "mu = ", formatC(mu),
  collapse = " "))
```

That's a relief.

## Conclusions

Yes, NAG can be thought of as being like classical momentum but where you do the
momentum step first and *then* the gradient step. But there's more to it than
that.

Both classical momentum and NAG use a weighted combination of the current and
past gradients to determine the next direction of descent. The difference is in
the distribution of weights: NAG places more emphasis on recent gradients. So it
might succeed better than classical momentum under conditions where the old
gradients quickly become uniformative (regions of high curvature?). Personally,
I find this view of NAG through the lens of a weighted history of gradients more
appealing than the Sutskever formulation because it leads to interesting ideas
like Quasi-Hyperbolic Momentum.

In terms of implementation, I pity anyone tasked with implementing Nesterov
momentum and demonstrating that they actually got it right. Imagine naively
implementing vanilla NAG, the Sutskever formulation and the Bengio formulation.
Using identical test functions, starting points and parameters, their output are
all different from each other *but they're all correct*!
